\documentclass{ximera}

\input{../preamble.tex}

\title{MATLAB}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{S:1.2}

We shall use \Matlab to compute addition and scalar
multiplication of vectors in two and three dimensions.  This
will serve the purpose of introducing some basic \Matlab
commands.

\subsection*{Entering Vectors and Vector Operations}

Begin a \Matlab session.  We now discuss how to enter a vector
into \Matlabp.  The syntax is straightforward; to enter the row
vector $x=(1,2,1)$ type\footnote{\Matlab has several useful line
editing features.  We point out two here:
\begin{enumerate}
\item[(a)]	Horizontal arrow keys ($\rightarrow,\leftarrow$)
	move the cursor one space without deleting a character.
\item[(b)]	Vertical arrow keys ($\uparrow,\downarrow$)
	recall previous and next command lines.
\end{enumerate} }

\begin{verbatim}
x = [1 2 1]
\end{verbatim}
\index{\computer![1 2 1]} and \Matlab responds with
\begin{verbatim}
x =
     1     2     1
\end{verbatim}

Next we show how easy it is to perform addition and scalar
multiplication\index{scalar multiplication!in \protect\Matlab}
in \Matlabp.  Enter the row vector $y=(2,-1,1)$ by typing
\begin{verbatim}
y = [2 -1 1]
\end{verbatim}
and \Matlab responds with
\begin{verbatim}
y =
     2    -1     1
\end{verbatim}
To add the vectors $x$ and $y$, type
\begin{verbatim}
x + y
\end{verbatim}
and \Matlab responds with
\begin{verbatim}
ans =
     3     1     2
\end{verbatim}
This vector is easily checked to be the sum of the vectors $x$
and $y$.  Similarly, to perform a scalar multiplication, type
\begin{verbatim}
2*x
\end{verbatim}
which yields
\begin{verbatim}
ans =
     2     4     2
\end{verbatim}
\Matlab subtracts the vector $y$ from the vector $x$ in the natural
way.  Type
\begin{verbatim}
x - y
\end{verbatim}
to obtain
\begin{verbatim}
ans =
    -1     3     0
\end{verbatim}

We mention two points concerning the operations that
we have just performed in \Matlabp.
\begin{enumerate}
\item[(a)] When entering a vector or a number, \Matlab
automatically echoes what has been entered.  {\em This echoing
can be suppressed by appending a semicolon to the line.\/} For
example, type
\begin{verbatim}
z = [-1 2 3];
\end{verbatim}
\index{\computer!;} and \Matlab responds with a new line awaiting a new
command.  To see the contents of the vector $z$ just type {\tt
z} and \Matlab responds with
\begin{verbatim}
z =
    -1     2     3
\end{verbatim}

\item[(b)] \Matlab stores in a new vector the information obtained by
algebraic manipulation.  Type
\begin{verbatim}
a = 2*x - 3*y + 4*z;
\end{verbatim}
Now type {\tt a} to find
\begin{verbatim}
a =
    -8    15    11
\end{verbatim}
We see that \Matlab has created a new row vector $a$ with the
correct number of entries.
\end{enumerate}
\noindent Note:  In order to use the result of a calculation later in a 
\Matlab session, we need to name the result of that calculation.  To recall 
the calculation {\tt 2*x - 3*y + 4*z}, we needed to name that calculation, 
which we did by typing {\tt a = 2*x - 3*y + 4*z}.  Then we were able to 
recall the result just by typing {\tt a}.


We have seen that we enter a row $n$ vector into \Matlab by
surrounding a list of $n$ numbers separated by spaces with
square brackets.  For example, to enter the $5$-vector
$w=(1,3,5,7,9)$ just type
\begin{verbatim}
w = [1 3 5 7 9]
\end{verbatim}
Note that the addition of two vectors is only defined when the
vectors have the same number of entries.  Trying to add the
$3$-vector $x$ with the $5$-vector $w$ by typing {\tt x + w} in \Matlab
yields the warning:
{\small
\begin{verbatim}
Arrays have incompatible sizes for this operation
\end{verbatim}
}

In \Matlab new rows are indicated by typing {\tt ;}. For
example, to enter the column vector
\[
z=\left(\begin{array}{r} -1 \\ 2\\ 3 \end{array}\right),
\]
just type:
\begin{verbatim}
z = [-1; 2; 3]
\end{verbatim}
\index{\computer![1; 2; 3]} and \Matlab responds with the column vector 
\begin{verbatim}
z =
    -1
     2
     3
\end{verbatim}
Individual entries of a vector can also be addressed.  For instance,
to display the first component of {\tt z} type {\tt z(1)}.

Note that \Matlab will add two row vectors only if they are the same size.  
Similarly for column vectors. Although \Matlab will add a row vector and 
a column vector to get a rectangular matrix (try typing {\tt x + z}), this a 
feature that will not be used in this text.

\subsection*{Entering Matrices}

Matrices\index{matrix} are entered into \Matlab row by row with
rows separated either by semicolons or by line returns.  To enter
the $2\times 3$ matrix
\[
A=\left(\begin{array}{ccc} 2 & 3 & 1 \\ 1 & 4 & 7
\end{array}\right),
\]
just type
\begin{verbatim}
A = [2 3 1; 1 4 7]
\end{verbatim}

\Matlab has very sophisticated methods for addressing the entries
of a matrix.  You can directly address individual entries,
individual rows, and individual columns.  To display the entry in
the $1^{st}$ row, $3^{rd}$ column of $A$, type {\tt A(1,3)}.  To
display the $2^{nd}$ column of $A$, type {\tt A(:,2)}; and to
display the $1^{st}$ row of $A$, type {\tt A(1,:)}.  For example,
to add the two rows of $A$ and store them in the vector $x$,
just type
\begin{verbatim}
x = A(1,:) + A(2,:)
\end{verbatim}
\index{\computer!:}

\Matlab has many operations involving matrices --- these will
be introduced later, as needed.



\includeexercises




\end{document}
