\documentclass{ximera}

\input{../preamble.tex}

\title{Vectors and Matrices}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{S:1.1}

In their elementary form, matrices and vectors are just lists of 
real numbers in different formats.  An $n$-vector is a list of $n$ 
numbers $(x_1,x_2,\ldots,x_n)$.  We may write this vector as a 
{\em row\/} vector\index{vector} as we have just done --- or as a 
{\em column\/} vector
\[
\vect{x}{n}.
\]
The set of all (real-valued) $n$-vectors is denoted by $\R^n$\index{$\R^n$}; 
so points in $\R^n$ are called vectors.  The sets $\R^n$ when $n$ is small
are very familiar sets.  The set $\R^1=\R$ is the real number
line, and the set $\R^2$ is the Cartesian plane\index{Cartesian plane}.
The set $\R^3$ consists of points or vectors in three dimensional space.

An $m\times n$ {\em matrix\/}\index{matrix} is a rectangular array
of numbers with $m$ rows and $n$ columns.  A general $2\times 3$
matrix has the form
\[
A=\left(\begin{array}{ccc} a_{11} & a_{12} & a_{13} \\
     a_{21} & a_{22} & a_{23} \end{array}\right).
\]
We use the convention that matrix entries $a_{ij}$ are indexed
so that the first subscript $i$ refers to the {\em row\/}
\index{row} while the second subscript $j$ refers to the {\em
column\/}\index{column}.  So the entry $a_{21}$ refers to the
matrix entry in the $2^{nd}$ row, $1^{st}$ column.

An $n\times m$ matrix $A$ and an $n'\times m'$ matrix $B$ are equal
precisely when the sizes of the matrices are equal ($n=n'$ and $m=m'$)
and when each of the corresponding entries are equal ($a_{ij}=b_{ij}$).

There is some redundancy in the use of the terms ``vector'' and
``matrix''.  For example, a row $n$-vector may be thought of as a 
$1\times n$ matrix, and a column $n$-vector may be thought of as a 
$n\times 1$ matrix.  There are situations where matrix notation is 
preferable to vector notation and vice-versa.


\subsection*{Addition and Scalar Multiplication of Vectors}

There are two basic operations on vectors: addition and scalar
multiplication.  Let $x=(x_1,\ldots,x_n)$ and
$y=(y_1,\ldots,y_n)$ be $n$-vectors.  Then
\[
x+y=(x_1+y_1,\ldots,x_n+y_n);
\]
that is, {\em vector addition\/}\index{vector!addition} is
defined as componentwise addition.

Similarly, {\em scalar multiplication\/}\index{scalar multiplication}
is defined as
componentwise multiplication.  A {\em scalar\/} is just a
number. Initially, we use the term scalar\index{scalar} to refer to a real
number --- but later on we sometimes use the term scalar
to refer to a {\em complex\/} number.  Suppose $r$ is a real
number; then the multiplication of a vector by the scalar $r$
is defined as
\[
rx = (rx_1,\ldots,rx_n).
\]

Subtraction of vectors\index{vector!subtraction} is defined simply as
\[
x-y = (x_1-y_1,\ldots,x_n-y_n).
\]
Formally, subtraction of vectors may also be defined as
\[
x-y = x+(-1)y.
\]
Division of a vector $x$ by a scalar $r$ is defined to be
\[
\frac{1}{r} x.
\]
The standard difficulties concerning division by zero still hold.

\subsection*{Addition and Scalar Multiplication of Matrices}

Similarly, we add two $m\times n$ matrices\index{matrix!addition}
by adding corresponding
entries, and we multiply a scalar times a matrix by multiplying each
entry of the matrix by that scalar\index{matrix!scalar multiplication}.
For example,
\[
\mattwo{0}{2}{4}{6} + \mattwo{1}{-3}{1}{4} = \mattwo{1}{-1}{5}{10}
\]
and
\[
4\mattwo{2}{-4}{3}{1} = \mattwo{8}{-16}{12}{4}.
\]
The main restriction on adding two matrices is that the matrices must
be of the same size.  So you cannot add a $4\times 3$ matrix to $6\times 2$
matrix --- even though they both have twelve entries.



\EXER

\TEXER

\noindent In Exercises~\ref{c1.1.1A} -- \ref{c1.1.1C}, let $x=(2,1,3)$ and 
$y=(1,1,-1)$ and compute the given expression.
\begin{exercise}  \label{c1.1.1A}
  $x+y\begin{prompt}
    = \left(\answer{3},\answer{2},\answer{2}\right).
  \end{prompt}$.
\end{exercise}
\begin{exercise}  \label{c1.1.1B}
  $2x-3y\begin{prompt}
    = \left(\answer{1},\answer{-1},\answer{9}\right)
  \end{prompt}$.
  \begin{hint}
    $2x - 3y = (4,2,6) - (3,3,-3)$.
  \end{hint}
  \begin{hint}
    $(4,2,6) - (3,3,-3) = (1,-1,9)$.
  \end{hint}  
\end{exercise}
\begin{exercise}  \label{c1.1.1C}
  $4x\begin{prompt}
    = \left(\answer{8},\answer{4},\answer{12}\right)
    \end{prompt}$.
\end{exercise}

\begin{exercise} \label{c1.1.2}
Let $A$ be the $3\times 4$ matrix
\[
A=\left(\begin{array}{rrrr} 2 & -1 & 0 & 1 \\ 3 & 4 & -7 & 10\\
        6 & -3 & 4 & 2 \end{array}\right).
\]
\begin{enumerate}
\item[(a)]  For which $n$ is a row of $A$ a vector in $\R^n$? \begin{prompt}$n = \answer{4}$\end{prompt}.
\item[(b)]  What is the $2^{nd}$ column of $A$?
  \begin{prompt}
    \[
      \left(\begin{array}{r} \answer{-1} \\ \answer{4} \\ \answer{-3} \end{array} \right)
    \]
  \end{prompt}
\item[(c)] Let $a_{ij}$ be the entry of $A$ in the $i^{th}$ row
  and the $j^{th}$ column.  What is $a_{23}-a_{31}$?
  \begin{prompt}
    \[
      a_{23}-a_{31} = \answer{-13}.
    \]
  \end{prompt}
\end{enumerate}
\end{exercise}

\noindent For each of the pairs of vectors or matrices in
Exercises~\ref{c1.1.3a} -- \ref{c1.1.3e}, decide whether addition
of the members of the pair is possible; and, if addition is possible,
perform the addition.
\begin{exercise}\label{c1.1.3a}
  $x=(2,1)$ and $y=(3,-1)$.
  
  \begin{multipleChoice}
    \choice[correct]{Addition is possible.}
    \choice{Addition is not possible.}
  \end{multipleChoice}
  \begin{exercise}
    $x + y = \left(\answer{5},\answer{0}\right)$.
  \end{exercise}
\end{exercise}

\begin{exercise}\label{c1.1.3b}
  $x=(1,2,2)$ and $y=(-2,1,4)$.
  
  \begin{multipleChoice}
    \choice[correct]{Addition is possible.}
    \choice{Addition is not possible.}
  \end{multipleChoice}
  \begin{exercise}
    $x + y = \left(\answer{-1},\answer{3},\answer{6}\right)$.
  \end{exercise}
\end{exercise}

\begin{exercise}\label{c1.1.3c}
  $x=(1,2,3)$ and $y=(-2,1)$.
  
  \begin{multipleChoice}
    \choice{Addition is possible.}
    \choice[correct]{Addition is not possible.}
  \end{multipleChoice}  
\end{exercise}

\begin{exercise}\label{c1.1.3d}
  $A=\mattwo{1}{3}{0}{4}$ and $B=\mattwo{2}{1}{1}{-2}$.
  
  \begin{multipleChoice}
    \choice[correct]{Addition is possible.}
    \choice{Addition is not possible.}
  \end{multipleChoice}
  \begin{exercise}
    $A + B = \mattwo{\answer{3}}{\answer{4}}{\answer{1}}{\answer{2}}$.
  \end{exercise}
\end{exercise}

\begin{exercise}\label{c1.1.3e}
  $A=\left(\begin{array}{rrr} 2 & 1 & 0\\ 4 & 1 & 0\\
             0 & 0 & 0\end{array}\right)$ and $B=\mattwo{2}{1}{1}{-2}$.
         
         \begin{multipleChoice}
           \choice{Addition is possible.}
           \choice[correct]{Addition is not possible.}
         \end{multipleChoice}         
\end{exercise}

\noindent In Exercises~\ref{c1.1.4A} -- \ref{c1.1.4B}, let
$A=\mattwo{2}{1}{-1}{4}$ and $B=\mattwo{0}{2}{3}{-1}$ and compute the given 
expression.
\begin{exercise}\label{c1.1.4A}
  $4A+B\begin{prompt}= \mattwo{\answer{8}}{\answer{6}}{\answer{-1}}{\answer{15}}\end{prompt}$.
\end{exercise}
\begin{exercise}\label{c1.1.4B}
  $2A-3B\begin{prompt}=\mattwo{\answer{4}}{\answer{-4}}{\answer{-11}}{\answer{11}}\end{prompt}$.
\end{exercise}

\end{document}
