\documentclass{article}
\usepackage{hyperref}

\title{Instructor Guide}
\author{Jim Fowler \and Marty Golubitsky}
\textwidth=6.0in
\evensidemargin=-1.0in
\oddsidemargin= 0.0in

\begin{document}

\maketitle


\section{Set up book files to create worksheets}\label{instructor-guide}

The following instructions assume you are using a Mac.

\subsection{Get the .cls files.}\label{get-the-.cls-files.}

In the Terminal, run

\begin{verbatim}
mkdir -p ~/Library/texmf/tex/latex
git clone https://github.com/XimeraProject/ximeraLatex ~/Library/texmf/tex/latex/ximeraLatex
\end{verbatim}

\subsection{Download resources}\label{download-resources}

First, download the book.

\begin{verbatim}
mkdir ~/linear-algebra
git clone https://github.com/mooculus/laode ~/linear-algebra/laode
\end{verbatim}

Then, download the worksheet builder.

\begin{verbatim}
git clone https://github.com/kisonecat/worksheet-builder ~/linear-algebra/worksheet-builder
\end{verbatim}

And install the other software needed to run the worksheet builder.

\begin{verbatim}
gem install optparse-pathname
\end{verbatim}

If you lack permissions to run \texttt{gem}, try

\begin{verbatim}
gem install --user-install optparse-pathname
\end{verbatim}

to install it locally.

\subsection{Build the book}\label{build-the-book}

Compile the book. This step is needed because various \texttt{aux} files
are created which will be read by the worksheet builder.

\begin{verbatim}
cd ~/linear-algebra/laode
lualatex --shell-escape linearAlgebra
lualatex --shell-escape linearAlgebra
\end{verbatim}



\section{Build a worksheet}

You can \href{https://osu.box.com/v/laode-labels}{view a PDF of the
textbook} which includes the labels that you can use in the
\texttt{\textbackslash{}exercise} command below.

\subsection{Try it}\label{try-it}

Now you are ready to try building a worksheet.

Go to the worksheet builder directory.

\begin{verbatim}
cd ~/linear-algebra/worksheet-builder
\end{verbatim}

Then create a file called \texttt{homework0.tex} with the following
content:

\begin{verbatim}
\documentclass{article}

\input{preamble}

\title{Homework 0}\author{Author Name}\date{Jan 1, 2019}

\begin{document}
\maketitle

\exercise{c5.1.4g}

\exercise{c5.1.7b}

\end{document}
\end{verbatim}

Now if you run

\begin{verbatim}
./build-both.sh --root=$HOME/linear-algebra/laode sample.tex
\end{verbatim}

two PDF files will be created: * one with solutions called
\texttt{homework0s.pdf} and * one without solutions called
\texttt{homework0e.pdf} and two TeX files will be created * one with
solutions called \texttt{homework0s.tex} and * one without solutions
called \texttt{homework0e.tex} and you may edit the TeX files by hand if
you wish.

\section{How to add exercises}\label{how-to-add-exercises}

\subsection{If you are not github enabled}

In this repository, there is a
\href{https://github.com/mooculus/laode/blob/master/howToContribute/template.tex}{template.tex}
file which can be used as a starting point for creating your own
exercises.

Once you have edited the template, please send the template to
fowler@math.osu.edu so that it can be added to the textbook repository.

\subsection{If you are github enabled}

\subsubsection{Fork the repository}

Go to https://github.com/mooculus/laode and click `Fork`

You should have a copy of the repository at https://github.com/YOUR-GITHUB-USERNAME/laode where **YOUR-GITHUB-USERNAME** is your GitHub username.

\subsubsection{Go to your local copy of the master repo}

cd ~/linear-algebra/laode
\\
set up the remote to point to your personal copy of the repo by running
\\
git remote rm origin
git remote add origin https://github.com/YOUR-GITHUB-USERNAME/laode
git remote add upstream https://github.com/mooculus/laode 


You will find it convenient to have symlinks to the numbered sections.  To create such symblinks, run
\\
ruby make-links.rb
\\
which will populate a subdirectory `by-number` with links to the sections of the textbook.

\subsubsection{Locate the appropriate section of the textbook}

Suppose you would like to add an exercise to Section 2.2, {\em The
Geometry of Low-Dimensional Solutions}.
\\
You will find the text for that section in [../solvingLinearEquations/theGeometryOfLowDimensionalSolutions.tex](../solvingLinearEquations/theGeometryOfLowDimensionalSolutions.tex), and a subdirectory of the same name ([../solvingLinearEquations/theGeometryOfLowDimensionalSolutions](../solvingLinearEquations/theGeometryOfLowDimensionalSolutions)) houses the exercise files.
\\
To look at the existing exercises, run
\\
cd ../by-number/SECTION.NUMBER
\\
Here, `SECTION.NUMBER` is the desired textbook section.

## Copy the template to the appropriate location

Pick a new, unused number, and copy [template.tex](template.tex) to
the appropriate directory with the other exercises for that section.

To do this, copy the file to a new number with
```
cp ~/linear-algebra/laode/howToContribute/template.tex 9999.tex
```
Here, `9999` is your chosen unused number.

## Write and save your exercise

Open the template you have copied to the appropriate directory and add
your content.

## Commit your changes

Commit your new exercise with
```
git add 9999.tex
git commit -m 'Description of my new exercise'
git push master origin
```

## Submit a pull request

In order for your exercise to appear in the master copy of the
textbook, submit a pull request.  You can do this by going to
https://github.com/YOUR-GITHUB-USERNAME/laode and clicking on `New Pull
Request`.









\section{Installing MATLAB}\label{installing-matlab}

\subsection{For Ohio State faculty and
students}\label{for-ohio-state-faculty-and-students}

Computers not only enable the completion of complicated calculations,
but can also improve the conceptual understanding of mathematics. This
course relies on MATLAB, a popular software package for matrix
computations. MATLAB is available free-of-charge to Ohio State students.
To download and activate your free copy of MATLAB, please follow the
steps below.

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  Launch Ohio State's IT Self-Service. Links to an external site. tool
  and click Login at the upper right.
\item
  Click Order Services. It is marked with a shopping cart but MATLAB
  will not cost you anything.
\item
  Click the Software Request link located under the ``Software
  Services'' heading.
\item
  Complete the Software Request Form, choosing MATLAB. Agree to terms
  and conditions.
\item
  Click Check Out.
\item
  Click Submit Order at the lower right.
\item
  Your order confirmation page lists your selections and their costs, as
  well as provides a Request Number (i.e., REQ12345) for tracking
  purposes.
\end{enumerate}

The rest of the instructions for installing MATLAB will automatically be
emailed to your Ohio State email account. Instructions will vary as to
whether you are on Windows, Mac OS, or Linux.

\subsection{Getting the m\_files}\label{getting-the-m_files}

To produce m\_files with filenames coming from the numeric labels in the
textbook, run

\begin{verbatim}
cd ~/linear-algebra/laode/m_files
ruby numeric-labels.rb
\end{verbatim}

Then to see the m\_files, run

\begin{verbatim}
cd ~/linear-algebra/laode/m_files/linearAlgebra
ls
\end{verbatim}

\section{Set up canvas}\label{set-up-canvas}

To set up a course shell in Canvas (``Carmen'') populated with linear
algebra content, email fowler@math.osu.edu.

\section{Syllabus}\label{syllabus}

To receive a sample syllabus and calendar, email fowler@math.osu.edu.

\section{Old exams}\label{old-exams}

Contact Marty Golubitsky (golubitsky.4@osu.edu) for old Math 2568 exams.


\end{document}
