\documentclass{ximera}

\input{./preamble.tex}

\title{mo13.tex}

\begin{document}
\begin{abstract}
BADBAD
\end{abstract}
\maketitle

\chapter{Matrix Normal Forms}

\subsection*{Section~\protect{\ref{S:RDM}} Real Diagonalizable Matrices}
\rhead{S:RDM}{REAL DIAGONALIZABLE MATRICES}

\exer{c10.3.1}
(a) The eigenvalues of $A$ are $\lambda_1 = 3$ and $\lambda_2 = -3$,
with corresponding eigenvectors $v_1 = (1,1)^t$ and $v_2 = (1,-1)^t$,
respectively.

(b) Let
\[
S = (v_1|v_2) = \mattwo{1}{1}{1}{-1}.
\]
Then $D = S^{-1}AS = \mattwo{3}{0}{0}{-3}$ is a diagonal matrix.

\exer{c10.3.3}
The eigenvalues of $A$ are $\lambda_1 = 1$ and $\lambda_2 = -1$.  The
eigenvector associated to $\lambda_1$ is $v_1 = (1,1,1)^t$.  There are
two eigenvectors associated to $\lambda_2$: $v_2 = (1,0,0)^t$ and
$v_3 = (0,1,2)^t$.
\[
S = (v_1|v_2|v_3) = \matthree{1}{1}{0}{1}{0}{1}{1}{0}{2}.
\]

\exer{c10.3.5}
Let $A = P^{-1}BP$ for some invertible matrix $P$, and let
$D = S^{-1}AS$, where $D$ is a diagonal matrix.  Then
\[
D = S^{-1}AS = S^{-1}(P^{-1}BP)S = (S^{-1}P^{-1})B(PS) = 
(PS)^{-1}B(PS).
\]
Therefore, $B$ is also similar to $D$, so $B$ is real diagonalizable.

\exer{c10.3.6A}  We assume that $Av=\lambda v$ and that $AB=BA$.  It follows that 
$ABv=BAv=\lambda Bv$.  Therefore $Bv$ is an eigenvector of $A$ with eigenvalue
$\lambda$.  Since $\lambda$ has only one independent eigenvector, it follows that 
$v$ is a multiple of $v$; that is, there is a scalar $\mu$ such that $Bv=\mu v$.

\newpage
\exer{c10.3.6C} Since $A$ is diagonalizable, there is an invertible matrix $S$ such 
that $S\inv AS$ is diagonal.  The diagonal entries of  $S\inv AS$ are the eigenvalues
of $A$; that is, the diagonal entries equal $\pm 1$.  Therefore, $(S\inv AS)^2=I_n$.
But $(S\inv AS)^2 = S\inv A^2 S$.   Therefore, $S\inv A^2 S=I_n$ which implies that
$A^2=I_n$.

\exer{c10.3.7}
Verify that the eigenvalues of $C$ are real and distinct using the
\Matlab command {\tt eig(C)}, which yields:
\begin{verbatim}
ans =
   -4.0000
  -12.0000
   -8.0000
  -16.0000
\end{verbatim}
In order to find the matrix $S$, either use the {\tt null} command to
find the eigenvectors of $C$ individually, or type
{\tt [S,D] = eig(C)} to obtain the matrix $S$ and the diagonal matrix
$D = S^{-1}CS$:
\begin{verbatim}
S =
    0.5314   -0.5547    0.0000    0.4082
   -0.4871    0.5547   -0.4082   -0.8165
    0.6199   -0.5547    0.8165    0.4082
   -0.3100    0.2774   -0.4082    0.0000

D =
   -4.0000         0         0         0
         0  -12.0000         0         0
         0         0   -8.0000         0
         0         0         0  -16.0000
\end{verbatim}

\exer{c10.3.8b}
\ans Matrix $B$ is not real diagonalizable.

\soln Compute the eigenvalues of $B$ using \Matlabp.  If a matrix is
real diagonalizable, it has real eigenvalues.  Matrix $B$ has complex
eigenvectors, and is therefore not real diagonalizable.



\subsection*{Section~\protect{\ref{S:CSE}} Complex Simple Eigenvalues}
\rhead{S:CSE}{COMPLEX SIMPLE EIGENVALUES}

\exer{c10.4.3} 
\ans
\[
T = \cmattwo{1 - i}{1 + i}{2i}{-2i}
\AND S = \mattwo{1}{-1}{0}{2}.
\]

\soln $T$ is the matrix whose columns are the complex eigenvectors
of $A$.  Multiplying $T^{-1}AT$ yields
\[
\cmattwo{2 + i}{0}{0}{2 - i},
\]
the matrix with the eigenvalues of $A$ along the main diagonal.  The
first column of the matrix $S$ contains the real part of the eigenvector
$(1 + i, -2i)^t$, and the second column contains the imaginary part of
this eigenvector. Multiplying $S^{-1}AS$ yields
\[
\cmattwo{2}{-1}{1}{2} = \mattwo{\sigma}{-\tau}{\tau}{\sigma}
\]
where $\sigma - i\tau$ is the eigenvalue $2 - i$ associated to
eigenvector $(1 + i, -2i)^t$.

\exer{c10.4.rotate}
Matrix $A$ rotates a vector by $45^{\circ}$ counterclockwise, then expands
the vector by a factor of $\sqrt{2}$.  Let $v_1 = Av_0$, $v_2 =
Av_1$, and so on.  Note that $v_8 = 16v_0$.  That is, after eight
iterations, the vector points in its original direction and has length
$16|v_0|$.

\para Let $w_1 = Bv_0$ and $w_2 = Bw_1$.  Then, $w_8 = 16v_0
= v_8$.  The rotation and dilatation caused by matrix $B$ coincides
with that of matrix $A$ every 8 iterations, or $360^\circ$.  However,
$B$ does not cause a single constant rotation or dilatation on each
iteration, as does $A$.

\exer{c10.4.7a}
\ans The matrices are:
\begin{verbatim}
T =
   0.9690             0.0197 + 0.3253i   0.0197 - 0.3253i
   0.1840             0.0506 - 0.5592i   0.0506 + 0.5592i
   0.1647            -0.4757 - 0.5935i  -0.4757 + 0.5935i
S =
   0.9690             0.0197        0.3253   
   0.1840             0.0506       -0.5592 
   0.1647            -0.4757       -0.5935 
\end{verbatim}

\soln The matrix $T$ is the matrix whose columns are eigenvectors of $A$. 
Find $T$ in \Matlab by typing {\tt [T,D] = eig(A)}.  The second and third columns
of $S$ correspond to the real and imaginary parts of the second eigenvector
of $A$.  Verify these solutions by noting that
{\tt inv(T)*A*T} yields a diagonal matrix with the eigenvalues of $A$
along the diagonal, and that {\tt inv(S)*A*S} yields a block matrix
with blocks consisting of the real and imaginary parts of each complex
conjugate pair of eigenvalues.

\exer{c10.4.7c}
\ans The matrices are:
\begin{verbatim}
T =
  Columns 1 through 4 
  -0.1933-0.2068i  -0.1933+0.2068i  -0.6791+0.5708i  -0.6791-0.5708i
  -0.0362+0.4192i  -0.0362-0.4192i   0.2735-0.3037i   0.2735+0.3037i
   0.4084+0.1620i   0.4084-0.1620i   0.0881+0.0243i   0.0881-0.0243i
  -0.0000-0.0000i  -0.0000+0.0000i  -0.0000+0.0000i  -0.0000-0.0000i
  -0.1933-0.2068i  -0.1933+0.2068i  -0.1321-0.0365i  -0.1321+0.0365i
   0.2657-0.6317i   0.2657+0.6317i   0.1321+0.0365i   0.1321-0.0365i
  Columns 5 through 6 
   0.4205-0.1238i   0.4205+0.1238i
   0.0855+0.2601i   0.0855-0.2601i
  -0.1639-0.1479i  -0.1639+0.1479i
  -0.5203+0.1710i  -0.5203-0.1710i
   0.4205-0.1238i   0.4205+0.1238i
  -0.4205+0.1238i  -0.4205-0.1238i
S =
   -0.1933   -0.2068   -0.6791    0.5708    0.4205   -0.1238
   -0.0362    0.4192    0.2735   -0.3037    0.0855    0.2601
    0.4084    0.1620    0.0881    0.0243   -0.1639   -0.1479
   -0.0000   -0.0000   -0.0000    0.0000   -0.5203    0.1710
   -0.1933   -0.2068   -0.1321   -0.0365    0.4205   -0.1238
    0.2657   -0.6317    0.1321    0.0365   -0.4205    0.1238
\end{verbatim}

\soln The matrix $T$ is the matrix whose columns are eigenvectors of $A$. 
Find $T$ in \Matlab by typing {\tt [T,D] = eig(A)}.  The first two columns
of $S$ correspond to the real and imaginary parts of the first eigenvector
of $A$, the third and fourth columns of $S$ contain the real and imaginary parts of 
the third eigenvector, and the last two columns contain the real and imaginary 
parts of the fifth eigenvector.  Verify these solutions by noting that
{\tt inv(T)*A*T} yields a diagonal matrix with the eigenvalues of $A$
along the diagonal, and that {\tt inv(S)*A*S} yields a block matrix
with blocks consisting of the real and imaginary parts of each complex
conjugate pair of eigenvalues.



\subsection*{Section~\protect{\ref{S:MGE}} Multiplicity and Generalized Eigenvectors}
\rhead{S:MGE}{MULTIPLICITY AND GENERALIZED EIGENVECTORS}

\exer{c10.5.1a} The eigenvalues of matrix $A$ are:
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
eigenvalue & algebraic multiplicity & geometric multiplicity \\
\hline
$2$ & $1$ & $1$ \\
$3$ & $2$ & $1$ \\
$4$ & $1$ & $1$ \\
\hline
\end{tabular}
\end{center}

\newpage
\exer{c10.5.1c} The eigenvalues of matrix $C$ are:
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
eigenvalue & algebraic multiplicity & geometric multiplicity \\
\hline
$-1$ & $3$ & $2$ \\
$1$ & $1$ & $1$ \\
\hline
\end{tabular}
\end{center}

\exer{c10.5.2a} \ans $v_1=(-1,1)$ and $v_2 = (0,1)$ is a basis.

\soln  The characteristic polynomial of $A$ is $\lambda^2-4\lambda+4=(\lambda-2)^2$.
Therefore, the eigenvalues of $A$ both equal $2$.  Find the eigenvectors of $A$ by 
solving $(A-2I_2)v=0$.  That is, solve
\[
\mattwo{-1}{-1}{1}{1}v = 0.
\]
$v_1=(-1,1)$ is the only independent solution.  Choose any vector for $v_2$ that is
independent of $v_1$.

\exer{c10.5.2c} \ans $v_1=(9,1,-1)$, $v_2=(-2,0,1)$, $v_3=(9,1,-2)$ is a basis.

\soln  Determine either by direct calculation of the characteristic polyomial of $C$ 
or by using \Matlab that the eiegenvalues of $C$ are $-1,1,1$.  Let $v_1$ be the
eigenvector with $-1$ eigenvalue; so $v_1=(9,1,-1)$.  There is only one independent 
eigenvector associated with the eigenvalue $1$ and that eigenvector is
$v_2=(-2,0,1)$.  Let $v_3$ be any generalized eigenvector associated with the
eigenvalue $1$; one choice is $v_3=(9,1,-2)$.  These eigenvectors can be found by
direct calculation or by using \Matlab.  When using \Matlab find $v_1$ by typing {\tt
null(C+eye(3))}, $v_2$ by typing {\tt null(D-eye(3))}, and $v_3$ by typing 
{\tt null((C-eye(3))\^{}2)}

\exer{c10.5.3A} \ans The eigenvalue $2$ has algebraic multiplicity $4$ and geometric
multiplicity $1$.

\soln Using \Matlab to find eigenvalues of high algebraic multiplicity is numerically
dangerous. Type {\tt eig(A)} and obtain
\begin{verbatim}
   2.0000 + 0.0006i
   2.0000 - 0.0006i
   1.9994          
   2.0006          
\end{verbatim}
Since the coefficients of $A$ are all integers, you might be suspicious of the answer
and guess that all of the eigenvalues of $A$ equal $2$.  Type {\tt null(A-2*eye(4))}
and obtain
\begin{verbatim}
ans =
   -0.4804
   -0.8006
   -0.1601
    0.3203
\end{verbatim}
dividing by {\tt ans(3)} yields the eigenvector $v_1=(3,5,1,-2)$.  To check whether
the eigenvalue $2$ has algebraic multiplicity greater than $1$, type 
{\tt null((A-eye(4))\^{}2)} and obtain
\begin{verbatim}
ans =
    0.5071         0
    0.8452         0
    0.1690         0
         0    1.0000
\end{verbatim}
Thus $v_2=(0,0,0,1)$ is a generalized eigenvector of $A$ with index $2$ and eigenvalue
$2$.  To find generalized eigenvectors of index $3$ type {\tt null((A-eye(4))\^{}3)} 
and obtain
\begin{verbatim}
ans =
   -0.9487         0         0
         0    1.0000         0
   -0.3162         0         0
         0         0    1.0000
\end{verbatim}
Thus, $v_3=(0,1,0,0)$ is a generalized eigenvector of index $3$.  Type 
{\tt null((A-eye(4))\^{}4)} and obtain
\begin{verbatim}
ans =
     1     0     0     0
     0     1     0     0
     0     0     1     0
     0     0     0     1
\end{verbatim}
to see that $2$ is an eigenvalue of $A$ of algebraic multiplicity $4$ and geometric
multiplicity $1$.  



\subsection*{Section~\protect{\ref{S:JNF}} The Jordan Normal Form Theorem}
\rhead{S:JNF}{THE JORDAN NORMAL FORM THEOREM}

\exer{c10.5.2}
\ans Two such matrices are:
\[
\left(\begin{array}{rrrr}
2 & 1 & 0 & 0 \\
0 & 2 & 0 & 0 \\
0 & 0 & 2 & 1 \\
0 & 0 & 0 & 2 \end{array}\right)
\AND
\left(\begin{array}{rrrr}
2 & 1 & 0 & 0 \\
0 & 2 & 1 & 0 \\
0 & 0 & 2 & 0 \\
0 & 0 & 0 & 2 \end{array}\right)
\]

\exer{c10.5.2B} \ans There are six different Jordan normal form matrices.

\soln  There are $3$ different Jordan matrices associated with the eigenvalue $4$ and
$2$ different Jordan matrices associated with the eigenvalue $-3$.  Altogether, there
are $3\cdot 2=6$ different Jordan matrices.

\exer{c10.5.3a}
\ans The Jordan normal form of matrix $A$ is
\[
\cmattwo{\frac{3 + \sqrt{17}}{2}}{0}{0}{\frac{3 - \sqrt{17}}{2}}.
\]

\soln Matrix $A$ has two distinct real eigenvalues at
$\lambda = \frac{3 + \sqrt{17}}{2}$ and 
$\lambda = \frac{3 - \sqrt{17}}{2}$.

\exer{c10.5.4}
\ans The Jordan normal form of matrix $C$ is
\[
\matthree{-1}{0}{0}{0}{2}{1}{0}{0}{2}.
\]

\soln Matrix $C$ has two real eigenvalues.  The eigenvalue at $-1$ has
algebraic multiplicity of $1$, and the eigenvalue at $2$ has algebraic
multiplicity of $2$, and only $1$ linearly independent eigenvector.

\exer{c10.5.4b} \ans The Jordan normal form of matrix $E$ is
\[
\matthree{1}{1}{0}{0}{1}{1}{0}{0}{1}.
\]

\soln Matrix $E$ has one eigenvalue at $1$ of algebraic multiplicity of $3$
and geometric multiplicity $1$.

\exer{c10.5.5A} \ans $e^{tJ} =
\cmatthree{e^t}{0}{0}{0}{e^{-t}}{te^{-t}}{0}{0}{e^{-t}}$.

\soln The matrix $J$ is in block diagonal form so we can compute the
matrix exponential by computing the matrix exponential of each block.  In
particular, let 
\[
M = \mattwo{-1}{1}{0}{-1}.
\]
Then 
\[
e^{tM} = e^{-t}(I_2+tM) = e^{-t}\mattwo{1}{t}{0}{1}.
\]
Therefore,
\[
e^{tJ} = \mattwo{e^t}{0}{0}{e^{tM}} = 
\cmatthree{e^t}{0}{0}{0}{e^{-t}}{te^{-t}}{0}{0}{e^{-t}}.
\]

\exer{c10.5.5}
(a) The linear map $N^2$ satisfies
\[
N^2e_1 = N^2e_2 = 0, \AND N^2e_j = e_{j - 2}
\mbox{ for } j = 3,\ldots,n.
\]
The linear map $N^3$ satisfies
\[
N^3e_1 = N^3e_2 = N^3e_3 = 0, \AND N^3e_j = e_{j - 3} \mbox{ for }
j = 4,\ldots,n.
\]
By induction, $N^ne_j = 0$ for $j = 1,\dots,n$.  So $N^n = 0$, and $N$
is nilpotent.

(b) Let $\lambda$ be an eigenvalue of matrix $N$.  Then, there
exists a nonzero vector $v$ such that $Nv = \lambda v$.  Therefore,
$N^nv = \lambda^nv$.  If $N$ is nilpotent, then $\lambda^nv = 0$.  Since
$v$ is nonzero, $\lambda^n = 0$, which implies $\lambda = 0$.

(c) Let $M = P^{-1}NP$ be a matrix similar to nilpotent $n \times n$
matrix $N$.  Then $M^n = (P^{-1}NP)^n = P^{-1}N^nP$.  Since $N$ is
nilpotent, $N^n = 0$.  So $M^n = 0$, and $M$ is also nilpotent.

(d) By the Jordan normal form theorem, $N$ is similar to some matrix $W$ which
is block diagonal and has the matrices $N_j$ along the diagonal, where each
$N_j$ is a Jordan block.  By (a) of this problem, $N_j^n = 0$ for each
$j$.  $W^n$ is block diagonal with blocks $N_j^n$.  Since $N_j^n = 0$ for all
$j$, $W^n = 0$.  Thus, $W$ is nilpotent, and, by (c) of this problem, $N$
is also nilpotent.

\exer{E:jnfma}
(a) \ans The Jordan normal form of $A$ is
\[
J = \left(\begin{array}{rrrr}
3 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 \\
0 & 0 & 0 & -1 \end{array}\right).
\]

\soln
To find the Jordan normal form of matrix $A$, type {\tt eig(A)} to find
the eigenvalues.  Matrix $A$ has four distinct eigenvalues, so the
Jordan normal form is the diagonal matrix with these eigenvalues along
its diagonal.  

(b)\ans   The diagonalizing matrix is
\begin{verbatim}
S =
   -0.1387   -0.1543   -0.0000   -0.5774
    0.1387   -0.3086   -0.4082    0.0000
    0.1387    0.9258    0.8165    0.5774
   -0.9707   -0.1543    0.4082   -0.5774
\end{verbatim}

\soln The columns of matrix $S$ consist of the eigenvectors of $A$.

\exer{E:jnfmc}
(a) \ans The Jordan normal form of $A$ is
\[
J = \left(\begin{array}{rrrr}
 0 & 2 &  0 &  0 \\
-2 & 0 &  0 &  0 \\
 0 & 0 & -2 &  1 \\
 0 & 0 & -1 & -2 \end{array}\right).
\]

\soln Type {\tt eig(A)} to find that matrix $A$ has distinct eigenvalues
at $\pm 2i$ and $2 \pm i$.

(b) \ans   The diagonalizing matrix is
\begin{verbatim}
T =
   -0.2118   -0.0456    0.2211    0.0060
   -0.8548   -0.2507    0.8762    0.1803
   -0.3555   -0.0988    0.3529    0.0669
   -0.1437   -0.0531    0.1440    0.0344
\end{verbatim}

\soln The columns of matrix $S$ consist of the eigenvectors of $A$.

\exer{E:jnfme}
(a) \ans The Jordan normal form of $A$ is
\[
J = \left(\begin{array}{rrrr}
-1 &  1 &  0 & 0 \\
0  & -1 &  1 & 0 \\
0  &  0 & -1 & 0 \\
0  &  0 &  0 & 1 \end{array}\right).
\]

(b) \ans  The diagonalizing matrix is
\begin{verbatim}
S =
    -1     1     0     0
    -1     1     0     1
    -1     0     1     1
    -1     0     0     1
\end{verbatim}

\soln To find $S$, use \Matlab to see that $A$ has two eigenvalues: $-1$ of
algebraic multiplicity three and $2$ of multiplicity one.  The geometric 
multiplicity of $-1$ is one.  Just type {\tt null(A+eye(4))} to see that 
$(1,1,1,1)$ is the only eigenvector corresponding to this eigenvalue.  Choose 
{\tt v3} in the null space of $(A+I_4)^3$ but not in the null space of 
$(A+I_4)^2$.  For example let {\tt v3 = [0 0 1 0]'}.  Then set 
{\tt v2 = (A+eye(4))*v3} and {\tt v1 = (A+eye(4))*v2}.  Finally, set 
{\tt v4 = null(A-2*eye(4))} and {\tt S = [v1 v2 v3 v4]}.



\subsection*{Section~\protect{\ref{S:TransitionTheory}} *Appendix:Markov Matrix
Theory}
\rhead{S:TransitionTheory}{*APPENDIX:MARKOV MATRIX THEORY}

\exer{c10.6.1}
Let $\lambda$ be an eigenvalue of $A$, and let $v \neq 0$ be an eigenvector
associated to $\lambda$.  Then,
\[
0 = \lim_{k \to \infty}A^kv = \lim_{k \to \infty}\lambda^kv
= \left(\lim_{k \to \infty}\lambda^k\right)v.
\]
Since $v$ is nonzero,
$\dps \lim_{k \to \infty}\lambda^k = 0$.  This is true only when
$|\lambda| < 1$.



 
\end{document}
