\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}


\noindent A general $2\times 2$ diagonal matrix has the form
$\mattwo{a}{0}{0}{b}$.  Thus two numbers $a$ and $b$ are
needed to specify each $2\times 2$ diagonal matrix.  

In Exercises~\ref{c1.3.1a} -- \ref{c1.3.3c}, how many different numbers
are needed to specify each of the given matrices:
\begin{exercise}  \label{c1.3.1a}
An upper triangular $2\times 2$ matrix?

\begin{solution}
\ans $3$.

\soln A $2\times 2$ upper triangular matrix $A$ has the form $A = \left( \begin{array}{cc}
            a_{11} & a_{12} \\
            0 & a_{22} \end{array} \right)$.  Thus the number of entries needed to specify $A$ is $3$.  

\end{solution}
\end{exercise}

\begin{exercise}  \label{c1.3.1b}
A symmetric $2\times 2$ matrix?

\begin{solution}
\ans $3$.

\soln  A $2\times 2$ symmetric matrix $A$ has the form $A = \left( \begin{array}{cc}
            a_{11} & a_{12} \\
            a_{12} & a_{22} \end{array} \right)$.  Thus the number of entries needed to specify $A$ is $3$.  

\end{solution}
\end{exercise}

\begin{exercise}  \label{c1.3.2}
An $m\times n$ matrix?

\begin{solution}
\ans $mn$

\soln Each row of the matrix has $n$ entries and there are $m$ rows.  Hence the number of entries needed to specify the matrix is $mn$.

\end{solution}
\end{exercise}

\begin{exercise}  \label{c1.3.3a}
A diagonal $n\times n$ matrix? 

\begin{solution}
\ans $n$.

\soln There are $n$ diagonal entries in an $n\times n$ diagonal matrix and all other entries are $0$.
\end{solution}
\end{exercise}

\begin{exercise}  \label{c1.3.3b}
An upper triangular $n\times n$ matrix? 

\begin{solution}

\ans  $\frac{n(n + 1)}{2}$

\soln  Recall the summation formula
    \[
      1 + 2 + \cdots + k = \frac{k(k+1)}{2}.
    \]
Observe that the first row of an $n\times n$ upper triangular matrix has $n$ different nonzero entries; the second row has $n-1$ new nonzero entries; \ldots; and the last row has $1$ (new) nonzero entry.  Hence the number of different nonzero entries in an $n\times n$ upper triangular matrix is
$1 + 2 + \cdots + (n-1) + n = \frac{n(n + 1)}{2}$.

\end{solution}
\end{exercise}

\begin{exercise}  \label{c1.3.3c}
A symmetric $n\times n$ matrix?   


\begin{solution}
\ans  $\frac{n(n + 1)}{2}$

\soln  Recall the summation formula
    \[
      1 + 2 + \cdots + k = \frac{k(k+1)}{2}.
    \]
Observe that the first row of a symmetric $n\times n$ matrix has $n$ different nonzero entries; the second row has $n-1$ new nonzero entries; \ldots; and the last row has $1$ new nonzero entry.  Hence the number of different nonzero entries in an $n\times n$ symmetric matrix is
$1 + 2 + \cdots + (n-1) + n = \frac{n(n + 1)}{2}$.


\end{solution}
\end{exercise}
\end{document}
