\documentclass{ximera}

\input{../preamble.tex}

\title{Bifurcation Theory}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{C:BT}

\normalsize

Many applications are modeled by autonomous systems of differential 
equations that contain parameters. As these parameters
change, the stylized phase portraits of the differential
equations may also change; parameter values where these
changes occur are called {\em bifurcation\/}\index{bifurcation} 
values.  In this chapter we discuss how bifurcations occur.  To frame  
the discussion we introduce two systems of differential equations --- the 
Volterra-Lotka equations modeling the population evolution of two 
species (Section~\ref{S:TSPM}) and the CSTR equations \index{CSTR}
modeling a single exothermic (or heat producing) chemical reaction (Section~\ref{S:CSTR}).  In these models, we use scaling to identify the 
essential parameters, and we illustrate changes that can take place in 
phase portraits as a parameter is varied.  The information concerning 
changes in phase portraits is summarized in {\em bifurcation diagrams}, 
which are introduced by simple examples in Section~\ref{S:bifurcation}. 
Bifurcation diagrams are used in Section~\ref{S:CSTR} to summarize the 
results of numerical explorations on the CSTR.

In Chapter~\ref{C:NPS} we showed that stylized phase portraits of planar 
Morse-Smale \index{Morse-Smale system} systems can often be drawn by a 
combination of analysis and computer.  In this chapter we observe that 
bifurcations occur at parameter values where phase portraits of systems of 
differential equations are not Morse-Smale.  There are three ways that an 
autonomous planar system of differential equations can fail to be 
Morse-Smale.
\begin{itemize}
\item  There is a nonhyperbolic equilibrium.\index{nonhyperbolic} 
\index{equilibrium!nonhyperbolic}
\item  There is a periodic solution that is not a limit cycle.
\item  There is a trajectory that connects a saddle to itself or
that connects two different saddles.  
\end{itemize}
Typically, in a system of differential equations that depends on a single
parameter $\rho$, there are isolated values in $\rho$ where the 
differential equation fails to be Morse-Smale and at these bifurcation
values the phase portraits do actually change.

Section~\ref{S:bifurcation} discusses the typical bifurcations associated 
with nonhyperbolic equilibria: saddle-node bifurcations (where two equilibria 
are created) and Hopf bifurcations (where limit cycles are created).  These 
bifurcations are called {\em local bifurcations\/} as the changes in the phase 
portraits occur in a small neighborhood of the nonhyperbolic equilibrium.  

We also describe typical ways in which the remaining two failures in 
Morse-Smale occur.  Typically, when periodic solutions are not limit cycles, 
two periodic solutions collide and disappear (Section~\ref{S:GlobalBif}).  
Other global bifurcations occur when there are saddle-saddle connections.  
Homoclinic bifurcations occur when there is a connection from a saddle point 
to itself --- typically such bifurcations occur when a limit cycle disappears
(Section~\ref{S:bifurcation}).   Heteroclinic bifurcations occur when a 
trajectory connects two different saddles (Section~\ref{S:GlobalBif}).

Additional details concerning saddle-node and Hopf bifurcations are given in 
Sections~\ref{S:SNB} and \ref{S:HopfBif}.




\end{document}
