\documentclass{ximera}

\input{./preamble.tex}

\title{Higher Dimensional Systems}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{C:HDS}

\normalsize

In Chapter~\ref{chap:SolveOdes} we saw that equilibria and their 
stabilities completely determined the phase line dynamics for single
autonomous first order differential equations.  The stability of the
equilibria determine the direction in which one equilibrium is connected 
to the next.

In Chapter~\ref{C:NPS} we discussed the extent to which equilibria and 
their stability determined the phase planes of systems of two autonomous 
systems of ordinary differential equations.  When an equilibrium is 
hyperbolic, nonlinear planar systems behave much like their linearizations 
--- at least on a small neighborhood of the equilibrium.  However, away 
from equilibria, planar systems can have dynamically more interesting 
states: limit cycles.  We have seen that qualitatively we can understand the
global dynamics of most planar systems (the Morse-Smale ones) if we can find 
their equilibria, their periodic
solutions, and their connecting orbits.  Analytically, this is an impossible
problem to solve in closed form --- but numerically this kind of calculation
is often tractable.

In this chapter we discuss briefly the dynamics of systems of three or
more autonomous first order differential equations.  The situation is
very complicated --- even on the qualitative level.  On the positive
side, nonlinear systems do behave like their linear counterparts on a
neighborhood of hyperbolic equilibria, and the dynamics of the linearized
systems can be understood as a consequence of the Jordan normal form theorem.
These issues are discussed in Sections~\ref{sec:LinHomSys} and \ref{S:QT}.  
On the negative side, in Sections~\ref{S:NLD} and \ref{S:chaos} we will see 
that the dynamics of nonlinear systems away from equilibria are just much 
more complicated than their planar counterparts.  In particular, we will see 
that quasiperiodic motion may be expected (first in linear nonhyperbolic four 
dimensional systems and then in nonlinear three dimensional systems).  
Finally, we show that even complicated `chaotic' motion may be expected in 
three dimensions (the Lorenz equations). 

The discussion in Sections~\ref{S:NLD} and \ref{S:chaos} is predicated on 
being able to solve numerically systems of differential equations with more 
than two equations.  To do this, we must use the \Matlab differential 
equations solver {\tt ode45} directly.  We introduce this solver in 
Section~\ref{S:ode45} by solving certain one-dimensional differential 
equations.  In this section we also discuss how to store functions in \Matlab 
m-files.  In Section~\ref{S:ode45HD} we use {\tt ode45} to solve several 
sample differential equations in three and four dimensions, and we display 
the results using \Matlab graphics. 

The discussion in this chapter continues an important theme: what 
information can we learn about the dynamics and solutions of nonlinear 
systems of ordinary differential equations from numerical simulation.  
Indeed, what can mathematics say that will help in interpreting numerically 
obtained solutions?  We will see that even on the qualitative level, the 
situation is very complicated --- complicated enough to guarantee that closed 
form solutions are, in general, not an option.  It should be noted, however,
that closed form solutions do exist for many particular types of equations 
(not least of which are the linear constant coefficient systems).  In later
chapters we will discuss some of the techniques of integration that allow us
to solve certain special kinds of differential equations in closed form.


\end{document}
