\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}
\begin{exercise} \label{c5.5.4}
Show that the set $V$ of all $2\times 2$ matrices is a vector space.
Show that the dimension of $V$ is four by finding a basis of $V$
with four elements.  Show that the space $M(m,n)$ of all $m\times n$
matrices is also a vector space.  What is $\dim M(m,n)$?

\begin{solution}

The set $V$ is a vector space because the operations of addition and
scalar multiplication satisfy the eight properties of
vector spaces described in Table~\ref{vectorspacelist}.  For $2 \times
2$ matrices, matrix addition is defined for two matrices such that:
\[
\mattwo{a_1}{b_1}{c_1}{d_1} + \mattwo{a_2}{b_2}{c_2}{d_2} =
\mattwo{a_1 + a_2}{b_1 + b_2}{c_1 + c_2}{d_1 + d_2}
\]
and scalar multiplication is defined for a matrix and a scalar such that:
\[
s\mattwo{a}{b}{c}{d} = \mattwo{sa}{sb}{sc}{sd}.
\]
So, using these definitions, addition is commutative and associative,
and the additive identity is the $2 \times 2$ matrix of zeroes.  
If $W$ is a matrix, then its additive inverse is $-W$.  
More specifically, if 
\[
W = \mattwo{w_{11}}{w_{12}}{w_{21}}{w_{22}} \mbox{then}
\quad -W = \mattwo{-w_{11}}{-w_{12}}{-w_{21}}{-w_{22}}.
\]
Scalar multiplication is associative.  There is a multiplicative
identity, $I_2$, and scalar multiplication is distributive both
for scalars and for matrices.  So $V$ is a vector space.  One basis
for $V$ is
\[
\mattwo{1}{0}{0}{0}, \mattwo{0}{1}{0}{0}, \mattwo{0}{0}{1}{0}
\AND \mattwo{0}{0}{0}{1}.
\]

\para The set of $m \times n$ matrices is also a vector space, since
it also satisfies the eight properties of vector spaces.  In this
case, the additive identity is the $m \times n$ zero matrix, and
the multiplicative identity is $I_n$.  The dimension of the set is
$mn$, since one basis consists of the $mn$ matrices with $a_{ij}
= 1$ and all other entries $0$, for $1 \leq i \leq m$ and $1 \leq j
\leq n$.

\end{solution}
\end{exercise}
\end{document}
