\documentclass{ximera}

\input{../preamble.tex}

\title{Autonomous Planar Nonlinear Systems}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{C:NPS}

\normalsize

In Chapter~\ref{chap:SolveOdes} we discussed how the phase line for a single
autonomous nonlinear differential equation is determined by the equilibria of 
the differential equation.  Once the equilibria and their stability properties
are known, we can find the long time or asymptotic behavior of every solution 
to a single equation, even though we do not necessarily know a closed form 
formula for that solution.  

In this chapter we discuss the corresponding and more complicated results for 
planar systems of nonlinear autonomous differential equations.  We discuss 
precisely what information about planar phase portraits is needed in order to 
be able to tell the (asymptotic) fate of all solutions.  The information that 
we need includes the equilibria and their type (as discussed in 
Chapter~\ref{Chap:PlanarQ}), periodic solutions and their stability type, and
connecting trajectories.  Once we have the needed information, we can 
determine the qualitative features of all solutions to a planar system --- 
even though we cannot write a closed form formula for these solutions.  

In Section~\ref{S:introAPNS} we experiment numerically with nonlinear planar 
systems, introducing the information that is needed to draw a qualitative 
phase plane portrait of a differential equation.  We see that we will need to 
know the equilibria and their type (saddle, sink, or source), time periodic 
solutions and their stability, and trajectories that connect these solutions 
(including stable and unstable orbits emanating from saddles).  

In Section~\ref{S:linearization} we look more closely at the role of
linearized systems of differential equations near a hyperbolic equilibrium. 
In particular, we introduce the Jacobian matrix and show numerically that 
solutions near an equilibrium behave like solutions to the linearized 
system.  Then we use the results of Chapter~\ref{Chap:PlanarQ} to understand
specific features of the local behavior of nonlinear systems near a saddle,
sink, or source in terms of the corresponding features of solutions to linear 
differential equations.

The analytic discussion of aspects of time periodic solutions is introduced 
in Section~\ref{S:periodic}.  In particular, we show how 
periodic solutions to certain systems of differential equations can be 
constructed using phase-amplitude equations in polar coordinates.

This information is then synthesized in Section~\ref{S:SPP} in terms of 
stylized phase portraits for Morse-Smale planar systems of autonomous 
differential equations.  Morse-Smale systems are differential equations with 
properties that allow us to find qualitative phase plane portraits.  
Moreover, in a sense that we will not try to make precise, most planar 
autonomous systems are Morse-Smale.  


\end{document}
