\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}

\noindent In Exercises~\ref{c14.5.6} -- \ref{c14.5.9} solve the given initial 
value problem by an appropriate solution technique.
\begin{exercise} \label{c14.5.6}
$\dps \frac{dx}{dt} = \sec\left(\frac{x}{t}\right)+\frac{x}{t}$ where
$x(1)=\pi$.

\begin{solution}
\ans $x(t) = t\sin\inv(\ln|t|)$.

\soln  This differential equation is an equation with homogeneous
coefficients where $F(v) = v + \sec v$.  Using the substitution $v=x/t$, we
are led to the differential equation $t\dot{v}=\sec v$, which can be solved
by separation of variables.  That is, solve
\[
\cos v \frac{dv}{dt} = \frac{1}{t}
\]
by integration, obtaining
\[
\sin v = \ln|t| + c.
\]
We can solve for $c$ using the initial condition $x(1)=\pi$, which implies
that $v(1)=x(1)/1=\pi$.  Therefore, $c=0$.  Hence
\[
v = \sin\inv(\ln|t|).
\]
Using the identity $v=x/t$ leads to the answer. 
 

\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.7}
$\dps \frac{dx}{dt} = x+x^2$ where $x(2)=1$.

\begin{solution}
\ans $x(t) = \dps\frac{t}{4-t}$.

\soln  Although this equation is a Bernoulli equation, it is most directly
solved by separation of variables.  That is, solve
\[
\frac{1}{x(x+1)}\frac{dx}{dt} = 1.
\]
Using partial fractions, we obtain
\[
\left(\frac{1}{x}-\frac{1}{x+1}\right)\frac{dx}{dt} = 1
\]
On integration, we find
\[
\ln\left|\frac{x}{x+1}\right| = \ln|t| + c.
\]
Therefore,
\[
\frac{x}{x+1} = ct,
\]
for some constant $c$.  Using the initial condition $x(2)=1$ leads to
$c=\frac{1}{4}$.  After cross-multiplying we are led to the answer.

\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.8}
$\dps \frac{dx}{dt} = -\frac{x}{t}-t^3 x^3$ where $x(1)=1$.

\begin{solution}
\ans $x(t) = \dps\frac{1}{t^2}$.

\soln This differential equation is a Bernoulli equation with
\[
r(t) = -\frac{1}{t},\quad s(t) = -t^3 \AND p=3.
\]
After transformation the corresponding linear initial value problem
is given by
\[
\frac{dv}{dt} = \frac{2}{t} v + 2t^3,\quad v(1)=1.
\]
We solve this equation by variation of parameters.  The solution
of the homogeneous equation $\dot v = \frac{2}{t} v$ is
$v_h(t)=t^2$, and we use Theorem~\ref{thm:varpar} to find the solution
\[
v(t) = c(t) v_h(t) = t^2 v_h(t) = t^4.
\]
Back substitution now leads to the solution
\[
x(t) = (v(t))^{-\frac{1}{2}} = \frac{1}{t^2}.
\]


\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.9}
$\dps \frac{dx}{dt} = \frac{x(x+t)}{t^2}$ where $x(1)=1$.

\begin{solution}
\ans $x(t) = \dps\frac{t}{1-\ln|t|}$.

\soln  This differential equation is an equation with homogeneous
coefficients where $F(v) = v + v^2$.  Using the substitution $v=x/t$, we
are led to the differential equation $t\dot{v}=v^2$, which can be solved
by separation of variables.  That is, solve
\[
 \frac{1}{v^2} \frac{dv}{dt} = \frac{1}{t}
\]
by integration, obtaining
\[
-\frac{1}{v} = \ln|t| + c.
\]
We can solve for $c$ using the initial condition $x(1)=1$, which implies
that $v(1)=x(1)/1=1$.  Therefore, $c=-1$.  Hence
\[
v = \frac{1}{1-\ln|t|}.
\]
Using the identity $v=x/t$ leads to the answer.   Note that this equation is
also a Bernoulli equation and could be solved using the techniques for
Bernoulli equations, as well.

\end{solution}
\end{exercise}
\end{document}
