\documentclass{ximera}

\input{../preamble.tex}

\title{Closed Form Solutions for Planar ODEs}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{Chap:Planar}

\normalsize

In this chapter we describe two methods to find closed form
solutions to planar constant coefficient systems of linear differential
equations.  In Section~\ref{S:6.1} we show how uniqueness to
initial value problems implies that the space of solutions to a constant
coefficient system of $n$ linear differential equations is $n$ dimensional.
Using this observation we present a direct method for solving planar linear
systems in Section~\ref{S:TDM}.  This method extends the discussion of
solutions to systems whose coefficient matrices have distinct real
eigenvalues given in Section~\ref{S:IVPR} to the cases of complex
eigenvalues and equal real eigenvalues.

The second method for finding solutions is to use changes of coordinates to 
make the coefficient matrix of the differential equation as simple as possible.  
This idea leads to the notion of {\em similarity\/} of matrices, which is discussed 
in Section~\ref{S:6.5}, and leads to the second method for solving planar linear 
systems. Similarity also leads to the Jordan Normal Form theorem for 
$2\times 2$ matrices.
Both the direct method and the method based on similarity require 
being able to compute the eigenvalues and eigenvectors of the coefficient matrix.

The important subject of qualitative features of phase portraits of linear systems are explored in
Section~\ref{S:6.7}.  Specifically we discuss {\em saddles\/}, {\em sinks\/}, 
{\em sources\/} and asymptotic stability.  This discussion also uses similarity 
and Jordan Normal Form.  We find that the qualitative theory is determined 
by the eigenvalues and eigenvectors of the coefficient matrix --- which is not 
surprising given that we can classify matrices up to similarity by just knowing 
their eigenvalues and eigenvectors.  

Chapter~\ref{Chap:Planar} ends with three optional sections.  An elegant third 
way to derive closed form solutions to $n$-dimensional linear ODE systems is
given in Section~\ref{S:Matrixexp} using matrix exponentials.  This method 
leads to a proof of uniqueness of solutions to initial value problems of linear systems.  
See Theorem~\ref{T:linODEsoln}.  A proof of the Cayley Hamilton Theorem 
for $2\times 2$ matrices is given in Section~\ref{S:6.6}.
In the last section, Section~\ref{S:SOE}, we obtain solutions to second order 
equations by reducting them to first order systems.


\end{document}
