\documentclass{ximera}

\input{./preamble.tex}

\title{Laplace Transforms}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{C:LT}

\normalsize

Laplace transforms are used to solve forced linear differential equations.  
In Section~\ref{sec:2norderinhom} we used the method of undetermined 
coefficients to solve forced equations when the forcing term $g(t)$ is of a 
special form, namely, when $g(t)$ is a linear combination of the functions 
$t^je^{\lambda t}$, $t^je^{\sigma t}\sin(\tau t)$, and 
$t^je^{\sigma t}\cos(\tau t)$.  Although Laplace transforms can be used to 
solve such systems as well, it is usually more efficient to use the method of 
undetermined coefficients when that method is applicable.  The strength of the 
method of Laplace transforms is that it can be used to solve forced linear 
differential equations when the forcing term is more general.  Specifically,
Laplace transforms can be used to solve forced equations when the forcing
term is either discontinuous (a step function) or an impulse function (a Dirac 
delta function).

The idea behind the {\em Laplace transform method} is that it is possible to 
transform any function so that there is a simple relationship between the 
transform of that function and the transform of its derivative.  It is this 
observation, coupled with linearity, that leads to a useful and elegant
method for solving linear, higher order, inhomogeneous differential equations.
In Section~\ref{S:13.1} we discuss this property and the Laplace
transform method for solving differential equations.  In Section~\ref{S:13.3}
we formally introduce the Laplace transform (as an improper integral), show
that this definition satisfies the basic properties introduced in
Section~\ref{S:13.1}, and compute the Laplace transform for a variety of
functions including step functions and Dirac delta functions.

When using the method of Laplace transforms to solve linear differential
equations, it is immediately clear that this method demands the computation of 
partial fraction expansions.  Some of the details of partial fraction
expansions are discussed in Section~\ref{S:PF}.  

In Section~\ref{S:13.4} we solve several linear differential equations with
discontinuous forcing using the methods developed in the first three sections. 
RLC circuits provide an excellent example of a physical problem modeled by
second order linear forced differential equations, and this model is discussed
in Section~\ref{S:SOFE}.  Discontinuous periodic forcing (AC current) also
occurs in these models and their analysis is discussed in that section.


\end{document}
