\documentclass{ximera}

\input{../preamble.tex}

\title{The Method of Laplace Transforms}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

 \label{S:13.1}

In this section we describe the basic properties of Laplace transforms and 
show how these properties lead to a method for solving forced equations.   
We also discuss the kind of information that we will need about Laplace 
transforms in order to solve a general second order forced equation. 

\subsection*{The Basic Properties}

Suppose that there is an operation or transform that transforms functions $x$ 
defined on a variable $t$ to functions ${\cal L}[x]$ defined on a variable 
$s$ in such a way that:
\begin{equation}  \label{Tprops}
\begin{array}{cl}
(a) & {\cal L}\; \mbox{is linear},\\
(b) & {\cal L}\; \mbox{is invertible},\; \mbox{and}\\
(c) & {\cal L}[\dot{x}](s) = s{\cal L}[x](s) - x(0).
\end{array}
\end{equation}\index{Laplace transform!general properties}
Note that \Ref{Tprops}(c) relates the transform of the derivative of a 
function to the transform of the function itself.  We call a transform that 
satisfies \Ref{Tprops} the {\em Laplace transform\/}, as there is only one 
transform that satisfies these three properties.

\subsubsection*{The Laplace Transform of $e^{at}$}

We can use properties \Ref{Tprops} to compute the Laplace transform of the
function $x(t) = e^{at}$.  To perform this calculation we need only recall 
that $e^{at}$ is the unique solution to the differential equation 
$\dot{x}=ax$ with initial value $x(0)=1$.  In essence, we can compute the 
Laplace transform of $e^{at}$ without actually having defined the Laplace 
transform.

It follows from the differential equation $\dot{x}=ax$ and \Ref{Tprops}(a) 
that
\[
{\cal L}[\dot{x}](s) = {\cal L}[ae^{at}](s) = a{\cal L}[e^{at}](s)
\]
and from \Ref{Tprops}(c) that
\[
{\cal L}[\dot{x}](s) = s{\cal L}[e^{at}](s)-1.
\]
Equating these two expressions for ${\cal L}[\dot{x}]$ leads to
\[
s{\cal L}[e^{at}](s)-1 = a{\cal L}[e^{at}](s),
\]
and hence to 
\begin{equation} \label{eq:Lexp}
{\cal L}[e^{at}](s) = \frac{1}{s-a}.
\end{equation}

\subsubsection*{Solving an Inhomogeneous Equation by Laplace Transforms}

Properties \Ref{Tprops} and formula \Ref{eq:Lexp} allow us to solve the 
initial value problem 
\begin{equation} \label{eq:expODE}
\begin{array}{crcl}
(a) & \dot{x} - 3x & = & e^{2t}\\
(b) & x(0) & = & 1.
\end{array}
\end{equation}

Before proceeding, note that \Ref{eq:expODE} can be solved directly using
the method of undetermined coefficients.  \index{undetermined coefficients}  
Indeed, undetermined coefficients show that $x_p(t) = 2e^{2t}$ is a 
particular solution to \Ref{eq:expODE}(a).  Since the general solution to the 
homogeneous equation is $\alpha e^{3t}$, the general solution to the 
inhomogeneous equation \Ref{eq:expODE}(a) is:
\[
x(t) = 2e^{2t} + \alpha e^{3t}.
\]
Setting $\alpha=-1$ solves the initial value problem \Ref{eq:expODE}(b).

We illustrate the method of Laplace transforms by solving \Ref{eq:expODE} 
using Laplace transforms.  Apply the transform ${\cal L}$ to both sides of 
\Ref{eq:expODE}, obtaining 
\[
{\cal L}[\dot{x}-3x](s) = {\cal L}[e^{2t}](s).
\]
From \Ref{eq:Lexp} with $a=2$, we see that
\[
{\cal L}[e^{2t}](s) = \frac{1}{s-2}
\]
Using \Ref{Tprops}(c,a), we see that  
\[
{\cal L}[\dot{x}-3x](s) = (s-3){\cal L}[x](s) - 1.
\]
Hence 
\[
(s-3){\cal L}[x](s) = 1 + \frac{1}{s-2}.
\]

Next, we solve for ${\cal L}[x](s)$, obtaining
\[
{\cal L}[x](s) = \frac{1}{s-3} + \frac{1}{(s-2)(s-3)}.
\]
Using partial fractions\index{partial fractions}, we find  
\[
{\cal L}[x](s)  = \frac{2}{s-3} - \frac{1}{s-2}.
\]
We describe the method of partial fractions in more detail in 
Section~\ref{S:PF}.  The particular uses of partial fractions in this section 
all follow from Exercise~\ref{Ex:pf}. 

Finally, from \Ref{eq:Lexp} we see that 
\[
{\cal L}[2e^{3t}-e^{2t}](s) = \frac{2}{s-3} - \frac{1}{s-2},
\]
and using \Ref{Tprops}(b) we conclude that
\[
x(t) = 2e^{3t}-e^{2t} 
\]
is the unique solution to the initial value problem \Ref{eq:expODE}. 

\subsection*{Three Steps in Solving Equations by Laplace Transforms} 

We can summarize the method for solving ordinary differential equations by
Laplace transforms in three steps.  In this summary it will be useful to 
have defined the inverse Laplace transform.
\begin{Def}  \label{D:invLaplace}
The {\em inverse Laplace transform\/} of a function $Y(s)$ is the function 
$y(t)$ satisfying ${\cal L}[y(t)](s) = Y(s)$, and is denoted by  
${\cal L}\inv[Y(s)]$.
\end{Def}\index{Laplace transform!inverse}

The summary of the Laplace transform method is: 
\begin{enumerate}
\item	Compute the Laplace transforms of both sides of the differential 
equation in \Ref{e:2ndforced} using the linearity of Laplace transforms, the 
formulas for Laplace transforms of derivatives \Ref{Tprops}(c), and specific 
Laplace transforms such as in \Ref{eq:Lexp}.
\item	Explicitly solve the transformed equation for 
${\cal L}[x(t)](s)=Y(s)$.
\item	Find a function $x(t)$ whose Laplace transform is $Y(s)$; that is, 
find ${\cal L}\inv[Y(s)]$.
\end{enumerate}

\subsubsection*{An Example of a First Order Forced Equation}

As an example, we use this three step method to solve the initial value 
problem:
\begin{equation}  \label{eq:dxx2}
\begin{array}{rcl}
\dot x - x & = & 2\\
  x(0) & = & 4.
\end{array}
\end{equation}

The first step is to apply the Laplace transform to both sides of the 
differential equation.  Using formula \Ref{Tprops}(c) and the linearity of 
${\cal L}$, we obtain
\[
\frac{2}{s} = {\cal L}[2] = {\cal L}[\dot x]-{\cal L}[x] = 
(s-1){\cal L}[x]-4.
\]
Note that $e^{0t}=1$ so that ${\cal L}[1] = \frac{1}{s-0}=\frac{1}{s}$.

The second step requires solving this equation explicitly for ${\cal L}[x]$
obtaining
\begin{equation}  \label{eq:Lapdxx2}
{\cal L}[x] = \frac{\frac{2}{s}+4}{s-1} = \frac{2+4s}{s(s-1)}
= \frac{6}{s-1} - \frac{2}{s}.
\end{equation}
Note that we have simplified the right hand side by use of partial fractions.

The third step requires using \Ref{eq:Lexp} to see that
\[
x(t) = {\cal L}\inv\left[\frac{6}{s-1} - \frac{2}{s}\right](t) = 
6{\cal L}\inv\left[\frac{1}{s-1}\right](t) -
2{\cal L}\inv\left[\frac{1}{s}\right](t) = 6e^t-2.
\]


\subsection*{Laplace Transforms of Second Order Equations}

We end this section with a discussion of the type of information that
we will need to solve initial value problems of second order inhomogeneous 
\index{inhomogeneous} linear equations by the method of Laplace transforms. 
Consider the differential equation
\begin{equation}  \label{e:2ndforced}
\begin{array}{rcl} 
\ddot{x} + a\dot{x} + bx & = & g(t)\\
x(0) & = & x_0\\
\dot{x}(0) & = & \dot{x}_0,
\end{array}
\end{equation}
where $a,b\in\R$ are constants.  The function $g(t)$ is called the 
{\em forcing\/} term. 

\subsubsection*{A Second Order Homogeneous Example}
\index{differential equation!higher order}

We begin our discussion by considering a simple example of an unforced 
equation:
\begin{equation}  \label{Lexam2}
\begin{array}{rcl}
\ddot{x} + 3\dot{x} +2x & = & 0  \\
 x(0) & = & 1 \\
\dot{x}(0) & = & 2.
\end{array}
\end{equation}
We begin by applying \Ref{Tprops}(c) twice to obtain
\[
\begin{array}{rcl}
{\cal L}[\ddot{x}](s) & = & s{\cal L}[\dot{x}](s) - \dot{x}(0)\\
& = & s(s{\cal L}[x](s) - x(0)) - \dot{x}(0)\\
& = & s^2{\cal L}[x](s) - sx(0) - \dot{x}(0).
\end{array}
\]
Where there is no ambiguity we will drop the argument $(s)$ in 
${\cal L}[x](s)$; this change should make subsequent formulas easier 
to read.  For instance, the previous equation is:
\begin{equation}  \label{eq:dderLap}
{\cal L}[\ddot{x}] = s^2{\cal L}[x] - sx(0) - \dot{x}(0).
\end{equation}
Now apply \Ref{Tprops} and \Ref{eq:dderLap} to the differential
equation \Ref{Lexam2} to obtain
\begin{eqnarray*}
0 & = & {\cal L}[\ddot{x} + 3\dot{x} +2x] \\
& = & (s^2{\cal L}[x] - sx(0) - \dot{x}(0)) + 
3(s{\cal L}[x] - x(0)) + 2{\cal L}[x]\\
& = & (s^2+3s+2){\cal L}[x] -(sx(0)+\dot{x}(0)+3x(0)) \\
& = & (s^2+3s+2){\cal L}[x] - (s+5).
\end{eqnarray*}

Next, use this equation and partial fractions to solve for ${\cal L}[x]$ as
\[
\dps{\cal L}[x] = \frac{s+5}{(s+1)(s+2)} = 4\frac{1}{s+1} - 3\frac{1}{s+2}.
\]
Using linearity and \Ref{eq:Lexp}, we see that 
\[
{\cal L}[4e^{-t}-3e^{-2t}] = 4\frac{1}{s+1} - 3\frac{1}{s+2}.
\]
Hence
\[
x(t) = 4e^{-t}-3e^{-2t}
\]
is the solution to \Ref{Lexam2}.


\subsubsection*{Information Needed to Solve Second Order Equations}

What information about Laplace transforms will we need to solve the 
initial value problem \Ref{e:2ndforced} in general?  We can answer this 
question just by taking the Laplace transform of \Ref{e:2ndforced} and 
solving for ${\cal L}[x]$.  Using \Ref{eq:dderLap} and \Ref{Tprops}(c), 
compute 
\begin{eqnarray*}
{\cal L}[g(t)] & = & {\cal L}[\ddot{x}+a\dot{x} + bx] \\
& = & {\cal L}[\ddot{x}] + a{\cal L}[\dot{x}]+b {\cal L}[x] \\ 
& = & (s^2{\cal L}[x]-sx_0-\dot{x}_0) +a(s{\cal L}[x]-x_0)+b{\cal L}[x]\\
& = & (s^2+as+b){\cal L}[x]-(x_0s+ax_0+\dot{x}_0).
\end{eqnarray*}
Then solve for ${\cal L}[x]$ as
\begin{equation}  \label{e:laplace2nd}
{\cal L}[x] = \frac{x_0s + ax_0+\dot{x}_0}{s^2+as+b} + \frac{G(s)}{s^2+as+b}
\end{equation}
where $G = {\cal L}[g(t)]$.  The third step in solving \Ref{e:2ndforced}
by Laplace transforms is to compute the inverse Laplace transform
\index{Laplace transform!inverse} of the right hand side of 
\Ref{e:laplace2nd}.  

To find the inverse Laplace transform of the first term, we use partial 
fractions.  For example, suppose that the polynomial $s^2+as+b$ 
has real distinct roots $r_1$ and $r_2$, then we can rewrite the first 
term as 
\[
\frac{c_1}{s-r_1} + \frac{c_2}{s-r_2}
\]
for some real constants $c_1$ and $c_2$.  We can now use \Ref{eq:Lexp}
to find the inverse Laplace transform of this first term.  

If this polynomial has a double real root or a complex conjugate pair of 
roots, then we need to find inverse Laplace transforms of functions like
\begin{equation}  \label{e:sampleLT}
\frac{1}{(s-r_1)^2}  \AND \frac{1}{(s-B)^2+C^2} \AND \frac{s}{(s-B)^2+C^2}.
\end{equation}

Finding the inverse Laplace transform for the second term is in general 
more difficult, since it depends on the hitherto unspecified function $g(t)$.
As it happens this inverse Laplace transform can be computed for a number 
of important functions, as we discuss in the next section.

To summarize: in order to use the method of Laplace transforms successfully
to solve forced second order linear equations, we must be able to
\begin{itemize}
\item	compute partial fraction\index{partial fractions} expansions,
\item	compute the inverse Laplace transforms
\index{Laplace transform!inverse} for functions in \Ref{e:sampleLT}, and
\item   compute the inverse Laplace transforms of the second term in 
\Ref{e:laplace2nd}, which involves the forcing function $g(t)$.
\end{itemize}


\EXER

\TEXER

\begin{exercise} \label{c13.1.0a}
Find the Laplace transform of $x(t)=e^{3t}-2e^{-4t}$.
\end{exercise}

\begin{exercise} \label{c13.1.0b}
Find the Laplace transform of $x(t)=14e^{-6t}+3e^{7t}$.
\end{exercise}

\begin{exercise}  \label{Ex:pf}
Given distinct roots $r_1$ and $r_2$, the method of partial fractions states 
that 
\begin{equation} \label{E:pfquad}
\frac{1}{(s-r_1)(s-r_2)} = \frac{a_1}{s-r_1} + \frac{a_2}{s-r_2}
\end{equation}
for some scalars $a_1$ and $a_2$.  By putting the right side of \Ref{E:pfquad}
over a common denominator, verify that $a_1$ and $a_2$ are found by solving 
the system of linear equations
\begin{eqnarray*}
a_1 + a_2 & = & 0\\
r_2a_1 + r_1a_2 & = & -1.
\end{eqnarray*}
\end{exercise}

\noindent In Exercises \ref{c13.1.0A} -- \ref{c13.1.0B} use partial fractions
to find a function $x(t)$ whose Laplace transform is the given function
$Y(s)$.  {\bf Hint:}  In your calculations, you may use the result of 
Exercise~\ref{Ex:pf}.
\begin{exercise} \label{c13.1.0A}
$Y(s) = \dps\frac{1}{s^2-4s+3}$.
\end{exercise}
\begin{exercise} \label{c13.1.0B}
$Y(s) = \dps\frac{s-1}{s^2+5s+6}$.
\end{exercise}


\noindent In Exercises~\ref{c13.1.1} -- \ref{c13.1.1a}, use Laplace transforms 
to compute the solution to the given initial value problem.
\begin{exercise} \label{c13.1.1}
$\ddot{x} + 3\dot{x} + 2x = 0, \quad x(0) = 1, \quad \dot{x}(0) = -1$. 
\end{exercise}
\begin{exercise} \label{c13.1.1a}
$\ddot{x} - 3\dot{x} - 4x = 0, \quad x(0) = 2, \quad \dot{x}(0) = -1$. 
\end{exercise}

\begin{exercise} \label{c13.4.4}
Let $\alpha,\beta,a,b,c$ be real numbers.  Show that the solution
of the initial value problem
\[
\ddot x +a\dot x +b x=c,\quad x(0)=\alpha,\quad \dot x(0)=\beta,
\]
has the Laplace transform
\[
{\cal L}[x] = \frac{\alpha s^2 +(\beta +a\alpha)s+c}{s(s^2+as+b)}.
\]
\end{exercise}

\begin{exercise}   \label{exer:kderLap}
Identity \Ref{eq:dderLap}\index{Laplace transform!derivative property} may be 
generalized to: 
\begin{equation}  \label{LapDerk}
{\cal L}\left[\frac{d^k y}{dt^k}\right] =
s^k{\cal L}[y] - s^{k-1} y(0) - \cdots - \frac{d^{k-1} y}{dt^{k-1}}(0).
\end{equation}
Use induction to verify \Ref{LapDerk}. 
\end{exercise}


\end{document}
