\documentclass{ximera}

\author{Matthew Carr \& Marty Golubitsky}

%\input{../../preamble.tex}
\input{../../preamble.tex}

\begin{document}


\begin{exercise}\label{mc.exerciseErr6_M}
%\RED{SECTION: Linear Dependence and Linear Independence}
Consider the matrix
 \[
A=\Matrix{1 & -6 & -1 \\ -3 & 0 & 2 \\ 0 & 1 & -2 \\ 0 & -2 & 4}\ldotp
\]
\begin{enumerate}[label=(\alph*)]
\item Show that the rows of $A$ are  linearly dependent.
\item Find two subsets $S_1$ and $S_2$ of rows of $A$ such that 
\begin{enumerate}[label=(\roman*)]
	\item $S_1\ne S_2$.
	\item $\mathrm{span}\,S_1=\mathrm{span}\,S_2$. 
	\item The vectors in $S_1$ are linearly independent and the vectors in $S_2$ are linearly independent. 
	\end{enumerate}
\end{enumerate}

\begin{solution}
\ans 
\begin{enumerate}[label=(\alph*)]
\item Let $\{R_1,R_2,R_3,R_4\}$ be the rows of $A$.  Linear dependence follows from 
\[
0R_1 + 0R_3+2R_3+R_4 = 0.
\] 
\item $S_1=\{R_1,R_2,R_3\}$ and $S_2=\{R_1,R_2,R_4\}$ satisfy the three conditions.
\end{enumerate}

\soln 

\begin{enumerate}[label=(\alph*)]

\item By inspection, $R_4=-2R_3$. This implies, by Lemma~\ref{L:lindep}, that the rows of $A$ are linearly dependent. 

\item Let $S_1=\{R_1,R_2,R_3\}$ and $S_2=\{R_1,R_2,R_4\}$.   
\begin{enumerate}
	\item The sets of vectors $S_1$ and $S_2$ are unequal.
	\item Verify that $\mathrm{span}\,S_1 = \mathrm{span}\,S_2$ as follows.  A linear combination of vectors in $S_1$ has the form
	\[
	r_1 R_1 + r_2 R_2 + r_3 R_3 = r_1 R_1 + r_2 R_2 - \frac{1}{2} r_3 R_4 
	\]
	which is a linear combination of vectors in $S_2$. 
	\item Use  Lemma~\ref{L:linindep} to see that $S_1$ is a linearly independent set of vectors. More precisely, the augmented matrix for the system of equations $r_1R_1+r_2 R_2+r_3 R_3=0$ row reduces to 
\[
\left(
\begin{array}{ccc|c}
 1 & 0 & 0 & 0 \\
 0 & 1 & 0& 0 \\
 0 & 0 & 1& 0 \\
\end{array}
\right)\ldotp
\]
Thus, $r_1 = r_2 = r_3 = 0$ so Lemma~\ref{L:linindep} implies $S_1$ is a linearly independent set of vectors. The same reasoning shows that $S_2$ is a linearly independent set of vectors.  Alternatively, since $R_4=-2R_3$, linear independence of $S_1$ implies linear independence of $S_2$.
\end{enumerate}
\end{enumerate}
\end{solution}
\end{exercise}

\end{document}