\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}
\begin{exercise} \label{c3.1.7}
As an application of \eqref{e:intcalcsoln} answer the following question
(posed by R.P. Agnew).
\begin{quote}
One day it started snowing at a steady rate.  A snowplow started at
noon and went two miles in the first hour and one mile in the second
hour.  Assume that the speed of the snowplow times the depth of the
snow is constant.  At what time did it start to snow?
\end{quote}

\noindent To set up this problem, let $d(t)$ be the depth of the snow
at time $t$ where $t$ is measured in hours and $t=0$ is noon.
Since the snow is falling at a constant rate $r$, $d(t)= r(t-t_0)$
where $t_0$ is the time that it started snowing.  Let $x(t)$ be the
position of the snowplow along the road.  The assumption that speed
times the depth equals a constant $k$ means that
\[
\frac{dx}{dt}(t) = \frac{k}{d(t)} = \frac{K}{t-t_0}
\]
where $K=k/r$.  The information about how far the snowplow goes in the
first two hours translates to
\[
x(1) = 2  \AND x(2) =3.
\]
Now solve the problem.

\begin{solution}

\ans The snow started falling at 11:23 am.

\soln Begin with the given equation
\[
\frac{dx}{dt} = \frac{K}{t-t_0}.
\]
In order to get a formula for $x(t)$, we take the integral of both sides of
this equation, obtaining
\[
x(t) = x(0) + \int_{0}^{t} \frac{K}{\tau-t_0} d\tau
\]
Note that $x(0) = 0$, since the snowplow started plowing at time $t=0$.  We
obtain by integration that
\[
x(t) = \left.K[\ln(\tau - t_0)]\right|_{0}^{t} = K(\ln|t - t_0| - \ln|-t_0|) = 
K\ln\left|\frac{t - t_0}{-t_0}\right| = K\ln\left|\frac{t_0-t}{t_0}\right|.
\]
We are given two values for $x$, namely, $x(1) = 2$ and $x(2) = 3$.  We can
substitute these values into the formula for $x(t)$ to get the system of 
equations
\[
\begin{array}{rcl}
2 & = & K\ln\left|\frac{t_0 - 1}{t_0}\right| \\
3 & = & K\ln\left|\frac{t_0 - 2}{t_0}\right|.
\end{array}
\]
Solving the second equation for $K$ and substituting into the first equation 
gives
\[
2\ln\left|\frac{t_0 - 2}{t_0}\right| = 3\ln\left|\frac{t_0 - 1}{t_0}\right|.
\]
Next take the exponential of both sides, then expand and solve for $t_0$. 
That is
\[
\begin{array}{rcl}
\left(\frac{t_0 - 2}{t_0}\right)^2 & = & \left(\frac{t_0 - 1}{t_0}\right)^3 \\
t_0\left(t_0 - 2\right)^2 & = & \left(t_0 - 1\right)^3 \\
0 & = & t_0^2 - t_0 - 1 \end{array}
\]
Note that $t_0 < 0$, since the snow began falling before the snowplow
started.  Hence
\[
t_0 = \frac{1 - \sqrt{5}}{2} \approx -0.618.
\]
So the snow started falling at $t_0 \approx -37$ minutes, that is, at
11:23 am.


\end{solution}
\end{exercise}
\end{document}
