\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}
\begin{computerExercise} \label{c3.1.9}
There are two banks in town --- Intrastate and Statewide.  You plan
to deposit \$5,000 in one of these banks for two years.  Statewide Bank's
best savings account pays $8\%$ interest per year compounded quarterly
and charges \$10 to open an account.  Intrastate Bank's best savings
account pays $7.75\%$ interest compounded daily.  Which bank will
pay you the most money when you withdraw your money?  Would your
answer change if you had planned to keep your money in the bank
for only one year?

\begin{solution}

\ans After two years, an account at Statewide returns more interest than
the Intrastate account.  However, after only one year, the Intrastate
account returns more.

\soln To find the return from each account, use the formula for
compound interest:
\[
P(t) = P_0\left(1 + \frac{r}{N}\right)^{Nt}
\]
If you deposit your money at Statewide, $P_0 = \$4\mbox{,}990$ because of the
fine.  Interest is compounded four times a year, so $N = 4$.  The interest
rate is 8\%, so $r = 0.08$.
After two years, an account at Statewide returns
\[
P_{S}(2) = \$4\mbox{,}990\left(1 + \frac{0.08}{4}\right)^{2(4)} = 
\$5\mbox{,}846.58
\]
If you deposit your money at Intrastate, $r = 0.0775$.  Interest is
compounded daily, so $N = 365$.  So, after two years, the account returns
\[
P_{I}(2) = \$5\mbox{,}000\left(1 + \frac{0.0775}{365}\right)^{2(365)} = 
\$5\mbox{,}838.19.
\]
After only 1 year, however,
\[
P_{S}(1) = \$4\mbox{,}900\left(1 + \frac{0.08}{4}\right)^{4} = \$5\mbox{,}401.37
\]
at Statewide, and
\[
P_{I}(1) = \$5\mbox{,}000\left(1 + \frac{0.0775}{365}\right)^{365} = 
\$5\mbox{,}402.87
\]
at Intrastate.

\end{solution}
\end{computerExercise}
\end{document}
