\documentclass{ximera}

\input{../preamble.tex}

\title{Spanning Sets and MATLAB}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

 \label{S:5.3}

In this section we discuss:
\begin{itemize}
\item	how to find a spanning set for the subspace of solutions to a
homogeneous system of linear equations using the \Matlab command {\tt null},
and
\item	how to determine when a vector is in the subspace spanned by a
set of vectors using the \Matlab command {\tt rref}.
\end{itemize}

\subsection*{Spanning Sets for Homogeneous Linear Equations}

In Chapter~\ref{lineq} we saw how to use Gaussian elimination,
back substitution, and \Matlab to compute solutions to a system
of linear equations.  For systems of
homogeneous equations\index{homogeneous}, \Matlab
provides a command to find a spanning set for the subspace of solutions.
That command is {\tt null}.  For example, if we type
\begin{verbatim}
A = [2 1 4 0; -1 0 2 1]
B = null(A)
\end{verbatim} \index{\computer!null}
then we obtain
\begin{verbatim}
B =
    0.4830         0
   -0.4140    0.8729
   -0.1380   -0.2182
    0.7591    0.4364
\end{verbatim}
The two columns of the matrix $B$ span the set of solutions of
the equation $Ax=0$.  In particular, the vector $(2,-8,1,0)$ is a
solution to $Ax=0$ and is therefore a
linear combination \index{linear!combination} of the
column vectors of {\tt B}.  Indeed, type
\begin{verbatim}
4.1404*B(:,1)-7.2012*B(:,2)
\end{verbatim}
and observe that this linear combination is the desired one.

Next we describe how to find the coefficients {\tt 4.1404} and
{\tt -7.2012} by showing that these coefficients themselves are
solutions to another system of linear equations.

\subsection*{When is a Vector in a Span?} \index{span}

Let $w_1,\ldots,w_k$ and $v$ be vectors in $\R^n$.  We now
describe a method that allows us to decide whether $v$ is in
$\Span\{w_1,\ldots,w_k\}$.  To answer this question one has
to solve a system of $n$ linear equations in $k$ unknowns.
The unknowns correspond to the coefficients in the linear
combination of the vectors $w_1,\ldots,w_k$ that gives $v$.

Let us be more precise.  The vector\index{vector} $v$ is in
$\Span\{w_1,\ldots,w_k\}$ if and only if there are constants
$r_1,\ldots,r_k$ such that the equation
\begin{equation}  \label{e:lindepeqn}
     r_1 w_1 + \cdots + r_k w_k = v
\end{equation}
is valid.  Define the $n\times k$ matrix $A$ as the one having
$w_1,\ldots,w_k$ as its columns; that is,
\begin{equation}  \label{E:Abycol}
A = (w_1| \cdots |w_k).
\end{equation}
Let  $r$ be the $k$-vector
\[
r= \left(\begin{array}{c} r_1 \\ \vdots \\ r_k\end{array}\right).
\]
Then we may rewrite equation \eqref{e:lindepeqn} as
\begin{equation}  \label{E:Ar=v}
   Ar=v.
\end{equation}
To summarize:
\begin{lemma}
Let $w_1,\ldots,w_k$ and $v$ be vectors in $\R^n$.  Then $v$
is in $\Span\{w_1,\ldots,w_k\}$ if and only if the system of linear
equations \eqref{E:Ar=v} has a solution where $A$ is the $n\times k$
defined in \eqref{E:Abycol}.
\end{lemma}

To solve \eqref{E:Ar=v} we row reduce the
augmented matrix\index{matrix!augmented} $(A|v)$.
For example, is $v=(2,1)$ in the span of $w_1=(1,1)$ and $w_2=(1,-1)$?
That is, do there exist scalars $r_1,r_2$ such that
\[
r_1\vectwo{1}{1} + r_2\vectwo{1}{-1} = \vectwo{2}{1}?
\]
As noted, we can rewrite this equation as
\[
\mattwo{1}{1}{1}{-1}\vectwo{r_1}{r_2} = \vectwo{2}{1}.
\]
We can solve this equation by row reducing the augmented
matrix
\[
\left(\begin{array}{rr|r}
1 & 1 & 2 \\ 1 & -1 & 1 \end{array}\right)
\]
to obtain
\[
\left(\begin{array}{rr|r}
1 & 0 & \frac{3}{2} \\ 0 & 1 & \frac{1}{2}
\end{array}\right).
\]
So $v = \frac{3}{2}w_1 + \frac{1}{2}w_2$.

Row reduction to reduced echelon form\index{echelon form!reduced}
has been preprogrammed in the
\Matlab command {\tt rref}. \index{\computer!rref}  Consider the
following example.  Let
\begin{equation}  \label{e:w1w2}
     w_1=(2,0,-1,4) \AND w_2=(2,-1,0,2)
\end{equation}
and ask the question whether $v=(-2,4,-3,4)$ is in $\Span\{w_1,w_2\}$.

In \Matlab load the matrix $A$ having $w_1$ and
$w_2$ as its columns and the vector $v$ by typing {\tt e5\_3\_5}
\begin{matlabEquation}  \label{e:Aandv}
A=\left(\begin{array}{rr} 2 & 2 \\ 0 & -1 \\ -1 & 0 \\ 4 & 2
\end{array}\right) \AND
v=\left(\begin{array}{r} -2 \\ 4 \\ -3 \\ 4 \end{array}\right).
\end{matlabEquation}%
We can solve the system of equations using \Matlabp.
First, form the augmented matrix by typing
\begin{verbatim}
aug = [A v]
\end{verbatim}
Then solve the system by typing {\tt rref(aug)} to obtain
\begin{verbatim}
ans =
     1   0   3
     0   1  -4
     0   0   0
     0   0   0
\end{verbatim}
It follows that $(r_1,r_2)=(3,-4)$ is a solution and $v=3w_1-4w_2$.

Now we change the $4^{th}$ entry in $v$ slightly by typing
{\tt v(4) = 4.01}.  There is no solution to the system of equations
\[
Ar = \left(\begin{array}{r} -2 \\ 4 \\ -3 \\ 4.01
\end{array}\right)
\]
as we now show.  Type
\begin{verbatim}
aug = [A v]
rref(aug)
\end{verbatim}
which yields
\begin{verbatim}
ans =
     1    0    0
     0    1    0
     0    0    1
     0    0    0
\end{verbatim}
This matrix corresponds to an inconsistent\index{inconsistent} system;
thus $v$ is no longer in the span\index{span} of $w_1$ and $w_2$.

\EXER

\CEXER

\noindent In Exercises~\ref{c5.3.1a} -- \ref{c5.3.1c} use the {\tt null}
command in \Matlab to find all the solutions of the linear system of
equations $Ax=0$.  \index{\computer!null}
\begin{exercise} \label{c5.3.1a}
\begin{matlabEquation} \label{e:BCDa}
          A=    \left(\begin{array}{cccc}
                -4 & 0 & -4 & 3\\
                -4 & 1 & -1 & 1
                \end{array}\right) \quad
\end{matlabEquation}

\begin{solution}

Type {\tt null(A)} in \Matlab to find that the set of solutions to
$Ax = 0$ is spanned by the vectors
\[
\left(\begin{array}{r} 0.3225 \\ 0.8931 \\ -0.0992 \\ 0.2977
\end{array}\right) \AND \left(\begin{array}{r} 0 \\ -0.1961 \\
0.5883 \\ 0.7845 \end{array}\right).
\]

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.3.1b}
\begin{matlabEquation} \label{e:BCDb}
           A=    \left(\begin{array}{rr}
                1 & 2 \\
                1 & 0 \\
                3 & -2
                \end{array}\right) \quad
\end{matlabEquation}

\begin{solution}
\ans The solution to $Ax = 0$ is the vector $(0,0).$

\soln This can be demonstrated by typing {\tt null(A)} in \Matlab, which
yields
\begin{verbatim}
ans =
   Empty matrix: 2-by-0
\end{verbatim}

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.3.1c}
\begin{matlabEquation} \label{e:BCDc}
          A=      \left(\begin{array}{rrr}
               1  &  1  &  2\\
              -1  &  2  & -1
                \end{array}\right).
\end{matlabEquation}

\begin{solution}
The set of solutions to $Ax = 0$ is spanned by the vector
\[
\vecthree{-0.8452}{-0.1690}{0.5071}.
\]


\end{solution}
\end{exercise}

\begin{exercise} \label{c5.3.2}
Use the {\tt null} command in \Matlab to verify your answers to
Exercises~\ref{c5.2.2a} and \ref{c5.2.2b}.

\begin{solution}

Enter matrix {\tt A} into \Matlab and type {\tt null(A)}, obtaining
\begin{verbatim}
ans =
   -0.8957         0
    0.4414    0.1204
   -0.0519    0.9631
    0.0130   -0.2408
         0         0
\end{verbatim}
The \Matlab answer is a valid solution if the vectors found
by row reduction can be written as linear combinations of the \Matlab
answers.  In \Matlab, row reduce the augmented matrices {\tt null(A)|x},
and {\tt null(A)|y} where $x = (-2,1,0,0,0)$ and $y = (-1,0,-4,1,0)$ 
to find that
\[
\left(\begin{array}{r} -2 \\ 1 \\ 0 \\ 0 \\ 0 \end{array}\right) =
2.2328\left(\begin{array}{r} -0.8957 \\ 0.4414 \\ -0.0519 \\ 0.0130 \\ 0
\end{array}\right) + 0.1204\left(\begin{array}{r} 0 \\ 0.1204 \\ 0.9631 \\
-0.2408 \\ 0 \end{array}\right)
\]
and
\[
\left(\begin{array}{r} -1 \\ 0 \\ -4 \\ 1 \\ 0 \end{array}\right) =
1.1164\left(\begin{array}{r} -0.8957 \\ 0.4414 \\ -0.0519 \\ 0.0130 \\ 0
\end{array}\right) - 4.0931\left(\begin{array}{r} 0 \\ 0.1204 \\ 0.9631 \\
-0.2408 \\ 0 \end{array}\right).
\]

Enter matrix {\tt B} into \Matlab and type {\tt null(B)}, obtaining
\begin{verbatim}
ans =
   -0.9661         0
    0.2070   -0.5976
   -0.1380   -0.7171
    0.0690    0.3586
\end{verbatim}
Again, the \Matlab answer can be checked by verifying by row reduction
of the augmented matrix that the vectors $(-3,1,0,0)$ and $(-5,0,-2,1)$
can be written as linear combinations of the \Matlab solution vectors.
In particular,
\[
\left(\begin{array}{r} -3 \\ 1 \\ 0 \\ 0 \end{array}\right) =
3.1053\left(\begin{array}{r} -0.9661 \\ 0.2070 \\ -0.1380 \\ 0.0690
\end{array}\right) - 0.5976\left(\begin{array}{r} 0 \\ -0.5976 \\ -0.7171 \\
0.3586 \end{array}\right)
\]
and
\[
\left(\begin{array}{r} -5 \\ 0 \\ -2 \\ 1 \end{array}\right) =
5.1755\left(\begin{array}{r} -0.9661 \\ 0.2070 \\ -0.1380 \\ 0.0690
\end{array}\right) + 1.7928\left(\begin{array}{r} 0 \\ -0.5976 \\ -0.7171 \\
0.3586 \end{array}\right).
\]

\end{solution}
\end{exercise}

\begin{exercise} \label{c5.3.3}
Use row reduction to find the solutions to $Ax=0$ where $A$ is
given in \eqref{e:BCDa}.  Does your answer agree with the \Matlab
answer using {\tt null}?  If not, explain why.

\begin{solution}

\ans The solution set of $Bx = 0$ is
\[
\left(\begin{array}{r} x_1 \\ x_2 \\ x_3 \\ x_4 \end{array}\right)
= \left(\begin{array}{c} -x_3 + \frac{3}{4}x_4 \\ -3x_3 + 2x_4 \\
x_3 \\ x_4 \end{array}\right) = x_3\left(\begin{array}{r} -1 \\ -3 \\
1 \\ 0 \end{array}\right) + x_4\left(\begin{array}{r} \frac{3}{4} \\
2 \\ 0 \\ 1 \end{array}\right).
\]

\soln Row reduce $B$:
\[
\left(\begin{array}{rrrr} -4 & 0 & 4 & 3 \\ -4 & 1 & -1 & 1
\end{array}\right) \longrightarrow \left(\begin{array}{rrrr}
1 & 0 & 1 & -\frac{3}{4} \\ 0 & 1 & 3 & -2 \end{array}\right).
\]

The solution obtained by row reduction is not the same as the one
obtained using {\tt null}, but the solution vectors are linear
combinations of the \Matlab solution vectors, so the answers are
equivalent.  By row reducing the matrix {\tt [null(B) x]}, where
$x = (-1,-3,1,0)$, we find that
\[
\left(\begin{array}{r} -1 \\ -3 \\ 1 \\ 0 \end{array}\right) =
-3.1009\left(\begin{array}{r} 0.3225 \\ 0.8931 \\ -0.0992 \\ 0.2977
\end{array}\right) + 1.1767\left(\begin{array}{r} 0 \\ -0.1961 \\
0.5883 \\ 0.7845 \end{array}\right).
\]
By row reducing the matrix {\tt [null(B) y]} where $y = (\frac{3}{4},
2,0,1)$ we find that:
\[
\left(\begin{array}{r} \frac{3}{4} \\ 2 \\ 0 \\ 1 \end{array}\right) =
2.3257\left(\begin{array}{r} 0.3225 \\ 0.8931 \\ -0.0992 \\ 0.2977
\end{array}\right) + 0.3922\left(\begin{array}{r} 0 \\ -0.1961 \\
0.5883 \\ 0.7845 \end{array}\right).
\]

\end{solution}
\end{exercise}

\noindent In Exercises~\ref{c5.3.4a} -- \ref{c5.3.4c} let $W\subset\R^5$
be the subspace spanned by the vectors
\begin{matlabEquation}\label{MATLAB:65}
     w_1=(2,0,-1,3,4),\quad w_2=(1,0,0,-1,2),\quad w_3=(0,1,0,0,-1).
\end{matlabEquation}
Use \Matlab to decide whether the given vectors are elements of $W$.
\begin{exercise} \label{c5.3.4a}
$v_1=(2,1,-2,8,3)$.

\begin{solution}
\ans Vector $v_1$ is an element of $W$.

\soln The vector $v_1$ is an element of $W$ if there exist scalars $a$,
$b$, and $c$ such that
\[
aw_1 + bw_2 + cw_3 = v_1.
\]
Using \Matlab, create the matrix {\tt A = [w1' w2' w3']}, which has
$w_1$, $w_2$, and $w_3$ as its columns.  Then create the augmented
matrix {\tt aug1 = [A v1']}.  The command {\tt rref(aug1)} yields
\begin{verbatim}
ans =
     1     0     0     2
     0     1     0    -2
     0     0     1     1
     0     0     0     0
     0     0     0     0
\end{verbatim}
Since there is no pivot point in the last column, the linear system
$aw_1 + bw_2 + cw_3 = v_1$ is consistent, and $v_1 = 2w_1 - 2w_2 + w_3$.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.3.4b}
$v_2=(-1,12,3,-14,-1)$.

\begin{solution}
\ans Vector $v_2$ is not an element of $W$.

\soln Create the augmented matrix {\tt aug2 = [A v2']}.  Row reducing
{\tt aug2} yields
\begin{verbatim}
ans =
     1     0     0     0
     0     1     0     0
     0     0     1     0
     0     0     0     1
     0     0     0     0
\end{verbatim}
There is a pivot point in the last column, so the linear system
$aw_1 + bw_2 + cw_3 = v_2$ is inconsistent.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.3.4c}
$v_3=(-1,12,3,-14,-14)$.

\begin{solution}
\ans Vector $v_3$ is an element of $W$.

\soln Row reduce the augmented matrix {\tt aug3 = [A v3']} to obtain
\begin{verbatim}
ans =
     1     0     0    -3
     0     1     0     5
     0     0     1    12
     0     0     0     0
     0     0     0     0
\end{verbatim}
which is the matrix of a consistent linear system.  Therefore,
$v_3 = -3w_1 + 5w_2 + 12w_3$.



\end{solution}
\end{exercise}

\end{document}
