\documentclass{ximera}

\input{../preamble.tex}

\title{Vector Spaces}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{C:vectorspaces}

\normalsize

In Chapter~\ref{lineq} we discussed how to solve systems of $m$
linear equations in $n$ unknowns.  We found that solutions of
these equations are vectors $(x_1,\ldots,x_n)\in\R^n$.  
In Chapter~\ref{chap:matrices} we discussed how the notation of
matrices and matrix multiplication drastically simplifies the
presentation of linear systems and how matrix multiplication
leads to linear mappings.  We also discussed briefly how linear
mappings lead to methods for solving linear systems ---
superposition, eigenvectors, inverses.  In
Chapter~\ref{chap:SolveOdes} we discussed how to solve systems
of $n$ linear differential equations in $n$ unknown functions.
These chapters have
provided an introduction to many of the ideas of linear algebra
and now we begin the task of formalizing these ideas.

Sets having the two operations of vector addition and scalar
multiplication are called {\em vector spaces\/}.  This concept
is introduced in Section~\ref{S:5.1} along with the two
primary examples --- the set $\R^n$ in which solutions to systems
of linear equations sit and the set $\CCone$ of differentiable
functions in which solutions to systems of ordinary differential
equations sit.  Solutions to systems of homogeneous linear equations
form subspaces of $\R^n$ and solutions of systems of linear
differential equations form subspaces of $\CCone$.  These issues
are discussed in Sections~\ref{S:5.1} and \ref{S:5.2}.

When we {\em solve\/} a homogeneous system of equations, we write
every solution as a superposition of a finite number of specific
solutions.  Abstracting this process is one of the main points of this
chapter.  Specifically, we show that every vector in many commonly
occurring vector spaces
(in particular, the subspaces of solutions) can be written as a
{\em linear combination\/} (superposition) of a few 
solutions.  The minimum number of solutions needed is called
the {\em dimension\/} of that vector space.  Sets of vectors that 
generate all solutions by superposition and that consist of that minimum 
number of vectors are called {\em bases\/}.  These ideas are discussed
in detail in Sections~\ref{S:5.3}--\ref{S:5.5}.  The proof of
the main theorem (Theorem~\ref{basis=span+indep}), which gives a
computable method for determining when a set is a basis, is given in
Section~\ref{S:5.6}.  This proof may be omitted on a first reading,
but the statement of the theorem is most important and must be
understood.


\end{document}
