\documentclass{ximera}

\input{../preamble.tex}

\title{Vector Spaces and Subspaces}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

 \label{S:5.1}

Vector spaces abstract the arithmetic properties of addition and
scalar multiplication of vectors.  In $\R^n$ we know how to add
vectors and to multiply vectors by scalars.  Indeed, it is
straightforward to verify that each of the eight
properties listed in Table~\ref{vectorspacelist} is valid for
vectors in $V=\R^n$.  Remarkably, sets that satisfy these eight
properties have much in common with $\R^n$.  So we define:
\begin{definition}  \label{vectorspace}
Let $V$ be a set having the two operations of addition and scalar
multiplication.  Then $V$ is a {\em vector space\/} if the eight
properties listed in Table~\ref{vectorspace} hold.  The elements
of a vector space are called {\em vectors}\index{vector}.
\end{definition} \index{vector!space}\index{vector!space}

The vector $0$ mentioned in (A3) in Table~\ref{vectorspacelist}
is called the {\em zero vector}\index{zero vector}.

\begin{table*}
\caption{Properties of Vector Spaces: suppose $u,v,w\in V$ and
$r,s\in\R$.}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
(A1) & Addition is commutative & $v+w=w+v$ \\
\hline
(A2) & Addition is associative & $(u+v)+w = u+(v+w)$ \\
\hline
(A3) & Additive identity $0$ exists & $v+0=v$ \\
\hline
(A4) & Additive inverse $-v$ exists & $v+(-v) = 0$ \\
\hline
(M1) & Multiplication is associative & $(rs)v = r(sv)$ \\
\hline
(M2) & Multiplicative identity exists & $1v=v$ \\
\hline
(D1) & Distributive law for scalars & $(r+s)v = rv+sv$ \\
\hline
(D2) & Distributive law for vectors & $r(v+w) = rv+rw$ \\
\hline
\end{tabular}
\end{center} \index{scalar multiplication} \index{vector!addition}
\index{associative} \index{distributive}
\index{commutative} \index{inverse}
\label{vectorspacelist}
\end{table*}

When we say that a vector space $V$ has the two operations of
addition and scalar multiplication we mean that the sum of two
vectors in $V$ is again a vector in $V$ and the scalar product
of a vector with a number is again a vector in $V$.  These two
properties are called {\em closure under addition\/} and
{\em closure under scalar multiplication}.
\index{closure!under addition}
\index{closure!under scalar multiplication}

In this discussion we focus on just two types of vector spaces:
$\R^n$ and function spaces.  The reason that we make this choice
is that solutions to linear equations are vectors in $\R^n$ while
solutions to linear systems of differential equations are vectors
of functions.

\subsubsection*{An Example of a Function Space}
\index{function space}

For example, let ${\cal F}$ denote the set of all functions $f:\R\to\R$.
Note that functions like $f_1(t)=t^2-2t+7$ and $f_2(t)=\sin t$ are in
${\cal F}$ since they are defined for all real numbers $t$, but that functions
like $g_1(t)=\frac{1}{t}$ and $g_2(t)=\tan t$ are not in ${\cal F}$ since
they are not defined for all $t$.

We can add two functions $f$ and $g$ by defining the function $f+g$ to be:
\[
(f+g)(t) = f(t) + g(t).
\]
We can also multiply a function $f$ by a scalar $c\in\R$ by defining the
function $cf$ to be:
\[
(cf)(t) = cf(t).
\]
With these operations of addition and scalar multiplication, ${\cal F}$
is a vector space; that is, ${\cal F}$ satisfies the eight vector space
properties in Table~\ref{vectorspacelist}.  More precisely:
\begin{itemize}
\item[(A3)] Define the zero function ${\cal O}$ by
\[
     {\cal O}(t) = 0 \quad \mbox{for all $t\in\R$.}
\]
For every $x$ in ${\cal F}$ the function ${\cal O}$ satisfies:
\[
     (x+{\cal O})(t) = x(t) + {\cal O}(t) = x(t) + 0 = x(t).
\]
Therefore, $x+{\cal O}=x$ and ${\cal O}$ is the additive
identity in ${\cal F}$.
\item[(A4)] Let $x$ be a function in ${\cal F}$ and define $y(t)=-x(t)$.
Then $y$ is also a function in ${\cal F}$, and
\[
    (x+y)(t) =  x(t) + y(t) = x(t) + (-x(t)) = 0 = {\cal O}(t).
\]
Thus, $x$ has the additive inverse $-x$.
\end{itemize}
After these comments it is straightforward to verify that the
remaining six properties in Table~\ref{vectorspacelist} are
satisfied by functions in ${\cal F}$.

\subsubsection*{Sets that are not Vector Spaces}
\index{vector!space}\index{vector!space}

It is worth considering how closure under vector addition and
scalar multiplication can fail.  Consider the following three
examples.

\begin{enumerate}
\item[(i)] Let $V_1$ be the set that consists of just the $x$
and $y$ axes in the plane.  Since $(1,0)$ and $(0,1)$ are in
$V_1$ but
\[
(1,0) + (0,1) = (1,1)
\]
is not in $V_1$, we see that $V_1$ is not closed under vector
addition. On the other hand, $V_1$ is closed under scalar
multiplication.  \index{closure!under addition}
\index{closure!under scalar multiplication}


\item[(ii)] Let $V_2$ be the set of all vectors $(k,\ell)\in\R^2$
where $k$ and $\ell$ are integers.  The set $V_2$ is closed under
addition but not under scalar multiplication since
$\half(1,0)=(\half,0)$ is not in $V_2$.

\item[(iii)] Let $V_3=[1,2]$ be the closed interval in $\R$. The
set $V_3$ is neither closed under addition ($1+1.5=2.5\not\in
V_3$) nor under scalar multiplication ($4\cdot 1.5 = 6\not\in
V_3$).  Hence the set $V_3$ is not closed under vector addition
and not closed under scalar multiplication.
\end{enumerate}

\subsection*{Subspaces}

\begin{definition} \label{subspaces}
Let $V$ be a vector space.  A nonempty subset $W\subset V$ is a
{\em subspace\/} if $W$ is a vector space using the operations of 
addition and scalar multiplication defined on $V$.
\end{definition} \index{subspace}

Note that in order for a subset $W$ of a vector space $V$ to be a 
subspace it must be {\em closed under addition\/} and {\em closed under
scalar multiplication\/}.  That is, suppose $w_1,w_2\in W$ and 
$r\in\R$.  Then
\begin{itemize}
\item[(i)]  $w_1+w_2 \in W$, and
\item[(ii)]  $rw_1\in W$.
\end{itemize}
\index{vector!addition} \index{scalar multiplication}

The $x$-axis and the $xz$-plane are examples of subsets of $\R^3$
that are closed under addition and closed under scalar multiplication.
Every vector on the $x$-axis has the form $(a,0,0)\in\R^3$.
The sum of two vectors $(a,0,0)$ and $(b,0,0)$ on the $x$-axis is
$(a+b,0,0)$ which is also on the $x$-axis.  The $x$-axis
is also closed under scalar multiplication as $r(a,0,0)=(ra,0,0)$,
and the $x$-axis is a subspace of $\R^3$.  Similarly, every vector
in the $xz$-plane in $\R^3$ has the form $(a_1,0,a_3)$. As in the
case of the $x$-axis, it is easy to verify that this set of vectors
is closed under addition and scalar multiplication.  Thus, the
$xz$-plane is also a subspace of $\R^3$.

In Theorem~\ref{T:subspaces} we show that every subset of a vector space 
that is closed under addition and scalar multiplication is a subspace.  
To verify this statement, we need the following lemma in which some special 
notation is used.  Typically, we use the same notation $0$ to denote the
real number zero and the zero vector\index{zero vector}.  In the following
lemma it is convenient to distinguish the two different uses of $0$, and we 
write the zero vector in boldface.

\begin{lemma}  \label{lem:AddId}
Let $V$ be a vector space, and let ${\bf 0}\in V$ be the zero vector.  Then
\[
0v={\bf 0} \quad \mbox{and}\quad (-1)v=-v
\]
for every vector in $v\in V$.
\end{lemma}

\begin{proof} Let $v$ be a vector in $V$ and use (D1) to compute
\[
     0v + 0v  = (0+0)v = 0v.
\]
By (A4) the vector $0v$ has an additive inverse $-0v$.  Adding $-0v$ 
to both sides yields 
\[
(0v + 0v) + (-0v) = 0v + (-0v) = {\bf 0}.
\] 
Associativity of addition (A2) now implies 
\[
0v + (0v + (-0v)) = {\bf 0}.
\]
A second application of (A4) implies that 
\[
0v + {\bf 0} = {\bf 0}
\]
and (A3) implies that $0v={\bf 0}$.

Next, we show that the additive inverse $-v$ of a vector $v$ is unique. 
That is, if $v + a = {\bf 0}$, then $a=-v$.  

Before beginning the proof, note that commutativity of addition (A1) together 
with (A3) implies that ${\bf 0}+v=v$.  Similarly, (A1) and (A4) imply that 
$-v+v={\bf 0}$.  

To prove uniqueness of additive inverses, add $-v$ to both sides of the 
equation $v + a = {\bf 0}$ yielding 
\[
-v + (v + a) = -v + {\bf 0}. 
\]
Properties (A2) and (A3) imply
\[
(-v + v) + a = -v.
\]
But 
\[
(-v + v) + a = {\bf 0} + a = a.
\]
Therefore $a = -v$, as claimed.

To verify that $(-1)v = -v$, we show that $(-1)v$ is the
additive inverse of $v$.  Using (M1), (D1), and the fact that $0v = {\bf 0}$, 
calculate
\[
     v + (-1)v = 1v + (-1)v =(1-1)v = 0v = {\bf 0}.
\]
Thus, $(-1)v$ is the additive inverse of $v$ and must equal $-v$, as
claimed.  \end{proof}

\begin{theorem}  \label{T:subspaces}
Let $W$ be a subset of the vector space $V$. If $W$ is closed under addition 
and closed under scalar multiplication, then $W$ is a subspace.
\end{theorem}\index{subspace}
\begin{proof}  We have to show that $W$ is a vector space using the operations 
of addition and scalar multiplication defined on $V$.  That is, we need to
verify that the eight properties listed in Table~\ref{vectorspacelist} are 
satisfied.  Note that properties (A1), (A2), (M1), (M2), (D1), and (D2) are 
valid for vectors in $W$ since they are valid for vectors in $V$.

It remains to verify (A3) and (A4).  Let $w\in W$ be any vector.
Since $W$ is closed under scalar multiplication, it follows that
$0w$ and $(-1)w$ are in $W$. Lemma~\ref{lem:AddId} states that
$0w=0$ and $(-1)w=-w$; it follows that $0$ and $-w$ are in $W$.
Hence, properties (A3) and (A4) are valid for vectors in $W$,
since they are valid for vectors in $V$.  \end{proof}

\subsubsection*{Examples of Subspaces of $\R^n$}

\begin{example}  \label{EX:subspaces}
{\rm
\begin{itemize}
\item[(a)] Let $V$ be a vector space.  Then the subsets $V$ and $\{0\}$ are
always subspaces of $V$.  A subspace $W\subset V$ is {\em proper\/} if
$W\neq 0$ and $W\neq V$. \index{subspace!proper}
\item[(b)] Lines through the origin are subspaces of $\R^n$.  Let $w\in \R^n$
be a nonzero vector and let $W=\{rw:r\in\R\}$.  The set $W$ is closed under
addition and scalar multiplication and is a subspace of $\R^n$ by 
Theorem~\ref{T:subspaces}.
The subspace $W$ is just a {\em line through the origin\/} in $\R^n$, since
the vector $rw$ points in the same direction as $w$ when $r>0$ and the exact
opposite direction when $r<0$.
\item[(c)]  Planes containing the origin are subspaces of $\R^3$.  To verify
this point, let $P$ be a plane through the origin and let $N$ be a vector
perpendicular\index{perpendicular} to $P$.  Then $P$ consists of all vectors
$v\in\R^3$ perpendicular to $N$; using the dot-product (see Chapter~\ref{lineq},
\eqref{XX_0}) we recall that such vectors satisfy the linear equation
$N\cdot v = 0$.  By superposition\index{superposition}, the set of all
solutions to this equation is closed under addition and scalar multiplication
and is therefore a subspace by Theorem~\ref{T:subspaces}.
\end{itemize}
}
\end{example}
In a sense that will be made precise all subspaces of $\R^n$ can be written
as the span of a finite number of vectors generalizing
Example~\ref{EX:subspaces}(b) or as solutions to a system of linear equations
generalizing Example~\ref{EX:subspaces}(c).


\subsubsection*{Examples of Subspaces of the Function Space ${\cal F}$}
\index{function space!subspace of}\index{subspace!of function space}

Let ${\cal P}$ be the set of all polynomials in ${\cal F}$.  The sum of
two polynomials is a polynomial and the scalar multiple of a polynomial is a
polynomial.  Thus, ${\cal P}$ is closed under addition and scalar
multiplication, and ${\cal P}$ is a subspace of ${\cal F}$.

As a second example of a subspace of ${\cal F}$, let $\CCone$ be the set of
all continuously differentiable functions $u:\R\to\R$.  A function $u$ is
in $\CCone$ if $u$ and $u'$ exist and are continuous for all $t\in\R$.
Examples of functions in $\CCone$ are:
\begin{enumerate}
\item[(i)] Every polynomial $p(t)=a_mt^m+a_{m-1}t^{m-1}+\cdots
+a_1t+a_0$ is in $\CCone$. \index{polynomial}
\item[(ii)]  The function $u(t)=e^{\lambda t}$ is in $\CCone$
for each constant $\lambda\in\R$.
\item[(iii)]  The trigonometric functions $u(t)=\sin(\lambda t)$
and $v(t)=\cos(\lambda t)$ are in $\CCone$ for each constant
$\lambda\in\R$. \index{trigonometric function}
\item[(iv)] $u(t)=t^{7/3}$ is twice differentiable everywhere and is
in $\CCone$.
\end{enumerate}
Equally there are many commonly used functions that are not in
$\CCone$.  Examples include:
\begin{enumerate}
\item[(i)] $u(t)=\frac{1}{t-5}$ is neither defined nor
continuous at $t=5$.
\item[(ii)] $u(t)=|t|$ is not differentiable (at $t=0$).
\item[(iii)] $u(t)=\csc(t)$ is neither defined nor continuous at
$t=k\pi$ for any integer $k$.
\end{enumerate}

The subset $\CCone\subset{\cal F}$  is a subspace and hence a vector space.
\index{vector!in $\CCone$}
The reason is simple.  If $x(t)$ and $y(t)$ are continuously differentiable,
then
\[
\frac{d}{dt}(x+y) = \frac{dx}{dt}+\frac{dy}{dt}.
\]
Hence $x+y$ is differentiable and is in $\CCone$ and $\CCone$ is closed under
addition.  Similarly, $\CCone$ is closed under scalar multiplication.  Let
$r\in\R$ and let $x\in\CCone$. Then
\[
\frac{d}{dt}(rx)(t) = r\frac{dx}{dt}(t).
\]
Hence $rx$ is differentiable and is in $\CCone$.

\subsubsection*{The Vector Space $(\CCone)^n$}

Another example of a vector space that combines the features of
both $\R^n$ and $\CCone$ is $(\CCone)^n$.  Vectors
$u\in (\CCone)^n$ have the form
\[
u(t) = (u_1(t),\ldots,u_n(t)),
\]
where each coordinate function $u_j(t)\in\CCone$.  Addition and
scalar multiplication in $(\CCone)^n$ are defined coordinatewise
--- just like addition and scalar multiplication in $\R^n$.
That is, let $u,v$ be in $(\CCone)^n$ and let $r$ be in $\R$, then
\begin{eqnarray*}
(u+v)(t) & = & (u_1(t)+v_1(t),\ldots,u_n(t)+v_n(t)) \\
(ru)(t) & = & (ru_1(t),\ldots,ru_n(t)).
\end{eqnarray*}
The set $(\CCone)^n$ satisfies the eight properties of vector spaces and
is a vector space.  Solutions to systems of $n$ linear ordinary differential
equations are vectors in $(\CCone)^n$.


\EXER

\TEXER

\begin{exercise} \label{c5.1.1}
Verify that the set $V_1$ consisting of all scalar multiples of
$(1,-1,-2)$ is a subspace of $\R^3$.

\begin{solution}

The set $V_1 \subset \R^3$ is a subspace.
The set $V_1$ contains all vectors $(a,-a,-2a)$,
where $a \in \R$.  We can show that it is closed under
vector addition, since
\[
a(1,-1,-2) + b(1,-1,-2) = (a + b)(1,-1,-2) \in V_1
\]
where $a$ and $b$ are scalars.  The set $V_1$ is closed under scalar
multiplication since
\[
b(a(1,-1,-2) = (ba)(1,-1,-2) \in V_1.
\]

\end{solution}
\end{exercise}

\begin{exercise} \label{c5.1.2}
Let $V_2$ be the set of all $2\times 3$ matrices.   Verify that
$V_2$ is a vector space.

\begin{solution}

The set $V_2$ is a vector space.
Let $A$ and $B$ be matrices in $V_2$:
\[
A = \left(\begin{array}{rrr} a_{11} & a_{12} & a_{13} \\
a_{21} & a_{22} & a_{23} \end{array}\right) \AND
B = \left(\begin{array}{rrr} b_{11} & b_{12} & b_{13} \\
b_{21} & b_{22} & b_{23} \end{array}\right).
\]
The set $V_2$ is closed under vector addition because
\[
A + B
= \left(\begin{array}{rrr} a_{11} + b_{11} & a_{12} + b_{12} &
a_{13} + b_{13} \\ a_{21} + b_{21} & a_{22} + b_{22} &
a_{23} + b_{23} \end{array}\right) \in V_2,
\]
and is closed under scalar multiplication because, for $r \in \R$,
\[
rA = \left(\begin{array}{rrr} ra_{11} & ra_{12} & ra_{13} \\
ra_{21} & ra_{22} & ra_{23} \end{array}\right) \in V_2.
\]
Next, verify that $V_2$ satisfies the eight properties of vector
spaces, shown in Table~\ref{vectorspacelist}.

\begin{itemize}
\item[(A1)] Addition is commutative in $V_2$, since
\[
A + B  = \left(\begin{array}{rrr} a_{11} + b_{11} & a_{12} + b_{12} &
a_{13} + b_{13} \\ a_{21} + b_{21} & a_{22} + b_{22} &
a_{23} + b_{23} \end{array}\right) = B + A.
\]
\item[(A2)] Given a third matrix $C \in V_2$,
\[
(A + B) + C = \left(\begin{array}{rrr} a_{11} + b_{11} + c_{11}
& a_{12} + b_{12} + c_{12} & a_{13} + b_{13} + c_{13} \\ a_{21} + b_{21}
+ c_{21} & a_{22} + b_{22} + c_{22} & a_{23} + b_{23} + c_{23}
\end{array}\right) = A + (B + C).
\]
Thus, addition is associative.
\item[(A3)] The additive identity is the matrix
\[
0 = \left(\begin{array}{rrr} 0 & 0 & 0 \\ 0 & 0 & 0 \end{array}\right).
\]
\item[(A4)] For each matrix $A \in V_2$, there exists an element $-A$ such
that $A + (-A) = 0$, where
\[
-A = \left(\begin{array}{rrr} -a_{11} & -a_{12} & -a_{13} \\ -a_{21} &
-a_{22} & -a_{23} \end{array}\right).
\]
\item[(M1)] Scalar multiplication is associative in $V_2$ since, for any
$r$, $s \in \R$ and $A \in V_2$,
\[
(rs)A = \left(\begin{array}{rrr} rsa_{11} & rsa_{12} & rsa_{13} \\
rsa_{21} & rsa_{22} & rsa_{23} \end{array}\right) = r(sA).
\]
\item[(M2)] There exists a multiplicative identity in $V_2$, since $1A = A$
for any $A \in V_2$.
\item[(D1)] For any scalars $r$, $s \in \R$, and any matrix $A \in V_2$,
\[
(r + s)A = \left(\begin{array}{rrr}
ra_{11} + sa_{11} & ra_{12} + sa_{12} & ra_{13} + sa_{13} \\
ra_{21} + sa_{21} & ra_{22} + sa_{22} & ra_{23} + sa_{23} \end{array}\right)
= rA + sA.
\]
\item[(D2)] For any scalar $r \in \R$ and any matrices $A, B \in V_2$,
\[
r(A + B) = \left(\begin{array}{rrr} ra_{11} + rb_{11} & ra_{12} + rb_{12} &
ra_{13} + rb_{13} \\ ra_{21} + rb_{21} & ra_{22} + rb_{22} &
ra_{23} + rb_{23} \end{array}\right)
= rA + rB.
\]
\end{itemize}
Thus, $V_2$ is a vector space, since it is closed under addition
and scalar multiplication and satisfies the eight properties of vector spaces.

\end{solution}
\end{exercise}

\begin{exercise} \label{c5.1.3}
Let
\[
A=\left(\begin{array}{rrr} 1 & 1 & 0\\ 1 & -1 & 1 \end{array}
\right).
\]
Let $V_3$ be the set of vectors $x\in\R^3$ such that $Ax=0$.
Verify that $V_3$ is a subspace of $\R^3$.  Compare $V_1$ with
$V_3$.

\begin{solution}

The set $V_3$ is a subspace of $R^3$ since the solution set to
any equation $Ax = 0$ is a space.  This is demonstrated by the
principle of superposition introduced in Section~\ref{S:Superposition}.
Also, $V_3 = V_1$.  

\para We can show that $V_3 = V_1$ by row reducing to find the
solutions to $Ax = 0$:
\[
\left(\begin{array}{rrr} 1 & 1 & 0 \\ 1 & -1 & 1
\end{array}\right) \longrightarrow \left(\begin{array}{rrr} 1 & 0 &
\frac{1}{2} \\ 0 & 1 & -\frac{1}{2} \end{array}\right).
\]
So all vectors in $V_3$ are of the form $x = s(-\frac{1}{2},
\frac{1}{2}, 1)$, where $s \in \R$.  The vector $x$ is an element
of $V_1$ for each $s$.


\end{solution}
\end{exercise}

\noindent In Exercises~\ref{c5.1.4a} -- \ref{c5.1.4f} you are given a
vector space $V$ and a subset $W$.  For each pair, decide whether or
not $W$ is a subspace of $V$.
\begin{exercise} \label{c5.1.4a}
$V=\R^3$ and $W$ consists of vectors in $\R^3$
     that have a $0$ in their first component.

\begin{solution}
$W$ is a subspace of $V$, since $W$ is closed under
addition and scalar multiplication.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.4b}
$V=\R^3$ and $W$ consists of vectors in $\R^3$
     that have a $1$ in their first component.

\begin{solution}
\ans $W$ is not a subspace of $V$.

\soln The subset $W$ is closed neither under addition nor under scalar
multiplication.  For example, let $w_1 = (1,4,2)$ and $w_2 = (1,-1,3)$
be elements of $W$.  Then,
\[
w_1 + w_2 = (1,4,2) + (1,-1,3) = (2,3,5)
\]
which is not an element of $W$.


\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.4d}
$V=\R^2$ and $W$ consists of vectors in $\R^2$
     for which the sum of the components is $1$.

\begin{solution}
\ans $W$ is not a subspace of $V$.

\soln The subset $W$ is closed neither under addition nor under scalar
multiplication.  For example, let $w_1 = (3,-2)$ and $w_2 = (0,1)$ be
elements of $W$.  Then,
\[
w_1 + w_2 = (3,-2) + (0,1) = (3,-3).
\]
The sum of the elements $3 - 3 = 0 \neq 1$.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.4c}
$V=\R^2$ and $W$ consists of vectors in $\R^2$
     for which the sum of the components is $0$.

\begin{solution}
$W$ is a subspace of $V$, since $W$ is closed under
addition and scalar multiplication.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.4g}
$V=\CCone$ and $W$ consists of functions
     $x(t)\in\CCone$ satisfying $\int_{-2}^4x(t)dt =0$.

\begin{solution}
$W$ is a subspace of $V$, since $W$ is closed under
addition and scalar multiplication.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.4e}
$V=\CCone$ and $W$ consists of functions
     $x(t)\in\CCone$ satisfying $x(1)=0$.

\begin{solution}
$W$ is a subspace of $V$, since $W$ is closed under
addition and scalar multiplication.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.4f}
$V=\CCone$ and $W$ consists of functions
     $x(t)\in\CCone$ satisfying $x(1)=1$.

\begin{solution}
\ans $W$ is not a subspace of $V$.

\soln The subset $W$ is closed neither under addition nor under scalar
multiplication.  For example, let $w_1(t) = t$ and let $w_2(t) = t^2$.
\[
(w_1 + w_2)(1) = w_1(1) + w_2(1) = 1 + 1 = 2,
\]
so $(w_1 + w_2)(t)$ is not an element of $W$.


\end{solution}
\end{exercise}

\noindent In Exercises~\ref{c5.1.5a} -- \ref{c5.1.5e} which of the
sets $S$ are subspaces?
\begin{exercise} \label{c5.1.5a}
$S = \{(a,b,c)\in\R^3 : a\ge 0, \; b\ge 0,\; c\ge 0\}$.

\begin{solution}
\ans The set $S$ is not a subspace.

\soln The set $S$ is closed under addition but not under scalar
multiplication.  To demonstrate, let $x = (1,4,2)$ be an element of $S$. 
Then
\[
-2x = -2(1,4,2) = (-2,-8,-4)
\]
which is not an element of $S$.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.5b}
$S = \{(x_1,x_2,x_3)\in\R^3 : a_1x_1+a_2x_2+a_3x_3=0
\mbox{ where } a_1,a_2,a_3\in\R \mbox{ are fixed}\}$.

\begin{solution}
The set $S$ is a subspace, since it is closed under
addition and scalar multiplication.

 
\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.5c}
$S = \{(x,y)\in\R^2: (x,y) \mbox{ is on the line through }
(1,1) \mbox{ with slope } 1\}$.

\begin{solution}
The set $S$ is a subspace, since it is closed under
addition and scalar multiplication.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.5d}
$S = \{x\in\R^2: Ax=0\}$ where $A$ is a $3\times 2$ matrix.

\begin{solution}
The set $S$ is a subspace, since it is closed under
addition and scalar multiplication.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.5e}
$S = \{x\in\R^2: Ax=b\}$ where $A$ is a $3\times 2$ matrix
	and $b\in\R^3$ is a fixed nonzero vector.

\begin{solution}
\ans The set $S$ is not a subspace.

\soln The set $S$ is closed under neither addition nor scalar
multiplication.  For example, let $x_1$ and $x_2$ be solutions to the
equation $Ax = b$.  Then,
\[
A(x_1 + x_2) = Ax_1 + Ax_2 = b + b = 2b,
\]
so $(x_1 + x_2)$ is not an element of $S$.

\end{solution}
\end{exercise}

\begin{exercise} \label{c5.1.6}
Let $V$ be a vector space and let $W_1$ and $W_2$ be subspaces.
Show that the intersection $W_1\cap W_2$ is also a subspace of $V$.

\begin{solution}

The subset $W_1 \cap W_2$ is a subspace of $V$.  To show that this
subset is closed under addition and scalar multiplication, 
let $x$ and $y$ be vectors in $W_1 \cap W_2$.  It follows that
$x,y \in W_1$ and $x,y \in W_2$.  Therefore, by the
definition of a subspace, $x + y \in W_1$ and $x + y \in W_2$, so
$x + y \in W_1 \cap W_2$.  Also by definition, $rx \in W_1$ and
$rx \in W_2$, for some scalar $r$, so $rx \in W_1 \cap W_2$.

\end{solution}
\end{exercise}

\begin{exercise} \label{c5.1.7a}
For which scalars $a,b,c$ do the solutions to the equation
\[
ax+by = c
\]
form a subspace of $\R^2$?

\begin{solution}

\ans Let $V$ be the subset of solutions $(x,y)$ to $ax + by = c$.
The subset $V$ is a subspace when $c = 0$ and is not a subspace
when $c \neq 0$. 

\soln Let $(x_1,y_1)$ and $(x_2,y_2)$ be elements of $V$.  Then
\[
a(x_1 + x_2) + b(y_1 + y_2) = (ax_1 + by_1) + (ax_2 + by_2) =
c + c = 2c.
\]
Thus $V$ is closed under addition only when $2c = c$, so $c = 0$.
Similarly, for any scalar $r$,
\[
r(ax_1 + by_1) = cr.
\]
So $V$ is closed under scalar multiplication only when $rc = c$ for
any scalar $r$.  Thus, $c = 0$.

\end{solution}
\end{exercise}
\begin{exercise} \label{c5.1.7b}
For which scalars $a,b,c,d$ do the solutions to the equation
\[
ax+by+cz = d
\]
form a subspace of $\R^3$?

\begin{solution}

\ans By the same proof as in Exercise~\ref{c5.1.7a}, the solutions to the
equation $ax + by + cz = d$ form a subspace of $\R^3$ when $d = 0$,
and do not form a subspace when $d \neq 0$.

\end{solution}
\end{exercise}

\begin{exercise} \label{c5.1.8}
Show that the set of all solutions to the differential equation
$\dot{x}=2x$ is a subspace of $\CCone$.

\begin{solution}

The set of all solutions to the differential equation $\dot{x} = 2x$ is
indeed a subspace of $\CCone$.  To demonstrate, let $x_1$ and $x_2$
be elements of this set.  The set is closed under addition since
\[
\frac{d}{dt}(x_1 + x_2) = \frac{d}{dt}(x_1) + \frac{d}{dt}(x_2)
= 2x_1 + 2x_2 = 2(x_1 + x_2)
\]
and closed under scalar multiplication since, for any real scalar $r$,
\[
\frac{d}{dt}(rx_1) = r\frac{d}{dt}(x_1) = 2(rx_1).
\]
Note that this problem provides another example of the principle of
superposition.

\end{solution}
\end{exercise}

\begin{exercise} \label{c5.1.9}
Recall from equation~\eqref{e:solnODE} of Section~\ref{S:IVP&E}
that solutions to the system of differential equations
\[
\frac{dX}{dt} = \mattwo{-1}{3}{3}{-1} X
\]
are
\[
X(t) = \alpha e^{2t}\vectwo{1}{1} + \beta e^{-4t}\vectwo{1}{-1}.
\]
Use this formula for solutions to show that the set of solutions
to this system of differential equations is a vector subspace of
$(\CCone)^2$.

\begin{solution}

Let $V \subset (\CCone)^2$ be the set of solutions to \eqref{e:solnODE}.
The set is closed under both addition and scalar multiplication and
is a subspace.
To demonstrate, let
\[
x_1(t) = \alpha_1 e^{2t}\vectwo{1}{1} +
\beta_1 e^{-4t}\vectwo{1}{-1} \AND x_2(t) =
\alpha_2 e^{2t}\vectwo{1}{1} + \beta_2 e^{-4t}\vectwo{1}{-1}
\]
be elements of this set.  Adding $x_1$ and $x_2$ yields
\[
\left(\alpha_1 e^{2t}\vectwo{1}{1} +
\beta_1 e^{-4t}\vectwo{1}{-1}\right) +
\left(\alpha_2 e^{2t}\vectwo{1}{1}
+ \beta_2 e^{-4t}\vectwo{1}{-1}\right)
\]
\[
= (\alpha_1 + \alpha_2)e^{2t}\vectwo{1}{1} +
(\beta_1 + \beta_2) e^{-4t}\vectwo{1}{-1} \in V
\]
and multiplying $x_1$ by any real scalar $r$ yields
\[
rx_1 = r\left(\alpha_1 e^{2t}\vectwo{1}{1} +
\beta_1 e^{-4t}\vectwo{1}{-1}\right) = r\alpha_1 e^{2t}\vectwo{1}{1} +
r\beta_1 e^{-4t}\vectwo{1}{-1} \in V.
\]



\end{solution}
\end{exercise}

\end{document}
