\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}

\noindent In Exercises~\ref{c8.2.4a} -- \ref{c8.2.4e} we review some
of the theory of planar linear systems of differential equations that
is needed in the study of planar nonlinear systems.  For the given 
matrix $C$, determine whether the origin is a hyperbolic equilibrium 
for the differential equation $\dot{X}=CX$.  When the origin is 
hyperbolic, what kind of system is it?
\begin{exercise} \label{c8.2.4a}
$C=\mattwo{0.1}{2}{-1}{-3}$. 

\begin{solution}
\ans The origin is a nodal sink.

\soln A matrix $C$ with nonzero trace and determinant is a hyperbolic
equilibrium.  In this case, $\trace(C) = -2.9$ and $\det(C) = 1.7$.
Since the determinant is positive and the trace is negative, the origin
is a sink. Since the discriminant is $1.61>0$ the origin is a node.

\end{solution}
\end{exercise}
\begin{exercise} \label{c8.2.4b}
$C=\mattwo{4}{-2}{1}{5}$. 

\begin{solution}
\ans The origin is a spiral source.

\soln In this case, $\trace(C) = 9$ and $\det(C) = 22$, so the origin
is a hyperbolic equilibrium and a source.  Since the discriminant is 
$-1<0$, the origin is a spiral.

\end{solution}
\end{exercise}
\begin{exercise} \label{c8.2.4c}
$C=\mattwo{2}{-1}{4}{-2}$. 

\begin{solution}
\ans The origin is not hyperbolic.

\soln Since $\trace(C) = 0$ and $\det(C) = 0$, the origin is not 
hyperbolic. 

\end{solution}
\end{exercise}
\begin{exercise} \label{c8.2.4d}
$C=\mattwo{1}{3}{-2}{-6}$.

\begin{solution}
\ans The origin is not hyperbolic.

\soln Since $\det(C) = 0$ the origin is not hyperbolic.

\end{solution}
\end{exercise}
\begin{exercise} \label{c8.2.4e}
$C=\mattwo{11}{-8}{18}{-13}$.

\begin{solution}
\ans The origin is an improper nodal sink.

\soln Since $\trace(C) = -2$ and $\det(C) = 1$, the equilibrium
is hyperbolic. Since the trace is negative, the origin is a sink.
Since the discriminant is $0$, the origin is an improper node. 

\end{solution}
\end{exercise}
\end{document}
