\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}
\begin{exercise} \label{c8.2.6}
Consider the planar nonlinear system of ODEs
\begin{eqnarray*}
\frac{dx}{dt} & = & 1 - 2x + y + x^2 - xy \\
\frac{dy}{dt} & = & y - y^2
\end{eqnarray*}
\begin{itemize}
\item[(a)] Find all equilibria of this system.
\item[(b)] Are the equilibria asymptotically stable or not?
\item[(c)] Determine the type of each equilibrium.
\end{itemize}

\begin{solution}

(a) \ans Equilibria occur at the points $(1,0)$, $(1,1)$, and $(2,1)$.

\soln Solve $0 = y - y^2$ to obtain $y = 1$ and $y = 0$, then substitute
these values for $y$ into $0 = 1 - 2x + y + x^2 - xy$ to find the points
at which $\frac{dx}{dt} = \frac{dy}{dt} = 0$.

(b) \ans The equilibria at $(1,0)$ and $(2,1)$ are not asymptotically
stable.  The equilibrium at $(1,1)$ is stable.

\soln Compute the Jacobian matrix of the system:
\[
(dJ)_{(x,y)} = \cmattwo{-2 + 2x - y}{1 - x}{0}{1 - 2y}
\]
Specifically, at the equilibrium points, the Jacobian matrices are
\[
(dJ)_{(1,0)} = \mattwo{0}{0}{0}{1} \qquad (dJ)_{(1,1)} =
\mattwo{-1}{0}{0}{-1} \qquad (dJ)_{(2,1)} = \mattwo{1}{-1}{0}{-1}.
\]
An equilibrium is asymptotically stable if both eigenvalues are
negative.  The eigenvalues of $(1,0)$ are $0$ and $1 > 0$.
The point $(1,1)$ has a double eigenvalue at $-1 < 0$.
The eigenvalues of $(1,1)$ are $-1 < 0$ and $1 > 0$.

(c) The equilibrium at $(1,0)$ is a saddle node, since the Jacobian
has one zero eigenvalue and one nonzero eigenvalue.  The equilibrium
at $(1,1)$ is a focus sink, since the Jacobian is a multiple of $I_2$
with negative real eigenvalue.  The equilibrium at $(2,1)$ is a
saddle point, since the Jacobian has one negative and one positive
eigenvalue.

\end{solution}
\end{exercise}
\end{document}
