\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}
\begin{computerExercise} \label{c2.3.4}
{\bf Comment:} {\rm To understand the point of this exercise you
must begin by typing the \Matlab command {\tt format short e}.
This command will set a format in which you can see the
difficulties that sometimes arise in numerical computations.}

Consider the following two $3\times 3$-matrices:
\begin{matlabEquation}\label{MATLAB:14}
A=\left( \begin{array}{rrr}
     1  &  3  &  4\\
     2  &  1  &  1\\
    -4  &  3  &  5
\end{array}\right) \AND
B=\left( \begin{array}{rrr}
     3  &  1  &  4\\
     1  &  2  &  1\\
     3  & -4  &  5
\end{array}\right).
\end{matlabEquation}
Note that matrix $B$ is obtained from matrix $A$ by interchanging the
first two columns.
\begin{itemize}
\item[(a)] Use \Matlab to put $A$ into row echelon form using the
transformations
\begin{enumerate}
\item Subtract $2$ times the $1^{st}$ row from the $2^{nd}$.
\item Add $4$ times the $1^{st}$ row to the $3^{rd}$.
\item Divide the $2^{nd}$ row by $-5$.
\item Subtract $15$ times the $2^{nd}$ row from the $3^{rd}$.
\end{enumerate}
\item[(b)] Put $B$ by hand into row echelon form using the
transformations
\begin{enumerate}
\item Divide the $1^{st}$ row by $3$.
\item Subtract the $1^{st}$ row from the $2^{nd}$.
\item Subtract $3$ times the $1^{st}$ row from the $3^{rd}$.
\item Multiply the $2^{nd}$ row by $3/5$.
\item Add $5$ times the $2^{nd}$ row to the $3^{rd}$.
\end{enumerate}
\item[(c)] Use \Matlab to put $B$ into row echelon form using the
same transformations as in part (b).
\item[(d)] Discuss the outcome of the three transformations.  Is
there a difference in the results?  Would you expect to see a
difference?  Could the difference be crucial when solving a system
of linear equations?
\end{itemize}

\begin{solution}

(a)
\begin{verbatim}
A =
   1.0000e+00   3.0000e+00   4.0000e+00
            0   1.0000e+00   1.4000e+00
            0            0            0
\end{verbatim}

(b)
$B = \left(\begin{array}{rrr}
1 & \frac{1}{3} & \frac{4}{3} \\
0 & 1 & -\frac{1}{5} \\
0 & 0 & 0\end{array} \right)$

(c)
\begin{verbatim}
B =
   1.0000e+00   3.3333e-01   1.3333e+00
            0   1.0000e+00  -2.0000e-01
            0            0   2.2204e-16
\end{verbatim}


(d) Note that switching the first two columns of matrix {\tt A} produces
matrix {\tt B}.  Suppose that {\tt A} and {\tt B} represent the left-hand
sides of linear systems with the same vector representing the right-hand
sides.  If the solution of system {\tt A} is $(x,y,z) = (a,b,c)$, then the
solution of system {\tt B} should be $(x,y,z) = (b,a,c)$.  However,
according to the row reduced matrices produced by \Matlab, the system
corresponding to {\tt B} has a unique solution, and the system
corresponding to {\tt A} does not.  Row reducing {\tt B} by hand shows
that there is not a unique solution.  \Matlab calculations provide one because of 
roundoff error.  The division by 3 in the first step of the row reduction
for {\tt B} causes a rounding inaccuracy.  Because of this, \Matlab
eventually computes a very small nonzero value for {\tt B}$(3,3)$
rather than the correct answer of 0.

\end{solution}
\end{computerExercise}
\end{document}
