\documentclass{ximera}

\input{../preamble.tex}

\title{Simplification by Substitution}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{sec:SBS}

So far in this chapter --- as well as in Chapter~\ref{C:LDE} ---
we have seen how to find solutions of ordinary differential equations 
that have special forms.  (For instance, separation of variables can 
be applied to solve equations of the form $\frac{dx}{dt}=g(x) h(t)$.) 
However, ``most'' differential equations are not of the form needed to
apply one of these techniques.  But sometimes it is possible to
transform the equation into such a form.  This is accomplished
by substituting a new function for $x(t)$, so that the differential 
equation for this new function has a simpler form.  We illustrate this 
procedure in two cases.

\subsection*{Homogeneous Coefficients}
\index{homogeneous coefficients}

Differential equations of the form
\begin{equation}
\label{eq:homcoeff}
\frac{dx}{dt} = F\left(\frac{x}{t}\right),
\end{equation}
where $F:\R\to\R$ is continuous, are said to have {\em
homogeneous coefficients}\index{homogeneous coefficients}.  
In general, none of the techniques
described so far can be applied to this equation.  

Since the function $x(t)/t$ appears as the argument of $F$ in
the equation it seems plausible to write down \eqref{eq:homcoeff}
in terms of this function.  Having this in mind, define
\[
v(t) = \frac{x(t)}{t}.
\]
Then $x(t) = t v(t)$ and \eqref{eq:homcoeff} becomes
\[
v(t) + t \frac{dv}{dt}(t) = F(v(t)).
\]
When $t\not= 0$, this equation is equivalent to
\begin{equation}
\label{eq:homsep}
\frac{dv}{dt} = \frac{F(v)-v}{t}.
\end{equation}
We have arrived at an equation to which we can apply separation
of variables\index{separation of variables}.  
Once a solution $v(t)$ of \eqref{eq:homsep} is
found, we obtain a solution of \eqref{eq:homcoeff} by setting
\[
x(t) = t v(t).
\]
If an initial condition $x(t_0)=x_0$ is specified in
\eqref{eq:homcoeff}, then we have to solve \eqref{eq:homsep} with
the transformed initial condition $v(t_0)=x_0/t_0$.

\subsubsection*{An Example}
Consider the initial value problem
\begin{equation}
\label{eq:homcoex1}
\begin{array}{rcl}
\dps \frac{dx}{dt} & = & \dps \frac{2t^2+x^2}{tx} \\
x(2) & = & 6.
\end{array}
\end{equation}
{\rm Since the right hand side of this equation can be written as
\[
F\left(\frac{x}{t}\right) = 2\frac{t}{x}+\frac{x}{t},
\]
we see that \eqref{eq:homcoex1} has homogeneous 
coefficients\index{homogeneous coefficients}.  We
set $v(t) = x(t)/t$. By \eqref{eq:homsep} we have to solve the
initial value problem\index{initial value problem}
\[
\begin{array}{rcl}
\dps \frac{dv}{dt} & = & \dps \frac{F(v)-v}{t} = 
\frac{2\frac{1}{v}+v-v}{t}=\frac{2}{tv} \\
v(2) & = & 6/2 = 3.
\end{array}
\]
Separation of variables \index{separation of variables} shows that 
$v(t)$ has to satisfy
\[
\frac{v^2}{2} = \ln(t) - \ln(2) + \frac{9}{2}
\]
(see Section~\ref{sec:sov}). Therefore
\[
v(t) = \sqrt{9+2\ln\frac{t}{2}}.
\]
Finally, we obtain the solution
\[
x(t) = tv(t) = t\sqrt{9+2\ln\frac{t}{2}}.
\]}

\subsection*{Bernoulli's Equation}
\index{Bernoulli's equation}

Let $r,s:\R\rightarrow\R$ be continuous functions, and let $p\in\R$
be a real number.  Then an equation of the form
\begin{equation}
\label{eq:bern1}
\frac{dx}{dt} = r(t)x + s(t)x^p
\end{equation}
is called a {\em Bernoulli equation}.  For $p=0$ or $p=1$ the
equation is linear and we can, in principle, find all the
solutions by variation of parameters (see Chapter~\ref{C:LDE}).  
Hence we assume that
\[
p\not= 0,1.
\]
The idea is to substitute $x(t)$ in such a way that also for these
values of $p$ \eqref{eq:bern1} becomes linear in the new function $v(t)$.  
We try the guess
\[
v(t) = x(t)^{1/\alpha},
\]
where we choose the real constant $\alpha\not= 0$ later to simplify
the transformed equation.  Using the chain rule on $x(t) =
v(t)^\alpha$, we compute
\[
\frac{dx}{dt} = \alpha v^{\alpha-1} \frac{dv}{dt}.
\]
Substitution into \eqref{eq:bern1} yields
\[
\alpha v^{\alpha-1} \frac{dv}{dt} = r(t)v^\alpha + s(t)v^{p\alpha}.
\]
Thus
\[
\frac{dv}{dt}  = \frac{1}{\alpha} r(t) v + 
\frac{1}{\alpha} s(t) v^{p\alpha-\alpha+1}.
\]
To simplify this equation, choose the constant $\alpha$ so that
$p\alpha-\alpha+1=0$; that is, set
\[
\alpha = \frac{1}{1-p}.
\]
Then we arrive at the {\em linear\/} equation
\begin{equation}
\label{eq:bern2}
\frac{dv}{dt} = (1-p) r(t) v + (1-p) s(t),
\end{equation}
which we can, in principle, be solved by 
variation of parameters\index{variation of parameters}.
If this can be done, then we obtain a solution $x(t)$ of
\eqref{eq:bern1} by setting 
\[
x(t) = (v(t))^{\frac{1}{1-p}}.
\]
If an initial condition $x(t_0)=x_0$ is specified in
\eqref{eq:bern1},
then we have to solve \eqref{eq:bern2} with the transformed initial
condition
$v(t_0)=x_0^{1/\alpha}=x_0^{1-p}$.

\subsubsection*{An Example}
Consider the initial value problem
\[
\begin{array}{rcl}
\dps \frac{dx}{dt} & = &  3t^2 x + 3t^2x^{\frac{2}{3}}\\
x(0) & = & 27.
\end{array}
\]
{\rm The differential equation is a Bernoulli equation with
\[
r(t) = 3t^2,\quad s(t)=3t^2, \AND p=\frac{2}{3}.
\]
Hence we have to find a solution of the linear initial value
problem (see \eqref{eq:bern2})
\[
\begin{array}{rcl}
\dps \frac{dv}{dt} & = &  t^2 v + t^2 \\
v(0) & = & 27^{1/3} = 3.
\end{array}
\]
The solution to this differential equation can be obtained by variation of 
parameters and is 
\[
v(t) = -1 +4 e^{t^3/3}.
\]
Therefore, a solution of the Bernoulli equation is given by
\[
x(t) = v(t)^{\frac{1}{1-p}} = \left(-1 +4 e^{t^3/3}\right)^3.
\]}

\EXER

\TEXER

\noindent In Exercises~\ref{c14.5.1} -- \ref{c14.5.5} decide whether the 
given differential equation has homogeneous coefficients or is a Bernoulli 
equation.
\begin{exercise} \label{c14.5.1}
$\dps \frac{dx}{dt} = \frac{x^2}{t^3}$.

\begin{solution}
\ans The system is a Bernoulli equation.

\soln A Bernoulli equation has the form $\dot{x}=rx+sx^p$.  In this equation
take $r(t)=0$, $s(t)= \frac{1}{t^3}$, and $p=2$.

\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.2}
$\dps \frac{dx}{dt} = t$.

\begin{solution}
\ans The system is a Bernoulli equation.

\soln A Bernoulli equation has the form $\dot{x}=rx+sx^p$.  In this equation
take $r(t)=0$, $s(t)=t$, and $p=0$.

\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.4}
$\dps \frac{dx}{dt} = \frac{\cos t}{x^4}\sqrt{t}+x^2$.

\begin{solution}
\ans The system is a Bernoulli equation.

\soln A Bernoulli equation has the form $\dot{x}=rx+sx^p$.  In this equation
take $r(t)=\sin t$, $s(t)=t^2$, and $p=-\frac{1}{2}$.

\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.3}
$\dps \frac{dx}{dt} = \frac{t^2}{\sqrt{x}}+x\sin t$.

\begin{solution}
\ans The system is neither.

\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.5}
$\dps \frac{dx}{dt} = \frac{x}{t} +\frac{\sqrt{t}}{\sqrt{x}}$.

\begin{solution}
\ans The system has homogeneous coefficients and is a
Bernoulli equation.

\soln A Bernoulli equation has the form $\dot{x}=rx+sx^p$.  In this equation
take $r(t)=\frac{1}{t}$, $s(t)=\sqrt{t}$, and $p=-\frac{1}{2}$.  An equation
with homogeneous equations has the form $\dot{x}=F(x/t)$.  For this equation
$F(z) = z + \frac{1}{\sqrt{z}}$.

\end{solution}
\end{exercise}

\noindent In Exercises~\ref{c14.5.6} -- \ref{c14.5.9} solve the given initial 
value problem by an appropriate solution technique.
\begin{exercise} \label{c14.5.6}
$\dps \frac{dx}{dt} = \sec\left(\frac{x}{t}\right)+\frac{x}{t}$ where
$x(1)=\pi$.

\begin{solution}
\ans $x(t) = t\sin\inv(\ln|t|)$.

\soln  This differential equation is an equation with homogeneous
coefficients where $F(v) = v + \sec v$.  Using the substitution $v=x/t$, we
are led to the differential equation $t\dot{v}=\sec v$, which can be solved
by separation of variables.  That is, solve
\[
\cos v \frac{dv}{dt} = \frac{1}{t}
\]
by integration, obtaining
\[
\sin v = \ln|t| + c.
\]
We can solve for $c$ using the initial condition $x(1)=\pi$, which implies
that $v(1)=x(1)/1=\pi$.  Therefore, $c=0$.  Hence
\[
v = \sin\inv(\ln|t|).
\]
Using the identity $v=x/t$ leads to the answer. 
 

\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.7}
$\dps \frac{dx}{dt} = x+x^2$ where $x(2)=1$.

\begin{solution}
\ans $x(t) = \dps\frac{t}{4-t}$.

\soln  Although this equation is a Bernoulli equation, it is most directly
solved by separation of variables.  That is, solve
\[
\frac{1}{x(x+1)}\frac{dx}{dt} = 1.
\]
Using partial fractions, we obtain
\[
\left(\frac{1}{x}-\frac{1}{x+1}\right)\frac{dx}{dt} = 1
\]
On integration, we find
\[
\ln\left|\frac{x}{x+1}\right| = \ln|t| + c.
\]
Therefore,
\[
\frac{x}{x+1} = ct,
\]
for some constant $c$.  Using the initial condition $x(2)=1$ leads to
$c=\frac{1}{4}$.  After cross-multiplying we are led to the answer.

\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.8}
$\dps \frac{dx}{dt} = -\frac{x}{t}-t^3 x^3$ where $x(1)=1$.

\begin{solution}
\ans $x(t) = \dps\frac{1}{t^2}$.

\soln This differential equation is a Bernoulli equation with
\[
r(t) = -\frac{1}{t},\quad s(t) = -t^3 \AND p=3.
\]
After transformation the corresponding linear initial value problem
is given by
\[
\frac{dv}{dt} = \frac{2}{t} v + 2t^3,\quad v(1)=1.
\]
We solve this equation by variation of parameters.  The solution
of the homogeneous equation $\dot v = \frac{2}{t} v$ is
$v_h(t)=t^2$, and we use Theorem~\ref{thm:varpar} to find the solution
\[
v(t) = c(t) v_h(t) = t^2 v_h(t) = t^4.
\]
Back substitution now leads to the solution
\[
x(t) = (v(t))^{-\frac{1}{2}} = \frac{1}{t^2}.
\]


\end{solution}
\end{exercise}
\begin{exercise} \label{c14.5.9}
$\dps \frac{dx}{dt} = \frac{x(x+t)}{t^2}$ where $x(1)=1$.

\begin{solution}
\ans $x(t) = \dps\frac{t}{1-\ln|t|}$.

\soln  This differential equation is an equation with homogeneous
coefficients where $F(v) = v + v^2$.  Using the substitution $v=x/t$, we
are led to the differential equation $t\dot{v}=v^2$, which can be solved
by separation of variables.  That is, solve
\[
 \frac{1}{v^2} \frac{dv}{dt} = \frac{1}{t}
\]
by integration, obtaining
\[
-\frac{1}{v} = \ln|t| + c.
\]
We can solve for $c$ using the initial condition $x(1)=1$, which implies
that $v(1)=x(1)/1=1$.  Therefore, $c=-1$.  Hence
\[
v = \frac{1}{1-\ln|t|}.
\]
Using the identity $v=x/t$ leads to the answer.   Note that this equation is
also a Bernoulli equation and could be solved using the techniques for
Bernoulli equations, as well.

\end{solution}
\end{exercise}

\begin{exercise} \label{c14.5.10}
We set $v(t) = x(t)^{1-p}$ to transform the Bernoulli equation 
\eqref{eq:bern1} into the equation
\[
\frac{dv}{dt} = (1-p) r(t) v + (1-p) s(t).
\]
Now set $w(t) = v(\beta t)$ and show that $\beta$ can be chosen so that 
$w(t)$ is a solution to the linear differential equation
\[
\frac{dw}{dt} = r\left(\frac{t}{1-p}\right) w + s\left(\frac{t}{1-p}\right).
\]

\begin{solution}
\ans $\beta = 1/(1-p)$.

\soln Let $w(t) = v(\beta t)$ and compute
\[
\frac{dw}{dt}(t) = \beta \frac{dv}{dt}(\beta t).
\]
Using the linear differential equation for $v$, we obtain
\[
\frac{dw}{dt}(t) = \beta ((1-p) r(\beta t) v(\beta t) + (1-p) s(\beta t))
=  \beta (1-p) r(\beta t) w(t) + \beta (1-p) s(\beta t).
\]
Now choose $\beta = 1/(1-p)$.




\end{solution}
\end{exercise}


\end{document}
