\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}
\begin{exercise} \label{c8.3.5}
An autonomous planar system of differential equations is a {\em gradient\/} 
\index{gradient} system if there exists a real-valued differentiable function 
$f(x,y)$ such that 
\begin{equation} \label{E:gradsys}
\begin{array}{rcl}
\dot{x} & = & f_x(x,y) \\
\dot{y} & = & f_y(x,y).
\end{array}
\end{equation}
Prove that gradient systems never have periodic solutions.  

{\bf Hint:} Proceed in four steps as follows.
\begin{itemize}
\item[(a)]  Let $(x(t),y(t))$ be a solution to \eqref{E:gradsys} and let 
$h(t)=f(x(t),y(t))$.  So $h:\R\to\R$.  Use the chain rule to show that 
$h'(t)\geq 0$.
\item[(b)]  Show that $h'(t_0) = 0$ for some $t_0$ if and only if 
$(x(t),y(t)$ is an equilibrium of  \eqref{E:gradsys}.
\item[(c)]  Show that $h$ is monotonic increasing along any nonconstant 
solution of \eqref{E:gradsys}. 
\item[(d)]  Explain why a nonconstant solution of \eqref{E:gradsys} cannot be 
periodic.
\end{itemize}

\begin{solution}

\soln (a)  Use the chain rule to calculate
\[
\dot{h} = f_x\dot{x}+f_y\dot{y}.
\]
Using the gradient system of differential equations, observe that
\[
\dot{h} = f_x^2 + f_y^2\ge 0.
\]
(b) It follows from (a) that $\dot{h}(t_0)=0$ if and only if 
$f_x(x(t_0),y(t_0))=0$ and $f_y(x(t_0),y(t_0))=0$.  That is, 
$\dot{h}(t_0)=0$ if and only if $(x(t_0),y(t_0))$ is an equilibrium for the
gradient system.

\noindent (c) It follows from (b) that $\dot{h}$ is nowhere zero on a
nonconstant solution and from (a) that $\dot{h}(t)>0$ on a nonconstant
solution.  Therefore, $h(t)$ is monotonic increasing along any nonconstant
solution.

\noindent (d)  If $(x(t),y(t)$ is a nonconstant periodic solution of the
gradient system with period $T>0$, then $h(t+T)=h(t)$ for all $t$.  Thus,
$h(t)$ is not monotonic along this solution, contradicting (c).  Therefore,
nonconstant periodic solutions of gradient systems cannot exist.  

\end{solution}
\end{exercise}
\end{document}
