\documentclass{ximera}

\input{./preamble.tex}

\title{Matrix Mappings}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

  \index{matrix!mappings} \label{s:4.2}

Having illustrated the notational advantage of using matrices
and matrix multiplication, we now begin to discuss why there
is also a {\em conceptual advantage\/} to matrix
multiplication, a conceptual advantage that will help
us to understand how systems of linear equations and linear
differential equations may be solved.

Matrix multiplication allows us to view $m\times n$ matrices as
mappings from $\R^n$ to $\R^m$.  Let $A$ be an $m\times n$
matrix and let $x$ be an $n$ vector.  Then
\[
x \mapsto Ax
\]
defines a mapping from $\R^n$ to $\R^m$.

The simplest example of a matrix mapping is given by $1\times 1$
matrices.  Matrix mappings defined from $\R\to\R$ are
\[
x \mapsto ax
\]
where $a$ is a real number.  Note that the graph of this
function is just a straight line through the origin (with slope
$a$).  From this example we see that matrix mappings are very
special mappings indeed. In higher dimensions, matrix mappings
provide a richer set of mappings; we explore here {\em planar\/}
mappings\index{planar mappings} --- mappings of the plane into
itself --- using \Matlab graphics and the program {\sf map}.

The simplest planar matrix mappings are the
{\em dilatations\/}\index{dilatation}.
Let $A=cI_2$ where $c>0$ is a scalar.  When $c<1$ vectors are
contracted by a factor of $c$ and and these mappings are
examples of {\em contractions}\index{contraction}.
When $c>1$ vectors are
stretched or expanded by a factor of $c$ and these dilatations
are examples of {\em expansions}\index{expansion}.
We now explore some more complicated planar matrix mappings.

The next planar motions that we study are those given by the
matrices
\[
A=\mattwo{\lambda}{0}{0}{\mu}.
\]
Here the matrix mapping is given by $(x,y)\mapsto(\lambda x,\mu y)$;
that is, a mapping that independently stretches and/or contracts the
$x$ and $y$ coordinates.  Even these simple looking mappings can move 
objects in the plane in a somewhat complicated fashion.

\subsection*{The Program Map}  \index{\computer!map}

We can use \Matlab to explore planar matrix mappings in an efficient way 
using the program {\tt map}.  In \Matlab type the command
\begin{verbatim}
map
\end{verbatim}
and a menu appears labeled {\sf MAP Setup}.  The $2\times 2$ matrix
\[
\mattwo{0}{-1}{1}{0}
\]
has been pre-entered.  Click on the {\sf Proceed} button.  A
window entitled {\sf MAP Display} appears. Click on {\sf Icons}
and click on an icon --- say {\sf Dog}. \index{\computer!dog}
Then click in the {\sf MAP Display} window and a blue `{\sf Dog}' 
will appear in that window.  Now click on
the {\sf Map} button and a new version of the {\sf Dog} will
appear in yellow --- but the yellow {\sf Dog} is rotated about
the origin counterclockwise by $90^\circ$ from the blue dog.
Indeed, this matrix $A$ just rotates the plane counterclockwise
by $90^\circ$.  To verify this statement just click on {\sf Map}
again and see that the yellow dog rotates $90^\circ$
counterclockwise into the magenta dog.  Of course, the magenta
dog is just rotated $180^\circ$ from the original blue dog.
Clicking on {\sf Map} again produces a fourth dog --- this one
in cyan.  Finally one more click on the {\sf map} button will
rotate the cyan dog into a red dog that exactly covers the
original blue dog.

Choose another icon from the {\sf Icons} menu; a blue version of
this icon appears in the {\sf MAP Display} window.  Now click on
{\sf Map} to see that your chosen icon is just rotated
counterclockwise by $90^\circ$.

Other matrices will produce different motions of the plane.  You
may either type the entries of a matrix in the {\sf Map Setup}
window and click on the {\sf Proceed} button or recall one of
the pre-assigned matrices listed in the menu obtained by
clicking on {\sf Gallery}.  For example, clicking on
the {\sf Contracting rotation} button enters the matrix
\[
\mattwo{0.3}{-0.8}{0.8}{0.3}
\]
This matrix rotates the plane through an angle of approximately
$69.4^\circ$ counterclockwise and contracts the plane by a
factor of approximately $0.85$.  Now click on {\sf Dog} in the
{\sf Icons} menu to bring up the blue dog again.  Repeated
clicking on {\sf map} rotates and contracts the dog so that dogs
in a cycling set of colors slowly converge towards the origin in
a spiral of dogs.

\subsubsection*{Rotations}\index{rotation}

Rotating the plane counterclockwise through an angle $\theta$ is
a motion given by a matrix mapping.  We show that the matrix that
performs this rotation is:
\begin{equation} \label{e:rotmat}
R_\theta = \mattwo{\cos\theta}{-\sin\theta}{\sin\theta}{\cos\theta}.
\end{equation}
To verify that $R_\theta$ rotates the plane counterclockwise
through angle $\theta$, let $v_\varphi$ be the unit vector whose
angle from the horizontal is $\varphi$; that is,
$v_\varphi=(\cos\varphi,\sin\varphi)$.  We can write every vector in
$\R^2$ as $rv_\varphi$ for some number $r\ge 0$.   Using the 
trigonometric identities for the cosine and sine of the sum of two angles, we
have:
\begin{eqnarray*}
R_\theta (rv_\varphi) & = &
\mattwo{\cos\theta}{-\sin\theta}{\sin\theta}{\cos\theta}
\vectwo{r\cos\varphi}{r\sin\varphi} \\
& = & \vectwo{r\cos\theta\cos\varphi -r\sin\theta\sin\varphi}
{r\sin\theta\cos\varphi + r\cos\theta\sin\varphi}\\
& = & r\vectwo{\cos(\theta+\varphi)}{\sin(\theta+\varphi)}  \\
& = & rv_{\varphi+\theta}.
\end{eqnarray*}
This calculation shows that $R_\theta$ rotates every vector in the plane
counterclockwise through angle $\theta$.

It follows from \Ref{e:rotmat} that $R_{180^\circ} = -I_2$.  So rotating a
vector in the plane by $180^\circ$ is the same as reflecting the vector
through the origin.  It also follows that the movement associated with the
linear map $x\mapsto -cx$ where $c>0$ may be thought of as a dilatation
($x\mapsto cx$) followed by rotation through $180^\circ$ ($x\mapsto -x$).

We claim that combining dilatations with general rotations produces spirals.  
Consider the matrix
\[
S = \mattwo{c\cos\theta}{-c\sin\theta}{c\sin\theta}{c\cos\theta} = cR_\theta
\]
where $c<1$.  Then a calculation similar to the previous one shows that
\[
S(rv_\varphi) = c(rv_{\varphi+\theta}).
\]
So $S$ rotates vectors in the plane while contracting them by
the factor $c$.  Thus, multiplying a vector repeatedly by $S$ 
spirals that vector into the origin.  The example that we just 
considered while using {\sf map} is
\[
\mattwo{0.3}{-0.8}{0.8}{0.3}\cong\mattwo{0.85\cos(69.4^\circ)}
{-0.85\sin(69.4^\circ)}{0.85\sin(69.4^\circ)}{0.85\cos(69.4^\circ)},
\]
which has the general form of $S$.

\subsection*{A Notation for Matrix Mappings}
\index{matrix!mappings}

We reinforce the idea that matrices are mappings by introducing a notation 
for the mapping associated with an $m\times n$ matrix $A$.  Define
\[
L_A:\R^n\to\R^m
\]
by
\[
L_A(x) = Ax,
\]
for every $x\in\R^n$.

There are two special matrices:  the $m\times n$ zero matrix 
\index{matrix!zero} $O$ all of whose entries are $0$ and the 
$n\times n$ identity matrix \index{matrix!identity} $I_n$ whose diagonal 
entries are $1$ and whose off diagonal entries are $0$.  For instance,
\[
	I_3 = \left(
\begin{array}{rrr}
 1 & 0 & 0  \\
 0 & 1 & 0  \\
 0 & 0 & 1
\end{array}
\right).
\]

The mappings associated with these special matrices are also special.  
Let $x$ be an $n$ vector.  Then
\begin{equation} \label{multby0}
Ox=0,
\end{equation}
where the $0$ on the right hand side of \Ref{multby0} is the $m$
vector all of whose entries are $0$.  The mapping $L_O$ is the 
{\em zero mapping\/} \index{zero mapping} --- the mapping 
that maps every vector $x$ to $0$.

Similarly,
\[
I_nx=x
\]
for every vector $x$.  It follows that
\[
L_{I_n}(x) = x
\]
is the {\em identity mapping\/}, \index{identity mapping} since
it maps every element to itself.  It is for this reason that the
matrix $I_n$ is called the $n\times n$ {\em identity matrix\/}.

\EXER

\TEXER

\noindent In Exercises~\ref{c4.2.a1a} -- \ref{c4.2.a1c} find a nonzero
vector that is mapped to the origin by the given matrix.
\begin{exercise} \label{c4.2.a1a}
$A=\mattwo{0}{1}{0}{-2}$.
\end{exercise}
\begin{exercise} \label{c4.2.a1b}
$B=\mattwo{1}{2}{-2}{-4}$.
\end{exercise}
\begin{exercise} \label{c4.2.a1c}
$C=\mattwo{3}{-1}{-6}{2}$.
\end{exercise}


\begin{exercise} \label{c4.2.1a}
What $2\times 2$ matrix rotates the plane about the origin counterclockwise
by $30^\circ$?
\end{exercise}
\begin{exercise} \label{c4.2.1b}
What $2\times 2$ matrix rotates the plane clockwise by $45^\circ$?
\end{exercise}
\begin{exercise} \label{c4.2.1c}
What $2\times 2$ matrix rotates the plane clockwise by $90^\circ$ while
dilating it by a factor of $2$?
\end{exercise}

\begin{exercise} \label{c4.2.2a}
Find a $2\times 2$ matrix that reflects vectors in the $(x,y)$ plane across
the $x$ axis.
\end{exercise}
\begin{exercise} \label{c4.2.2b}
Find a $2\times 2$ matrix that reflects vectors in the $(x,y)$ plane across
the $y$ axis.
\end{exercise}
\begin{exercise} \label{c4.2.2c}
Find a $2\times 2$ matrix that reflects vectors in the $(x,y)$ plane across
the line $x=y$.
\end{exercise}

\begin{exercise} \label{c7.8.1}
The matrix
\[
A=\mattwo{1}{K}{0}{1}
\]
is a {\em shear}\index{shear}.  Describe the action of $A$ on the plane
for different values of $K$.
\end{exercise}

\begin{exercise} \label{c7.8.2}
Determine a rotation matrix that maps the vectors $(3,4)$ and
$(1,-2)$ onto the vectors $(-4,3)$ and $(2,1)$ respectively.
\end{exercise}



\begin{exercise} \label{c4.2.3}
Find a $2\times 3$ matrix $P$ that projects three dimensional $xyz$ space onto
the $xy$ plane.  {\bf Hint:} Such a matrix will satisfy
\[
P\left(\begin{array}{c} 0 \\ 0 \\ z \end{array}\right) = \vectwo{0}{0}
\AND
P\left(\begin{array}{c} x \\ y \\ 0 \end{array}\right) = \vectwo{x}{y}.
\]
\end{exercise}

\begin{exercise}	\label{c4.2.3a}
Show that every matrix of the form $\mattwo{a}{-b}{b}{a}$ corresponds to
rotating the plane through the angle $\theta$ followed by a dilatation 
$cI_2$ where
\begin{eqnarray*}
c & = & \sqrt{a^2+b^2}\\
\cos\theta & = & \frac{a}{c} \\
\sin\theta & = & \frac{b}{c}.
\end{eqnarray*}
\end{exercise}

\begin{exercise}  \label{c4.2.3b}
Using Exercise~\ref{c4.2.3a} observe that the matrix
$\mattwo{3}{4}{-4}{3}$ rotates the plane counterclockwise through
an angle $\theta$ and then dilates the planes by a factor of $c$.
Find $\theta$ and $c$.  Use {\tt map} to verify your results.
\end{exercise}


\CEXER


\noindent In Exercises~\ref{c4.2.a4a} -- \ref{c4.2.a4c} use {\sf map} to
find vectors that are stretched and/or contracted to a multiple of
themselves by the given linear mapping.  {\bf Hint:}  Choose a vector in the
{\sf MAP Display} window and apply {\sf Map} several times.
\begin{exercise} \label{c4.2.a4a}
$A=\mattwoc{2}{0}{1.5}{0.5}$.
\end{exercise}
\begin{exercise} \label{c4.2.a4b}
$B=\mattwo{1.2}{-1.5}{-0.4}{1.2}$.
\end{exercise}
\begin{exercise} \label{c4.2.a4c}
$C=\mattwo{2}{-1.25}{0}{-0.5}$.
\end{exercise}

\noindent In Exercises~\ref{c4.2.ba} -- \ref{c4.2.bc} use
Exercise~\ref{c4.2.3a} and {\sf map} to verify that the given matrices
rotate the plane through an angle $\theta$ followed by a dilatation $cI_2$.
Find $\theta$ and $c$ in each case.
\begin{exercise}  \label{c4.2.ba}
$A=\mattwo{1}{-2}{2}{1}$.
\end{exercise}
\begin{exercise}  \label{c4.2.bb}
$B=\mattwo{-2.4}{-0.2}{0.2}{-2.4}$.
\end{exercise}
\begin{exercise}  \label{c4.2.bc}
$C=\mattwoc{2.67}{1.3}{-1.3}{2.67}$.
\end{exercise}


\noindent In Exercises~\ref{c4.2.4a} -- \ref{c4.2.4e} use {\sf map} to
help describe the planar motions of the associated linear mappings for the
given $2\times 2$ matrix.
\begin{exercise} \label{c4.2.4a}
$A=\mattwo{\frac{\sqrt{3}}{2}}{\frac{1}{2}}{-\frac{1}{2}}{\frac{\sqrt{3}}{2}}$.
\end{exercise}
\begin{exercise} \label{c4.2.4b}
$B = \mattwo{\frac{1}{2}}{-\frac{1}{2}}{\frac{1}{2}}{\frac{1}{2}}$.
\end{exercise}
\begin{exercise} \label{c4.2.4c}
$C = \mattwo{0}{1}{1}{0}$.
\end{exercise}
\begin{exercise} \label{c4.2.4d}
$D = \mattwo{1}{0}{0}{0}$.
\end{exercise}
\begin{exercise} \label{c4.2.4e}
$E = \mattwo{\frac{1}{2}}{\frac{1}{2}}{\frac{1}{2}}{\frac{1}{2}}$.
\end{exercise}

\begin{exercise}  \label{c4.2.5}
The matrix 
\[
A = \mattwo{0}{-1}{-1}{0}
\]
reflects the $xy$-plane across the diagonal line $y=-x$ while the matrix
\[
B=\mattwo{-1}{0}{0}{-1}
\]
rotates the plane through an angle of $180^\circ$. Using the program 
{\sf map} verify that both matrices map the vector $(1,1)$ to its negative
$(-1,-1)$.  Now perform two experiments.  First using the {\sf icon} menu in 
{\sf map} place a dog icon at about the point $(1,1)$ and move that dog by 
matrix $A$.  Then replace the dog in its orginal position near $(1,1)$ and 
move that dog using matrix $B$.  Describe the difference in the result.
\end{exercise}   


\end{document}
