\documentclass{ximera}

\input{../preamble.tex}

\title{Matrices and Linearity}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

  \label{chap:matrices}

\normalsize

In this chapter we take the first step in abstracting vectors and matrices 
to mathematical objects that are more than just arrays of numbers.  We begin 
the discussion in Section~\ref{S:4.1} by introducing the multiplication of a 
matrix times a vector.  Matrix multiplication simplifies the way in which we 
write systems of linear equations and is the way by which we view matrices as 
mappings.  This latter point is discussed in Section~\ref{s:4.2}.

The mappings that are produced by matrix multiplication are special and
are called {\em linear mappings}.  Some properties of
linear maps are discussed in Section~\ref{S:linearity}.  One consequence of
linearity is the {\em principle of superposition\/} that enables
solutions to systems of linear equations to be built out of simpler
solutions.  This principle is discussed in Section~\ref{S:Superposition}.


In Section~\ref{S:4.6} we introduce 
multiplication of two matrices and discuss properties of this multiplication
in Section~\ref{S:4.7}.  Matrix multiplication is defined in terms of
composition of linear mappings which leads to an explicit formula for matrix
multiplication.   This dual role of multiplication of two matrices ---
first by formula and second as composition --- enables us to solve
linear equations in a conceptual way as well as in an algorithmic way.
The conceptual way of solving linear equations is through the use of
matrix inverses (or inverse mappings) which is described in
Section~\ref{S:SLS}.  In this section we also present
important properties of matrix inversion and a method of computation
of matrix inverses.  There is a simple formula for computing inverses
of $2\times 2$ matrices based on determinants.  The chapter ends with a 
discussion of determinants of $2\times 2$ matrices in Section~\ref{S:det2x2}.


\end{document}
