\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}
\begin{exercise} \label{c14.7.9}
Determine the equilibria in the pendulum equations \eqref{e:pendulum}.
Refer to the phase portrait of the ideal pendulum in Figure~\ref{F:ppen}
and describe in words the motion of the pendulum corresponding to a periodic 
solution surrounding one of the centers.  Describe the motion of the 
pendulum corresponding to the initial condition $(x(0),y(0))=(0,2)$.

\begin{solution}
\ans The equilibria of \eqref{e:pendulum} are
$(x,y)=(k\pi,0)$ where $k$ is an integer.

\soln Physically there are two different types of equilibria corresponding
to the two situations that the pendulum is in a steady straight down
position or in a steady straight up position.  The straight down
positions correspond to the equilibria for $k$ even and the straight
up positions to the equilibria with $k$ odd.  The straight down
states are centers whereas the straight up states are saddles.

The periodic solutions surrounding the centers (i.e.\ $k$ even) correspond to 
motions where the pendulum is hanging down and swinging back and forth.
Indeed, from Figure~\ref{F:ppen} we can see that the periodic solutions
cross the $x$-axis vertically, and these points correspond to the turning
points of the swinging motion of the pendulum where its velocity vanishes.

The orbits that do not intersect the $x$-axis -- for instance the
solution with initial condition $(0,2)$ -- correspond to motions where
the velocity of the pendulum never vanishes.  Thus, the pendulum is
in a spinning motion going in a clockwise ($y>0$) or counterclockwise
($y<0$) direction.



\end{solution}
\end{exercise}
\end{document}
