\documentclass{article}
\usepackage{hyperref}
\usepackage{fullpage}
\usepackage{enumerate}

\setlength{\parindent}{0in}
\setlength{\parskip}{1ex}


\title{Guide: Worksheet Construction}
\author{Jim Fowler \and Marty Golubitsky}

\usepackage{xcolor,verbatimbox}
\catcode`>=\active %
\catcode`<=\active %
\def\openesc{\color{red}}
\def\closeesc{\color{black}}
\def\vbdelim{\catcode`<=\active\catcode`>=\active%
\def<{\openesc}
\def>{\closeesc}}
\catcode`>=12 %
\catcode`<=12 %

\begin{document}

\maketitle
\tableofcontents

In order to create worksheets, you will need access to the exercise
files (which include short answers and solutions).  To access these, you
will need to know the \verb|\label| attached to the exercise.  At
\url{BROKEN LINK}, you will find a copy of the textbook with the
exercise \verb|\label|s displayed in the margin.

The following instructions assume you are using a Mac.

\section{Get the .cls files.}\label{get-the-.cls-files.}

In the Terminal, run

\begin{verbatim}
mkdir -p ~/Library/texmf/tex/latex
git clone https://github.com/XimeraProject/ximeraLatex ~/Library/texmf/tex/latex/ximeraLatex
\end{verbatim}

\section{Download resources}\label{download-resources}

There are three steps:
\begin{enumerate}[\bfseries(a)]
\item First, download the \LaTeX\ files for the textbook.

\begin{verbatim}
mkdir ~/linear-algebra
git clone https://github.com/mooculus/laode ~/linear-algebra/laode
\end{verbatim}

\item Then, download the worksheet builder using
\begin{verbatim}
git clone https://github.com/kisonecat/worksheet-builder ~/linear-algebra/worksheet-builder
\end{verbatim}
The result will be a directory
\verb|~/linear-algebra/worksheet-builder| populated with scripts like
\verb|build-worksheet.rb| which copy and rearrange the \LaTeX\ source
of the textbook to provide worksheets.

\item Finally, install the other software needed to run the worksheet builder.

\begin{verbatim}
gem install optparse-pathname
\end{verbatim}

If you lack permissions to run \texttt{gem}, try

\begin{verbatim}
gem install --user-install optparse-pathname
\end{verbatim}

to install it locally.
\end{enumerate}

\section{Build the book}\label{build-the-book}

Compile the book. This step is needed because various \texttt{aux} files
are created that will be read by the worksheet builder.

\begin{verbatim}
cd ~/linear-algebra/laode
lualatex --shell-escape linearAlgebra
lualatex --shell-escape linearAlgebra
\end{verbatim}

\section{Build a worksheet}\label{try-it}

Now you are ready to build a worksheet.

\begin{enumerate}[\bfseries(a)]
\item \textbf{Go to the worksheet builder directory.}

\begin{verbatim}
cd ~/linear-algebra/worksheet-builder
\end{verbatim}

There is a sample worksheet source at \verb|homework-sample.tex|.
To build the associated sample worksheet, run the command
\begin{verbatim}
./build-both.sh --root=$HOME/linear-algebra/laode homework-sample.tex
\end{verbatim}
Among many output files, you should see two PDF files
\begin{enumerate}
\item one worksheet with solutions called \texttt{homework-samples.pdf} and
\item one worksheet without solutions called \texttt{homework-samplee.pdf}.
\end{enumerate}
Here is a helpful mnemonic: the filename
\texttt{homework-sample\textcolor{red}{s}.pdf} has an \texttt{s} for
\texttt{s}olutions and the filename
\texttt{homework-sample\textcolor{red}{e}.pdf} has an \texttt{e} for
\texttt{e}xercises.  It is helpful to provide the PDF with solutions
both to the grader and to the students after the due date.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \textbf{Create the worksheet source.} You can create your own file
  with a name like \texttt{homework1.tex} with the following content:

\vspace{2ex}\hrule

\begin{verbnobox}[\vbdelim\mbox{}]
\documentclass{article}

\input{preamble}

\title{<Homework 1>}
\author{<Instructor Name>}
\date{Due: <Jan 1, 2078> at <9:10am>}

\begin{document}
\maketitle

\exercise{<label.for.problem.1>}

\exercise{<label.for.problem.2>}

\end{document}
\end{verbnobox}

\hrule\vspace{2ex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \textbf{Create the worksheet.}
Now run

\begin{verbatim}
./build-both.sh --root=$HOME/linear-algebra/laode homework1.tex
\end{verbatim}

Two PDF files will be created:
\begin{enumerate}
\item one with solutions called \texttt{homework1s.pdf} and
\item one without solutions called \texttt{homework1e.pdf}.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \textbf{If desired, edit the \LaTeX\ files.} Two \LaTeX\ files are also created:
\begin{enumerate}
\item one with solutions called \texttt{homework1s.tex} and
\item one without solutions called \texttt{homework1e.tex}.
\end{enumerate}
You may edit the \LaTeX\ files by hand if you wish.  For example, you
may wish to insert \verb|\clearpage| to provide more logical
page-breaks.  Be warned that these auto-generated \LaTeX\ files may be
challenging to read.
\end{enumerate}

\section{Send us new exercises}

Our goal is to increase the number of available exercises.  Your
contributions will be a tremendous help in reaching this goal.  A good
source of new exercises will be your exams.

It is easiest for us if you submit a separate \LaTeX\ file for each
new exercise.  It would be helpful if you use the format provided in
\href{https://github.com/mooculus/laode/blob/master/GUIDES/template.tex}{\texttt{template.tex}}.

As in the template, your contribution should include the exercise
written in \LaTeX, and below it the solution (preferably including an
answer).

Please include the section where the new exercise should appear.  It
would also be helpful if you provide with the label of the existing
exercise that the new exercise should follow.

MATLAB exercises should be marked as such.  If needed, please include
new \verb|.m|-files.

\end{document}
