\documentclass{article}
\usepackage{hyperref}
\usepackage{fullpage}
\usepackage{enumerate}

\setlength{\parindent}{0in}
\setlength{\parskip}{1ex}

\title{Guide: Worksheet Construction via CoCalc}
\author{Jim Fowler \and Marty Golubitsky}

\usepackage{xcolor,verbatimbox}
\catcode`>=\active %
\catcode`<=\active %
\def\openesc{\color{red}}
\def\closeesc{\color{black}}
\def\vbdelim{\catcode`<=\active\catcode`>=\active%
\def<{\openesc}
\def>{\closeesc}}
\catcode`>=12 %
\catcode`<=12 %

\begin{document}

\maketitle
\tableofcontents

In the instructor edition of the textbook, you will notice a
\verb|\label| attached to the exercise and displayed in the margin.
You can create worksheets from these labels.

\section{Access CoCalc}\label{co-calc}

Rather than installing the ``worksheet builder'' on your own computer,
you can use CoCalc, a cloud-hosted platform.

\begin{enumerate}[\bfseries(a)]
\item First, visit \url{https://cocalc.com/} and
\item sign up for an account.
\item Email \texttt{fowler@math.osu.edu} with the email address that you used to register for CoCalc.
\end{enumerate}
You will then be added to the ``Math 2568'' project.

\section{Build a worksheet}\label{try-it}

Once you are added to the Math 2568 project, you are ready to build a
worksheet.

\begin{enumerate}[\bfseries(a)]
  \item Visit \url{https://cocalc.com/}
  \item log in again.
  \item Click on ``projects'' along the top bar of CoCalc.
  \item Choose ``Math 2568,'' a shared project.
  \item Click on the ``worksheet-builder'' folder.
  \item Click on the button labeled ``$\oplus$ New'' to create a new file.
  \item In the text field labeled ``Name your file'' enter a name for your worksheet.
  \item Then click on ``LaTeX document.''
  \item Replace the provided content with a worksheet template, such as
\begin{verbatim}
\documentclass{article}

\input{worksheet-preamble}

\title{Math 2568 Homework 7}
\author{Dr. Instructor}
\date{Due: Jan 1, 2078 at 9:00am}

\begin{document}
\maketitle

\exercise{c4.10A.1b}
\exercise{c4.10A.4a}
\exercise{c5.1.3}

\end{document}
\end{verbatim}
  \item On the bottom-right under ``Build Control and Log'' there is a ``Build Command.''  Replace this with
    \texttt{./build-both.sh your-chosen-filename.tex} and click the ``Build'' button.
\item If you return to the ``worksheet-builder'' folder in CoCalc, you will see two PDF files, 
\begin{enumerate}
\item one worksheet with solutions called \texttt{homework-samples.pdf} and
\item one worksheet without solutions called \texttt{homework-samplee.pdf}.
\end{enumerate}
Here is a helpful mnemonic: the filename
\texttt{homework-sample\textcolor{red}{s}.pdf} has an \texttt{s} for
\texttt{s}olutions and the filename
\texttt{homework-sample\textcolor{red}{e}.pdf} has an \texttt{e} for
\texttt{e}xercises.  It is helpful to provide the PDF with solutions
both to the grader and to the students after the due date.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \textbf{If desired, edit the \LaTeX\ files.} Two \LaTeX\ files are also created:
\begin{enumerate}
\item one with solutions called \texttt{homework1s.tex} and
\item one without solutions called \texttt{homework1e.tex}.
\end{enumerate}
You may edit the \LaTeX\ files by hand if you wish.  For example, you
may wish to insert \verb|\clearpage| to provide more logical
page-breaks.  Be warned that these auto-generated \LaTeX\ files may be
challenging to read.
\end{enumerate}

\section{Send us new exercises}

Our goal is to increase the number of available exercises.  Your
contributions will be a tremendous help in reaching this goal.  A good
source of new exercises will be your exams.

It is easiest for us if you submit a separate \LaTeX\ file for each
new exercise.  It would be helpful if you use the format provided in
\href{https://github.com/mooculus/laode/blob/master/GUIDES/template.tex}{\texttt{template.tex}}.

As in the template, your contribution should include the exercise
written in \LaTeX, and below it the solution (preferably including an
answer).

Please include the section where the new exercise should appear.  It
would also be helpful if you provide with the label of the existing
exercise that the new exercise should follow.

MATLAB exercises should be marked as such.  If needed, please include
new \verb|.m|-files.

\end{document}
