\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}

In Exercises~\ref{A5.6.4}-\ref{A5.6.9} decide whether the statement is true or false, and explain your answer.

\begin{exercise}  \label{A5.6.4}
Every set of three vectors in $\R^3$ is a basis for $\R^3$.
\begin{solution}
\ans False  

\soln
The vectors could be linearly independent. For example $\{e_1,e_2,e_1+e_2\}$ is not a basis for $\R^3$.
\end{solution}
\end{exercise}

\begin{exercise}  \label{A5.6.5}
Every set of four vectors in $\R^3$ is linearly dependent.
\begin{solution}
\ans True

\soln
The dimension of $\R^3$ is three, so any set of more than three vectors in $R^3$ is necessarily linearly dependent.
\end{solution}
\end{exercise}

\begin{exercise}  \label{A5.6.6}
If $\{v_1,v_2\}$ is a basis for the plane $z = 0$ in $\R^3$, then $\{v_1,v_2,e_3\}$ is a basis for $\R^3$.
\begin{solution}
\ans True
\soln
The vector $e_3$ is not contained in the plane $z=0$, so Lemma~\ref{extendindep} implies that $\{v_1,v_2,e_3\}$  is linearly independent. Therefore, $\{v_1,v_2,e_3\}$  is a basis for $\R^3$, because any set of three linearly independent vectors in a vector space of dimension three is a basis for that vector space. 

Alternatively, $\{e_1,e_2,e_3\}$ is a basis for $\R^3$, so if $w\in \R^3$, then $w = a_1 e_1 +a_2 e_2 +a_3 e_3$ for some scalars $a_1$ and $a_2$. It follows that $a_1 e_1 + a_2 e_2$ is a vector in the plane $z=0$, so there exist scalars $b_1$ and $b_2$ so $a_1 e_1+ a_2 e_2 = b_1 v_1 +b_2 v_2$. Therefore, $w = b_1 v_1 + b_2 v_2 + a_3 e_3$, so $w\in \text{span}\{v_1,v_2,e_3\}$. Therefore, $\{v_1,v_2,e_3\}$ spans $\R^3$, and is a basis for $\R^3$ because any set of three linearly independent vectors in a vector space of dimension three is a basis for that vector space. 
\end{solution}
\end{exercise}

\begin{exercise}  \label{A5.6.7}
If $\{v_1,v_2,v_3\}$ is a basis for $\R^3$, the only subspaces of $R^3$ of dimension one are $\text{span}\{v_1\}, \text{span}\{v_2\}$, and $\text{span}\{v_3\}$. 
\begin{solution}
\ans False 
\soln For example, $\{e_1,e_2,e_3\}$ is a basis for $\R^3$ and $\text{span}\{e_1+e_2\}$ does not equal the $x$, $y$, or $z$-axes.
\end{solution}
\end{exercise}

\begin{exercise}  \label{A5.6.8}
The only subspace of $\R^3$ that contains finitely many vectors is $\{ 0 \}$.
\begin{solution}
\ans True
\soln If a subspace of $\R^3$ contains a nonzero vector, it must contain all scalar multiples of that vector.
\end{solution}
\end{exercise}

\begin{exercise}  \label{A5.6.9}
If $U$ is a subspace of $\R^3$ of dimension 1 and $V$ is a subspace of $\R^3$ of dimension 2, then $U\cap V=\{0\}$.
\begin{solution}
\ans False 
\soln $U\cap V$ is always subspace of $\mathbb{R}^3,$ but its dimension could be one. For example, if $U$ is the $x$-axis ( $U=\text{span}\{e_1\}$) and $V$ is the $xy$-plane ($V=\text{span}\{e_1,e_2\}$) then $U\cap V=U.$
$U\cap V$ is always subspace of $\R^3$, but its dimension could be one. For example, if $U$ is the $x$-axis ($U=\text{span}\{e_1\}$) and $V$ is the $xy$-plane ($V=\text{span}\{e_1,e_2\}$), then $U\cap V=U$.
\end{solution}
\end{exercise}
\end{document}
