\documentclass{ximera}

\input{../preamble.tex}

\title{Gaussian Elimination}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

  \label{S:Gauss}

A general system of $m$ {\em linear\/} equations \index{linear}
in $n$ unknowns has the form
\begin{equation}    \label{general}
\arraycolsep 2pt
\begin{array}{rcrcrcrcrl}
 a_{11}x_1 & + & a_{12}x_2 & + & \cdots & + & a_{1n}x_n & = &   b_1
& \\
 a_{21}x_1 & + & a_{22}x_2 & + & \cdots & + & a_{2n}x_n & = &   b_2
& \\
        & \vdots &      & \vdots &    & \vdots &     & \vdots &
  & \\
 a_{m1}x_1 & + & a_{m2}x_2 & + & \cdots & + & a_{mn}x_n & = &   b_m
& .
\end{array}
\end{equation}
The entries $a_{ij}$ and $b_i$ are constants.  Our task is to find
a method for solving \eqref{general} for the variables
$x_1,\ldots,x_n$.

\subsection*{Easily Solved Equations}

Some systems are easily solved.  The system of three
equations ($m=3$) in three unknowns ($n=3$)
\begin{equation}
\begin{array}{rcrcrcrl} \label{examp3}
  x_1 & + & 2x_2 & + & 3x_3 & = &  10  & \\
      &   &  x_2 & - & \frac{1}{5}x_3 & = & \frac{7}{5}  & \\
      &   &      &   &  x_3 & = &   3  & \\
\end{array}
\end{equation}
is one example.  The $3^{rd}$ equation states that $x_3=3$.
Substituting this value into the $2^{nd}$ equation allows us to
solve the $2^{nd}$ equation for $x_2=2$.  Finally, substituting
$x_2=2$ and $x_3=3$ into the $1^{st}$ equation allows us to
solve for $x_1=-3$.  The process that we have just described is
called {\em back substitution\/}.\index{back substitution}

Next, consider the system of two equations ($m=2$) in three
unknowns ($n=3$):
\begin{equation}  \label{e23}
\begin{array}{rcrcrcrl}
  x_1 & + & 2x_2 & + & 3x_3 & = &  10  & \\
      &   &      &   &  x_3 & = &   3  & \! . \\
\end{array}
\end{equation}
The $2^{nd}$ equation in \eqref{e23} states that $x_3=3$.
Substituting
this value into the $1^{st}$ equation leads to the equation
\[
x_1 = 1-2x_2.
\]
We have shown that every solution to \eqref{e23} has the form
$(x_1,x_2,x_3)=(1-2x_2,x_2,3)$ and that every vector
$(1-2x_2,x_2,3)$ is a solution of \eqref{e23}.  Thus, there is an
infinite number of solutions to \eqref{e23}, and these solutions
can be parameterized by one number $x_2$.

\subsection*{Equations Having No Solutions}

Note that the system of equations
\begin{eqnarray*}
x_1 - x_2 = 1\\
x_1 - x_2 = 2
\end{eqnarray*}
has no solutions.

\begin{definition}
A linear system of equations is {\em inconsistent\/} if the
system has no solutions and {\em consistent\/} if the system
does have solutions.
\end{definition} \index{consistent} \index{inconsistent}

As discussed in the previous section, \eqref{incon} is an example
of a linear system that \Matlab cannot solve.  In fact, that
system is inconsistent --- inspect the $2^{nd}$ and $4^{th}$
equations in \eqref{incon}.

Gaussian elimination is an algorithm for finding all solutions
to a system of linear equations by reducing the given system to
ones like \eqref{examp3} and \eqref{e23}, that are easily solved by
back substitution.  Consequently, Gaussian elimination can also be
used to determine whether a system is consistent or inconsistent.

\subsection*{Elementary Equation Operations}

There are three ways to change a system of equations without
changing the set of solutions; Gaussian elimination
\index{Gaussian elimination} is based on this observation.  The
three elementary operations are:
\begin{enumerate}
\item   Swap two equations.
\item   Multiply a single equation by a nonzero number.
\item   Add a scalar multiple of one equation to another.
\end{enumerate}

We begin with an example:
\begin{equation}
\begin{array}{rcrcrcrl}
  x_1 & + & 2x_2 & + & 3x_3 & = &  10  & \\
  x_1 & + & 2x_2 & + &  x_3 & = &   4  & \\
 2x_1 & + & 9x_2 & + & 5x_3 & = &  27  & .\\
\end{array}
\end{equation}
Gaussian elimination works by eliminating variables from the
equations in a fashion similar to the substitution method in the
previous section.  To begin, eliminate the variable $x_1$ from
all but the $1^{st}$ equation, as follows.  Subtract the
$1^{st}$ equation from the $2^{nd}$, and subtract twice the
$1^{st}$ equation from the $3^{rd}$, obtaining:
\begin{equation}
\begin{array}{rcrcrcrl}
  x_1 & + & 2x_2 & + & 3x_3 & = &  10  & \\
      &   &      &   &-2x_3 & = &  -6  & \\
      &   & 5x_2 & - &  x_3 & = &   7  & .\\
\end{array}
\end{equation}
Next, swap the $2^{nd}$ and $3^{rd}$ equations, so that
the coefficient of $x_2$ in the new $2^{nd}$ equation is nonzero.
This yields
\begin{equation}
\begin{array}{rcrcrcrl}
  x_1 & + & 2x_2 & + & 3x_3 & = &  10  & \\
      &   & 5x_2 & - &  x_3 & = &   7  & \\
      &   &      &   &-2x_3 & = &  -6  & .\\
\end{array}
\end{equation}
Now, divide the $2^{nd}$ equation by $5$ and the $3^{rd}$
equation by $-2$ to obtain a system of equations identical to
our first example \eqref{examp3}, which we solved by back
substitution.


\subsection*{Augmented Matrices}

The process of performing Gaussian elimination when the number
of equations is greater than two or three is painful.  The
computer, however, can help with the manipulations.  We begin by
introducing the {\em augmented matrix\/}. \index{matrix!augmented}
The augmented matrix associated with \eqref{general} has
$m$ rows and $n+1$ columns and is written as:
\begin{equation}  \label{augmented}
\left(
\begin{array}{rrrr|r}
 a_{11} & a_{12} & \cdots & a_{1n} &  b_1 \\
 a_{21} & a_{22} & \cdots & a_{2n} &  b_2 \\
 \vdots & \vdots &        & \vdots & \vdots \\
 a_{m1} & a_{m2} & \cdots & a_{mn} &  b_m
\end{array}
\right)
\end{equation}
The augmented matrix contains all of the information that is
needed to solve system \eqref{general}.

\subsection*{Elementary Row Operations} \index{elementary row
operations}

The elementary operations used in Gaussian elimination
can be interpreted as {\em row operations\/} on
the augmented matrix, as follows:
\begin{enumerate}
\item   Swap two rows.
\item   Multiply a single row by a nonzero number.
\item   Add a scalar multiple of one row to another.
\end{enumerate}
We claim that by using these elementary row operations
intelligently, we can always solve a consistent linear system
--- indeed, we can determine when a linear system is consistent
or inconsistent.  The idea is to perform elementary row
operations in such a way that the new augmented matrix has zero
entries below the diagonal.

We describe this process inductively.  Begin with the $1^{st}$
column.  We assume for now that some entry in this column is
nonzero.  If $a_{11}=0$, then swap two rows so that the
number $a_{11}$ is nonzero.  Then divide the $1^{st}$ row by
$a_{11}$ so that the leading entry in that row is $1$.  Now
subtract $a_{i1}$ times the $1^{st}$ row from the $i^{th}$ row
for each row $i$ from $2$ to $m$.  The end result is that the
$1^{st}$ column has a $1$ in the $1^{st}$ row and a $0$ in every
row below the $1^{st}$.  The result is
\[
\left(\begin{array}{cccc}  1 & * & \cdots & * \\
0 & * & \cdots & * \\ \vdots & \vdots & \vdots & \vdots \\
0 & * & \cdots & * \end{array} \right).
\]


Next we consider the $2^{nd}$ column.  We assume that some entry
in that column below the $1^{st}$ row is nonzero.  So, if
necessary, we can swap two rows below the $1^{st}$ row so
that the entry $a_{22}$ is nonzero.  Then we divide the $2^{nd}$
row by $a_{22}$ so that its leading nonzero entry is $1$.  Then
we subtract appropriate multiples of the $2^{nd}$ row from each
row below the $2^{nd}$ so that all the entries in the $2^{nd}$
column below the $2^{nd}$ row are $0$.  The result is
\[
\left(\begin{array}{cccc}  1 & * & \cdots & * \\
0 & 1 & \cdots & * \\ \vdots & \vdots & \vdots & \vdots \\
0 & 0 & \cdots & * \end{array} \right).
\]

Then we continue with the $3^{rd}$ column.  That's the idea.  However,
does this process always work and what happens if all of the entries
in a column are zero?  Before answering these questions we do
experimentation with \Matlabp.

\subsection*{Row Operations in \Matlab}\index{elementary row
operations!in \protect\Matlab}

In \Matlab the $i^{th}$ row of a matrix {\tt A} is specified by
{\tt A(i,:)}.  Thus to replace the $5^{th}$ row of a matrix {\tt
A} by twice itself, we need only type:
\begin{verbatim}
A(5,:) = 2*A(5,:)
\end{verbatim}
In general, we can replace the $i^{th}$ row of the matrix
{\tt A} by $c$ times itself by typing
\begin{verbatim}
A(i,:) = c*A(i,:)
\end{verbatim}
Similarly, we can divide the $i^{th}$ row of the matrix {\tt A}
by the nonzero number $c$ by typing
\begin{verbatim}
A(i,:) = A(i,:)/c
\end{verbatim}

The third elementary row operation is performed similarly.
Suppose we want to add $c$ times the $i^{th}$ row to the
$j^{th}$ row, then we type
\begin{verbatim}
A(j,:) = A(j,:) + c*A(i,:)
\end{verbatim}
For example, subtracting $3$ times the $7^{th}$ row from the
$4^{th}$ row of the matrix {\tt A} is accomplished by typing:
\begin{verbatim}
A(4,:) = A(4,:) - 3*A(7,:)
\end{verbatim}

The first elementary row operation, swapping two rows, requires
a different kind of \Matlab command.  In \Matlabp, the $i^{th}$
and $j^{th}$ rows of the matrix {\tt A} are permuted by the
command
\begin{verbatim}
A([i j],:) = A([j i],:)
\end{verbatim}
So, to swap the $1^{st}$ and $3^{rd}$ rows of the matrix
{\tt A}, we type
\begin{verbatim}
A([1 3],:) = A([3 1],:)
\end{verbatim}  \index{\computer!A([1 3],:)}

\subsection*{Examples of Row Reduction in \Matlab}

Let us see how the row operations can be used in \Matlabp.  As
an example, we consider the augmented matrix
\begin{matlabEquation}  \label{examp4}
\left(
\begin{array}{rrrr|r}
 1  &  3  &  0  & -1  &  -8\\
 2  &  6  & -4  &  4  &   4\\
 1  &  0  & -1  & -9  & -35\\
 0  &  1  &  0  &  3  &  10
\end{array}
\right)
\end{matlabEquation}

We enter this information into \Matlab by typing
\begin{verbatim}
e2_3_8
\end{verbatim}
which produces the result
\begin{verbatim}
A =
     1     3     0    -1    -8
     2     6    -4     4     4
     1     0    -1    -9   -35
     0     1     0     3    10
\end{verbatim}

We now perform Gaussian elimination on {\tt A}, and then solve the 
resulting system by back substitution.  Gaussian elimination uses 
elementary row operations to set the entries that are in the lower 
left part of {\tt A} to zero. These entries are indicated by
numbers in the following matrix:
\begin{verbatim}
    *     *     *     *     *
    2     *     *     *     *
    1     0     *     *     *
    0     1     0     *     *
\end{verbatim}

Gaussian elimination \index{Gaussian elimination} works
inductively.  Since the first entry in the matrix $A$ is equal
to $1$, the first step in Gaussian elimination is to set to zero
all entries in the $1^{st}$ column below the $1^{st}$ row.  We
begin by eliminating the {\tt 2} that is the first entry in the
$2^{nd}$ row of {\tt A}.  We replace the $2^{nd}$ row by the
$2^{nd}$ row minus twice the $1^{st}$ row.  To accomplish this
elementary row operation, we type
\begin{verbatim}
A(2,:) = A(2,:) - 2*A(1,:)
\end{verbatim}
and the result is
\begin{verbatim}
A =
     1     3     0    -1    -8
     0     0    -4     6    20
     1     0    -1    -9   -35
     0     1     0     3    10
\end{verbatim}
In the next step, we eliminate the {\tt 1} from the entry in the
$3^{rd}$ row, $1^{st}$ column of {\tt A}.  We do this by
typing
\begin{verbatim}
A(3,:) = A(3,:) - A(1,:)
\end{verbatim}
which yields
\begin{verbatim}
A =
     1     3     0    -1    -8
     0     0    -4     6    20
     0    -3    -1    -8   -27
     0     1     0     3    10
\end{verbatim}

Using elementary row operations, we have now set the entries
in the $1^{st}$ column below the $1^{st}$ row to $0$.  Next,
we alter the $2^{nd}$ column.  We begin by swapping
the $2^{nd}$ and $4^{th}$ rows so that the leading nonzero entry
in the $2^{nd}$ row is $1$.   To accomplish this swap, we type
\begin{verbatim}
A([2 4],:) = A([4 2],:)
\end{verbatim}
and obtain
\begin{verbatim}
A =
     1     3     0    -1    -8
     0     1     0     3    10
     0    -3    -1    -8   -27
     0     0    -4     6    20
\end{verbatim}
The next elementary row operation is the command
\begin{verbatim}
A(3,:) = A(3,:) + 3*A(2,:)
\end{verbatim}
which leads to
\begin{verbatim}
A =
     1     3     0    -1    -8
     0     1     0     3    10
     0     0    -1     1     3
     0     0    -4     6    20
\end{verbatim}
Now we have set all entries in the $2^{nd}$ column below
the $2^{nd}$ row to $0$.

Next, we set the first nonzero entry in the $3^{rd}$ row to {\tt 1}
by multiplying the $3^{rd}$ row by $-1$, obtaining
\begin{verbatim}
A =
     1     3     0    -1    -8
     0     1     0     3    10
     0     0     1    -1    -3
     0     0    -4     6    20
\end{verbatim}

Since the leading nonzero entry in the $3^{rd}$ row is $1$,  we
next eliminate the nonzero entry in the $3^{rd}$ column,
$4^{th}$ row. This is accomplished by the following \Matlab
command:
\begin{verbatim}
A(4,:) = A(4,:) + 4*A(3,:)
\end{verbatim}
Finally, divide the $4^{th}$ row by $2$ to obtain:
\begin{verbatim}
A =
     1     3     0    -1    -8
     0     1     0     3    10
     0     0     1    -1    -3
     0     0     0     1     4
\end{verbatim}

By using elementary row operations, we have arrived at the system
\begin{equation}
\arraycolsep 2pt
\begin{array}{rcrcrcrcrl}
  x_1 & + & 3x_2 &   &     & - &  x_4 & = &  -8 & \\
      &   &  x_2 &   &     & + & 3x_4 & = &  10 & \\
      &   &      &   & x_3 & - &  x_4 & = &  -3 & \\
      &   &      &   &     &   &  x_4 & = &   4 &\! ,
\end{array}
\end{equation}
that can now be solved by back substitution.  We obtain
\begin{equation} \label{ans1}
        x_4 = 4,\quad x_3 = 1,\quad x_2 = -2,\quad x_1 = 2.
\end{equation}
We return to the original set of equations corresponding to
\eqref{examp4}
\begin{matlabEquation}    \label{examp4_con}
\arraycolsep 2pt
\begin{array}{rcrcrcrcrl}
  x_1 & + & 3x_2 &   &      & - &  x_4 & = &  -8  & \\
 2x_1 & + & 6x_2 & - & 4x_3 & + & 4x_4 & = &   4  & \\
  x_1 &   &      & - &  x_3 & - & 9x_4 & = & -35  & \\
      &   &  x_2 &   &      & + & 3x_4 & = &  10  & \! .
\end{array}
\end{matlabEquation}
Load the corresponding linear system into \Matlab by typing
\begin{verbatim}
e2_3_11
\end{verbatim}
The information in \eqref{examp4_con} is contained in the
coefficient matrix {\tt C} and the right hand side {\tt b}.
A direct solution is found by typing
\begin{verbatim}
x = C\b
\end{verbatim}
which yields the same answer as in \eqref{ans1}, namely,
\begin{verbatim}
x =
    2.0000
   -2.0000
    1.0000
    4.0000
\end{verbatim}

\subsection*{Introduction to Echelon Form} \index{echelon form}

Next, we discuss how Gaussian elimination works in an example
in which the number of rows and the number of columns in the
coefficient matrix are unequal.  We consider the augmented
matrix\index{matrix!augmented}
\begin{matlabEquation}  \label{examp5}
\left(
\begin{array}{rrrrrr|r}
 1  &  0  & -2  &  3  &  4  &  0  &  1\\
 0  &  1  &  2  &  4  &  0  & -2  &  0\\
 2  & -1  & -4  &  0  & -2  &  8  & -4\\
-3  &  0  &  6  & -8  &-12  &  2  & -2
\end{array}
\right)
\end{matlabEquation}
This information is entered into \Matlab by typing
\begin{verbatim}
e2_3_12
\end{verbatim}
Again, the augmented matrix is denoted by {\tt A}.

We begin by eliminating the {\tt 2} in the entry in the $3^{rd}$
row, $1^{st}$ column.  To accomplish the corresponding
elementary row operation, we type
\begin{verbatim}
A(3,:) = A(3,:) - 2*A(1,:)
\end{verbatim}
resulting in
\begin{verbatim}
A =
     1     0    -2     3     4     0     1
     0     1     2     4     0    -2     0
     0    -1     0    -6   -10     8    -6
    -3     0     6    -8   -12     2    -2
\end{verbatim}
We proceed with
\begin{verbatim}
A(4,:) = A(4,:) + 3*A(1,:)
\end{verbatim}
to create two more zeros in the $4^{th}$ row.  Finally, we
eliminate the {\tt -1} in the $3^{rd}$ row, $2^{nd}$
column by
\begin{verbatim}
A(3,:) = A(3,:) + A(2,:)
\end{verbatim}
to arrive at
\begin{verbatim}
A =
     1     0    -2     3     4     0     1
     0     1     2     4     0    -2     0
     0     0     2    -2   -10     6    -6
     0     0     0     1     0     2     1
\end{verbatim}
Next we set the leading nonzero entry in the $3^{rd}$ row to $1$
by dividing the $3^{rd}$ row by $2$.   That is, we type
\begin{verbatim}
A(3,:) = A(3,:)/2
\end{verbatim}
to obtain
\begin{verbatim}
A =
     1     0    -2     3     4     0     1
     0     1     2     4     0    -2     0
     0     0     1    -1    -5     3    -3
     0     0     0     1     0     2     1
\end{verbatim}
We say that the matrix {\tt A} is in (row) {\em echelon form\/}
\index{echelon form} since the first nonzero entry in each row
is a {\tt 1}, each entry in a column below a leading {\tt 1} is
{\tt 0}, and the leading {\tt 1} moves to the right as you go
down the matrix.  In row echelon form, the entries where leading
$1$'s occur are called {\em pivots\/}.\index{pivot}

If we compare the structure of this matrix to the ones we have
obtained previously, then we see that here we have two columns
too many.  Indeed, we may solve these equations by back
substitution for any choice of the variables $x_5$ and $x_6$.

The idea behind back substitution \index{back substitution} is
to solve the last equation for the variable corresponding to the
first nonzero coefficient.  In this case, we use the $4^{th}$
equation to solve for $x_4$ in terms of $x_5$ and $x_6$, and
then we substitute for $x_4$ in the first three equations.  This
process can also be accomplished by elementary row operations.
Indeed, eliminating the variable $x_4$ from the first three
equations is the same as using row operations to set the first
three entries in the $4^{th}$ column to {\tt 0}.  We can do this
by typing
\begin{verbatim}
A(3,:) = A(3,:) + A(4,:);
A(2,:) = A(2,:) - 4*A(4,:);
A(1,:) = A(1,:) - 3*A(4,:)
\end{verbatim}
{\bf Remember:} By typing semicolons after the first two rows, we
have told \Matlab not to print the intermediate results.  Since
we have not typed a semicolon after the $3^{rd}$ row, \Matlab
outputs
\begin{verbatim}
A =
     1     0    -2     0     4    -6    -2
     0     1     2     0     0   -10    -4
     0     0     1     0    -5     5    -2
     0     0     0     1     0     2     1
\end{verbatim}
We proceed with back substitution by eliminating the nonzero
entries in the first two rows of the $3^{rd}$ column.  To do
this, type
\begin{verbatim}
A(2,:) = A(2,:) - 2*A(3,:);
A(1,:) = A(1,:) + 2*A(3,:)
\end{verbatim}
which yields
\begin{verbatim}
A =
     1     0     0     0    -6     4    -6
     0     1     0     0    10   -20     0
     0     0     1     0    -5     5    -2
     0     0     0     1     0     2     1
\end{verbatim}
The augmented matrix is now in {\em reduced echelon form}
\index{echelon form!reduced} and the corresponding system of
equations has the form
\begin{equation}   \label{e:refexamp5}
\arraycolsep 2pt
\begin{array}{rcrcrcrcrcrcrl}
  x_1 &   &     &   &     &  &     & - &  6x_5 & + &  4x_6 & = &
-6 & \\
      &   & x_2 &   &     &  &     & + & 10x_5 & - & 20x_6 & = &
0 & \\
      &   &     &   & x_3 &  &     & - &  5x_5 & + &  5x_6 & = &
-2 & \\
      &   &     &   &     &  & x_4 &   &       & + &  2x_6 & = &
1 &\! ,
\end{array}
\end{equation}
A matrix is in reduced echelon form \index{echelon form!reduced}
if it is in echelon form and if {\em every\/} entry in a column
containing a pivot, other than the pivot itself, is {\tt 0}.

Reduced echelon form allows us to solve directly this system of
equations in terms of the variables $x_5$ and $x_6$,
\begin{equation}  \label{e:refexamp6}
\left(\begin{array}{c} x_1\\x_2\\x_3\\x_4\\x_5\\x_6\end{array}\right) =
\left(\begin{array}{c} -6+6x_5-4x_6\\-10x_5+20x_6\\
-2+5x_5-5x_6\\1-2x_6\\x_5\\x_6\end{array}\right).
\end{equation}
It is important to note that every consistent system of linear equations
corresponding to an augmented matrix in reduced echelon form can be
solved as in \eqref{e:refexamp6} --- and this is one reason for emphasizing
reduced echelon form.  We will discuss the reduction to reduced echelon
form in more detail in the next section.

\EXER


\TEXER

\noindent In Exercises~\ref{c2.3.6a} -- \ref{c2.3.6c} determine
whether the given matrix is in reduced echelon form.
\begin{exercise} \label{c2.3.6a}
$\left(\begin{array}{rrrr}
1 & -1 &  0 &   1   \\
0 &  1 &  0 &  -6    \\
         0 &  0 &  1 &   0   \end{array}\right)$.
     \begin{multipleChoice}
       \choice{The matrix is in reduced echelon form.}         
       \choice[correct]{The matrix is not in reduced echelon form.}         
     \end{multipleChoice}

\begin{solution}
The matrix is not in reduced echelon form.


\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.6b}
$\left(\begin{array}{rrrr}
1 &  0 & -2 &   0   \\
0 &  1 &  4 &   0    \\
         0 &  0 &  0 &   1  \end{array}\right)$.
     \begin{multipleChoice}
       \choice[correct]{The matrix is in reduced echelon form.}         
       \choice{The matrix is not in reduced echelon form.}         
     \end{multipleChoice}     

\begin{solution}
The matrix is in reduced echelon form.

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.6c}
$\left(\begin{array}{rrrr}
0 &  1 &  0 &   3   \\
0 &  0 &  2 &   1    \\
         0 &  0 &  0 &   0   \end{array}\right)$.
     \begin{multipleChoice}
       \choice{The matrix is in reduced echelon form.}         
       \choice[correct]{The matrix is not in reduced echelon form.}         
     \end{multipleChoice}     

\begin{solution}
The matrix is not in reduced echelon form.

\end{solution}
\end{exercise}

\noindent In Exercises~\ref{c2.3.7a} -- \ref{c2.3.7c} we list
the reduced echelon form of an augmented matrix of a system of
linear equations.  Which columns in these augmented matrices
contain pivots?  Describe all solutions to these systems of
equations in the form of \eqref{e:refexamp6}.
\begin{exercise} \label{c2.3.7a}
$\left(\begin{array}{rrr|r}
 1  &  4 & 0 & 0\\
 0  &  0 & 1 & 5\\
 0  &  0 & 0 & 0
       \end{array}\right)$.
     \begin{selectAll}
       \choice[correct]{The $1^{st}$ column contains a pivot.}
       \choice{The $2^{nd}$ column contains a pivot.}
       \choice[correct]{The $3^{rd}$ column contains a pivot.}
       \choice{The $4^{th}$ column contains a pivot.}
       \choice{The $5^{th}$ column contains a pivot.}
     \end{selectAll}
     
     \begin{exercise}
       The solutions of the system are:
\[
\left(\begin{array}{r} x_1 \\ x_2 \\ x_3\end{array} \right)
= \left(\begin{array}{c} \answer{-4}x_2 \\ x_2 \\ \answer{5}\end{array} \right)
\]

\begin{solution}
The $1^{st}$ and $3^{rd}$ columns of the matrix contain
pivots.  The solutions of the system are:

\[
\left(\begin{array}{r} x_1 \\ x_2 \\ x_3\end{array} \right)
= \left(\begin{array}{c} -4x_2 \\ x_2 \\ 5\end{array} \right)
\]

\end{solution}
     \end{exercise}

\begin{solution}


\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.7b}
$\left(\begin{array}{rrrr|r}
 1  &  2 & 0 & 0 & 0\\
 0  &  0 & 1 & 1 & 0\\
 0  &  0 & 0 & 0 & 1
       \end{array}\right)$.
     \begin{selectAll}
       \choice[correct]{The $1^{st}$ column contains a pivot.}
       \choice{The $2^{nd}$ column contains a pivot.}
       \choice[correct]{The $3^{rd}$ column contains a pivot.}
       \choice{The $4^{th}$ column contains a pivot.}
       \choice[correct]{The $5^{th}$ column contains a pivot.}
     \end{selectAll}
     \begin{exercise}
       The system of linear equations corresponding to this matrix has\ldots
       \begin{multipleChoice}
         \choice[correct]{No solutions.}
         \choice{Exactly one solution.}
         \choice{Infinitely many solutions.}
       \end{multipleChoice}

\begin{solution}
The $1^{st}$, $3^{rd}$, and $5^{th}$ columns of the matrix
contain pivots.  Since the last row of the matrix translates to the linear
equation $0 = 1$, the system is inconsistent, and there are no solutions.

\end{solution}
     \end{exercise}

\begin{solution}


\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.7c}
$\left(\begin{array}{rrrr|r}
 1  & -6 & 0 & 0 & 1\\
 0  &  0 & 1 & 0 & 9 \\
 0  &  0 & 0 & 0 & 0
       \end{array}\right)$.
     \begin{selectAll}
       \choice[correct]{The $1^{st}$ column contains a pivot.}
       \choice{The $2^{nd}$ column contains a pivot.}
       \choice[correct]{The $3^{rd}$ column contains a pivot.}
       \choice{The $4^{th}$ column contains a pivot.}
       \choice{The $5^{th}$ column contains a pivot.}
     \end{selectAll}
     \begin{exercise}
       The solutions of the system are:
\[
\left(\begin{array}{r} x_1 \\ x_2 \\ x_3\\ x_4\end{array} \right)
= \left(\begin{array}{c} 1 + \answer{6}x_2 \\ x_2 \\ \answer{9} \\ x_4\end{array} \right)
\]

\begin{solution}
The $1^{st}$ and $3^{rd}$ columns of the matrix contain
pivots.  The solutions of the system are:

\[
\left(\begin{array}{r} x_1 \\ x_2 \\ x_3\\ x_4\end{array} \right)
= \left(\begin{array}{c} 1 + 6x_2 \\ x_2 \\ 9 \\ x_4\end{array} \right)
\]

\end{solution}
     \end{exercise}

\begin{solution}


\end{solution}
\end{exercise}


\begin{exercise} \label{c2.3.8}
(a) Consider the $2\times 2$ matrix
\begin{equation}  \label{e:2x2}
\mattwo{a}{b}{c}{1}
\end{equation}
where $a,b,c\in\R$ and $a\neq 0$.  Show that \eqref{e:2x2}
is row equivalent to the matrix
\[
\left(\begin{array}{cc} 1 & \frac{b}{a} \\ 0 & \frac{\answer[given]{a-bc}}{a}
\end{array} \right).
\]
(b)  Show that \eqref{e:2x2} is row equivalent to the
identity matrix if and only if $a\neq \answer[given]{bc}$.

\begin{hint}
Two matrices are row equivalent if there is a sequence of row operations
that leads from one to the other.  In this case: Divide the $1^{st}$ row of \eqref{e:2x2} by $a$.
\end{hint}
\begin{hint}
Then subtract $c$ times the $1^{st}$ row from the $2^{nd}$ row.
\end{hint}
\begin{hint}
The result is
\[
\left(\begin{array}{cc} 1 & \frac{b}{a} \\ 0 & \frac{a - bc}{a}\end{array}
\right).
\]
\end{hint}
\begin{hint}
For (b), if $a \neq bc$, then we can row reduce \eqref{e:2x2} to $I_2$.  First
perform the two row operations in part (a).
\end{hint}
\begin{hint}
  Then multiply the $2^{nd}$ row by $\frac{a}{a - bc}$.
\end{hint}
\begin{hint}
  Finally subtract $\frac{b}{a}$ times the second row from the first row.
\end{hint}
\begin{hint}
This will give us the identity matrix $I_2$.
\end{hint}
\begin{hint}
  If $a = bc$, then $\frac{a - bc}{a} = 0$, and the matrix is
\[
\left(\begin{array}{cc} 1 & \frac{b}{a} \\ 0 & 0\end{array} \right).
\]
which has only one pivot and cannot be row reduced to $I_2$.
\end{hint}

\begin{solution}

(a)
Two matrices are row equivalent if there is a sequence of row operations
that leads from one to the other.  In this case:

1. Divide the $1^{st}$ row of \eqref{e:2x2} by $a$.

2. Subtract $c$ times the $1^{st}$ row from the $2^{nd}$ row.

The result is
\[
\left(\begin{array}{cc} 1 & \frac{b}{a} \\ 0 & \frac{a - bc}{a}\end{array}
\right).
\]

(b) If $a \neq bc$, then we can row reduce \eqref{e:2x2} to $I_2$ by
performing the two row operations in part (a), followed by:

3. Multiply the $2^{nd}$ row by $\frac{a}{a - bc}$.

4. Subtract $\frac{b}{a}$ times the second row from the first row.

This will give us the identity matrix $I_2$.

\para If $a = bc$, then $\frac{a - bc}{a} = 0$, and the matrix is
\[
\left(\begin{array}{cc} 1 & \frac{b}{a} \\ 0 & 0\end{array} \right).
\]
which has only one pivot and cannot be row reduced to $I_2$.

\end{solution}
\end{exercise}

\begin{exercise} \label{c2.3.9}
Use row reduction and back substitution to solve the following
system of two equations in three unknowns:
\[
\begin{array}{rcrcrcrc}
 x_1 & - & x_2 & + & x_3 & = & 1 \\
2x_1 & + & x_2 & - & x_3 & = & -1
\end{array}
\]
\begin{hint}
   Row reduce the augmented matrix of the system:
\[
\left(\begin{array}{rrr|r} 1 & -1 & 1 & 1 \\ 2 & 1 & -1 & -1\end{array}\right)
\longrightarrow
\left(\begin{array}{rrr|r} 1 & 0 & 0 & 0 \\ 0 & 1 & -1 & -1\end{array}\right).
\]
\end{hint}
\begin{prompt}
  The solution to this system is
\[
\left(\begin{array}{c} x_1 \\ x_2 \\ x_3\end{array}\right) =
\left(\begin{array}{c} \answer{0} \\ x_3 - \answer{1} \\ x_3\end{array}\right),
\]
where $x_3$ is any real number.
\end{prompt}
Is $(1,2,2)$ a solution to this system?
\begin{multipleChoice}
  \choice{Yes, $(1,2,2)$ is a solution.}
  \choice[correct]{No, $(1,2,2)$ is not a solution.}
\end{multipleChoice}
If not, is there a solution for which $x_3=2$?
\begin{multipleChoice}
  \choice[correct]{Yes, there is a solution for which $x_3 = 2$.}
  \choice{No, there is no solution for which $x_3 = 2$.}  
\end{multipleChoice}
\begin{hint}
  There is a solution for which $x_3 = 2$, namely $(0,1,2)$.
\end{hint}

\begin{solution}

\ans The solution to this system is
\[
\left(\begin{array}{c} x_1 \\ x_2 \\ x_3\end{array}\right) =
\left(\begin{array}{c} 0 \\ x_3 - 1 \\ x_3\end{array}\right),
\]
where $x_3$ is any real number.

\soln Row reduce the augmented matrix of the system:
\[
\left(\begin{array}{rrr|r} 1 & -1 & 1 & 1 \\ 2 & 1 & -1 & -1\end{array}\right)
\longrightarrow
\left(\begin{array}{rrr|r} 1 & 0 & 0 & 0 \\ 0 & 1 & -1 & -1\end{array}\right).
\]
Although $(1,2,2)$ is not a solution to this system, there is a solution
for which $x_3 = 2$, namely $(0,1,2)$.

\end{solution}
\end{exercise}

\noindent In Exercises~\ref{c2.3.10a} -- \ref{c2.3.10b} determine the
augmented matrix and all solutions for each system of linear equations
\begin{exercise} \label{c2.3.10a}
$\begin{array}{rcl}
x-y+z & = & 1 \\
4x+y+z & = & 5 \\
   2x+3y-z & = & 2 \end{array}$.
 %BADBAD

\begin{solution}
The augmented matrix for this system is
\[
\left(\begin{array}{rrr|r} 1 & -1 & 1 & 1 \\ 4 & 1 & 1 & 5 \\
2 & 3 & -1 & 2\end{array}\right)
\]
We can row reduce this matrix to reduced echelon form, obtaining
\[
\left(\begin{array}{rrr|r} 1 & 0 & \frac{2}{5} & \frac{6}{5} \\
0 & 1 & -\frac{3}{5} & \frac{1}{5} \\ 0 & 0 & 0 & 1\end{array}\right)
\]
The last row of the reduced system implies that $0 = 1$, so the system
is inconsistent and has no solutions.

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.10b}
$\begin{array}{rcl}
2x-y+z+w & = & 1\\
   x+2y-z+w & = & 7 \end{array}$.
  %BADBAD

\begin{solution}
The augmented matrix for this system is
\[
\left(\begin{array}{rrrr|r} 2 & -1 & 1 & 1 & 1 \\ 1 & 2 & -1 & 1 & 7
\end{array}\right)
\]
which can be row reduced to
\[
\left(\begin{array}{rrrr|r} 1 & 0 & \frac{1}{5} & \frac{3}{5} &
\frac{9}{5} \\ 0 & 1 & -\frac{3}{5} & \frac{1}{5} & \frac{13}{5}
\end{array}\right).
\]
The solution set is therefore
\[
\left(\begin{array}{c} x_1 \\ x_2 \\ x_3 \\ x_4\end{array}\right) =
\left(\begin{array}{c} \frac{9}{5} - \frac{1}{5}x_3 - \frac{3}{5}x_4
\\ \frac{13}{5} + \frac{3}{5}x_3 - \frac{1}{5}x_4 \\ x_3 \\ x_4
\end{array}\right).
\]

\end{solution}
\end{exercise}

\noindent In Exercises~\ref{c2.3.11a} -- \ref{c2.3.11d} consider the
augmented matrices representing systems of linear equations, and decide
\begin{itemize}
\item[(a)] if there are zero, one or infinitely many solutions, and
\item[(b)] if solutions are not unique, how many variables can be
assigned arbitrary values.
\end{itemize}
\begin{exercise} \label{c2.3.11a}
$\left(\begin{array}{ccc|c} 1 & 0 & 0 &3 \\0 & 2 & 1 & 1\\ 0 & 0 & 0 & 0
       \end{array}\right)$.
     \begin{multipleChoice}
       \choice{There are no solutions.}
       \choice{There is exactly one solution.}
       \choice[correct]{There are infinitely many solutions.}
     \end{multipleChoice}
     \begin{hint}
       The row-reduced form of the matrix is:
\[
\left(\begin{array}{rrr|r} 1 & 0 & 0 & 3 \\ 0 & 1 &
\frac{1}{2} & \frac{1}{2} \\ 0 & 0 & 0 & 0\end{array}\right).
\]
     \end{hint}
     \begin{exercise}
       There is/are $\answer{1}$ variable(s) that may be assigned arbitrary values.
     \end{exercise}
     
\begin{solution}

\ans (a) The system has infinitely many solutions.

(b) One variable can be assigned arbitrary values.

\soln The row-reduced form of the matrix is:

\[
\left(\begin{array}{rrr|r} 1 & 0 & 0 & 3 \\ 0 & 1 &
\frac{1}{2} & \frac{1}{2} \\ 0 & 0 & 0 & 0\end{array}\right).
\]

\end{solution}
\end{exercise}

\begin{exercise} \label{c2.3.11b}
$\left(\begin{array}{cccc|c} 1 & 2 & 0 & 0 & 3\\ 0 & 1 & 1 & 0 & 1\\
         0 & 0 & 0 & 0 & 2 \end{array}\right)$.
     \begin{multipleChoice}
       \choice[correct]{There are no solutions.}
       \choice{There is exactly one solution.}
       \choice{There are infinitely many solutions.}
     \end{multipleChoice}
     \begin{hint}
       The row-reduced form of the matrix is:
\[
\left(\begin{array}{rrrr|r} 1 & 2 & 0 & 0 & 3 \\
0 & 1 & 1 & 0 & 1 \\ 0 & 0 & 0 & 0 & 1\end{array}\right).
\]
     \end{hint}

\begin{solution}

\ans The system has no solutions.

\soln The row-reduced form of the matrix is:
\[
\left(\begin{array}{rrrr|r} 1 & 2 & 0 & 0 & 3 \\
0 & 1 & 1 & 0 & 1 \\ 0 & 0 & 0 & 0 & 1\end{array}\right).
\]

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.11c}
$\left(\begin{array}{ccc|c}  1 & 0 & 2 & 1\\ 0 & 5 & 0 & 2 \\ 0 & 0 & 4 & 3
       \end{array}\right)$.
     \begin{multipleChoice}
       \choice{There are no solutions.}
       \choice[correct]{There is exactly one solution.}
       \choice{There are infinitely many solutions.}
     \end{multipleChoice}
     \begin{hint}
       The row-reduced form of the matrix is:
\[
\left(\begin{array}{rrr|r} 1 & 0 & 2 & 1 \\ 0 & 1 & 0 & \frac{2}{5}
\\ 0 & 0 & 1 & \frac{3}{4}\end{array}\right).
\]
     \end{hint}

\begin{solution}

\ans The system has a unique solution.

\soln The row-reduced form of the matrix is:
\[
\left(\begin{array}{rrr|r} 1 & 0 & 2 & 1 \\ 0 & 1 & 0 & \frac{2}{5}
\\ 0 & 0 & 1 & \frac{3}{4}\end{array}\right).
\]

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.11d}
$\left(\begin{array}{cccc|c} 1 & 0 & 2 & 0 & 3 \\ 2 & 3 & 6 & 1 & 16\\
         0 & 3 & 2 & 1 & 10 \\ 0 & 0 & 0 & 0 & 0  \end{array}\right)$.
     \begin{hint}
       The row-reduced form of the matrix is:
\[
\left(\begin{array}{rrrr|r} 1 & 0 & 2 & 0 & 3 \\ 0 & 1 & \frac{2}{3}
& \frac{1}{3} & \frac{10}{3} \\ 0 & 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 & 0
\end{array}\right).
\]
     \end{hint}
          \begin{multipleChoice}
       \choice{There are no solutions.}
       \choice{There is exactly one solution.}
       \choice[correct]{There are infinitely many solutions.}
     \end{multipleChoice}
     \begin{exercise}
       There is/are $\answer{2}$ variable(s) that may be assigned arbitrary values.

\begin{solution}

\ans (a) The system has infinitely many solutions.

(b) Two variables can be assigned arbitrary values.

\soln The row-reduced form of the matrix is:
\[
\left(\begin{array}{rrrr|r} 1 & 0 & 2 & 0 & 3 \\ 0 & 1 & \frac{2}{3}
& \frac{1}{3} & \frac{10}{3} \\ 0 & 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 & 0
\end{array}\right).
\]


\end{solution}
     \end{exercise}

\begin{solution}


\end{solution}
\end{exercise}

\noindent A system of $m$ equations in $n$ unknowns is linear if it has
the form \eqref{general}; any other system of equations is called 
{\em nonlinear\/}. \index{nonlinear}  In Exercises~\ref{c2.3.12a} -- 
\ref{c2.3.12e} decide whether each of the given systems of equations 
is linear or nonlinear.
\begin{exercise} \label{c2.3.12a}
\[
\begin{array}{rcl}
3x_1 - 2x_2 +14x_3-7x_4 & = & 35\\
2x_1 + 5x_2 - 3x_3 + 12x_4 & = & -1
\end{array}
\]
\begin{multipleChoice}
  \choice[correct]{The system is linear.}
  \choice{The system is not linear.}  
\end{multipleChoice}

\begin{solution}
\ans The system is linear.

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.12b}
\[
\begin{array}{rcl}
3x_1 + \pi x_2  & = & 0\\
2x_1 - e x_2    & = & 1
\end{array}
\]
\begin{multipleChoice}
  \choice[correct]{The system is linear.}
  \choice{The system is not linear.}  
\end{multipleChoice}

\begin{solution}
\ans The system is linear.

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.12c}
\[
\begin{array}{rcl}
3x_1x_2 -x_2  & = & 10\\
2x_1 -   x_2^2    & = & -5
\end{array}
\]
\begin{multipleChoice}
  \choice{The system is linear.}
  \choice[correct]{The system is not linear.}  
\end{multipleChoice}
\begin{hint}
  The term $3x_1x_2$ contains two variables, and the term $x_2^2$
contains a variable squared.  Thus, these terms are not linear.
\end{hint}

\begin{solution}
\ans The system is not linear.

\soln The term $3x_1x_2$ contains two variables, and the term $x_2^2$
contains a variable squared.  Thus, these terms are not linear.

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.12d}
\[
\begin{array}{rcl}
3x_1  -x_2  & = & \cos(12)\\
2x_1 -   x_2    & = & -5
\end{array}
\]
\begin{multipleChoice}
  \choice[correct]{The system is linear.}
  \choice{The system is not linear.}  
\end{multipleChoice}
\begin{hint}
  You may be distracted by $\cos(12)$ but note that $\cos(12)$ is constant.
\end{hint}

\begin{solution}
\ans The system is linear.

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.12e}
\[
\begin{array}{rcl}
3x_1  - \sin(x_2)  & = & 12\\
2x_1 -   x_3    & = & -5
\end{array}
\]
\begin{multipleChoice}
  \choice{The system is linear.}
  \choice[correct]{The system is not linear.}  
\end{multipleChoice}
\begin{hint}
  The term $\sin(x_2)$ is not linear.
\end{hint}

\begin{solution}
\ans The system is not linear.

\soln The term $\sin(x_2)$ is not linear.


\end{solution}
\end{exercise}

\CEXER



\noindent In Exercises~\ref{c2.3.1a} -- \ref{c2.3.1c} use elementary row
operations and \Matlab to put each of the given matrices into row echelon
form.  Suppose that the matrix is the augmented matrix for a system of
linear equations.  Is the system consistent or inconsistent?

\begin{exercise} \label{c2.3.1a}
\[
\left(\begin{array}{rrr}
 2 &  1  &  1   \\
 4 &  2  &  3
\end{array}\right).
\]

\begin{solution}
\ans The row echelon form matrix is
\begin{verbatim}
A = 
    1.0000    0.5000    0.5000
         0         0    1.0000
\end{verbatim}
\soln Enter the augmented matrix into \Matlab as $A$, then reduce to
row echelon form by typing:
\begin{verbatim}
A(1,:) = A(1,:)/2
A(2,:) = A(2,:) - 4*A(1,:)
\end{verbatim}
The linear system that this matrix represents is inconsistent, since
the $2^{nd}$ row of the reduced matrix represents the equation $0 = 1$.

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.1b}
\[
\left(\begin{array}{rrrr}
 3  & -4 & 0 & 2\\
 0  &  2 & 3 & 1\\
 3  &  1 & 4 & 5
\end{array}\right).
\]

\begin{solution}
The row-reduced matrix is:
\begin{verbatim}
A =
    1.0000   -1.3333         0    0.6667
         0    1.0000    1.5000    0.5000
         0         0    1.0000   -0.1429
\end{verbatim}
This matrix represents a consistent linear system.

\end{solution}
\end{exercise}
\begin{exercise} \label{c2.3.1c}
\[
\left(\begin{array}{rrrr}
 -2 & 1 &  9 & 1\\
  3 & 3 & -4 & 2\\
  1 & 4 &  5 & 5
\end{array}\right).
\]

\begin{solution}
The row-reduced matrix is:
\begin{verbatim}
A =
    1.0000   -0.5000   -4.5000   -0.5000
         0    1.0000    2.1111    0.7778
         0         0         0    2.0000
\end{verbatim}
This matrix represents an inconsistent linear system.


\end{solution}
\end{exercise}

\noindent {\bf Observation:} {\rm In standard format \Matlab displays
all nonzero real numbers with four decimal places while it displays
zero as $0$.  An unfortunate consequence of this display is that
when a matrix has both zero and noninteger entries, the columns
will not align --- which is a nuisance.  You can work with
rational numbers rather than decimal numbers by typing {\tt
format rational}.  Then the columns will align.}

\begin{exercise} \label{c2.3.2}
Load the following $6\times 8$ matrix $A$ into \Matlab by typing
{\tt e2\_3\_16}.
\begin{matlabEquation}\label{MATLAB:13}
A=\left(\begin{array}{rrrrrrrr}
0 & 0 &  0 &   1 &   3 &   5 & 0 &  9 \\
0 & 3 &  6 &  -6 &  -6 & -12 & 0 &  1 \\
0 & 2 &  4 &  -5 &  -7 &  14 & 0 &  1 \\
0 & 1 &  2 &   1 &  14 &  21 & 0 & -1 \\
0 & 0 &  0 &   2 &   4 &   9 & 0 &  7 \\
0 & 5 & 10 & -11 & -13 &   2 & 0 &  2
\end{array}\right)
\end{matlabEquation}
Use \Matlab to transform this matrix to row echelon form.

\begin{solution}
The row echelon form is:
\begin{verbatim}
A = 
       0   1.0000   2.0000   1.0000  14.0000  21.0000        0  -1.0000
       0        0        0   1.0000   3.0000   5.0000        0   9.0000
       0        0        0        0   1.0000  -0.5000        0  -4.7143
       0        0        0        0        0   1.0000        0   0.3457
       0        0        0        0        0        0        0   1.0000
       0        0        0        0        0        0        0        0
\end{verbatim}


\end{solution}
\end{exercise}

\begin{exercise} \label{c2.3.3}
Use row reduction and back substitution to solve the following
system of linear equations:
\[
\begin{array}{rcrcrcrcr}
2x_1 & + &  3x_2 &  - &  4x_3 & + &  x_4 &  = & 2 \\
3x_1 & - &   x_2 &  - &   x_3 & + & 2x_4 &  = & 4 \\
 x_1 & - &  7x_2 &  + &  5x_3 & - &  x_4 &  = & 6
\end{array}
\]

\begin{solution}

\ans The system's solution is
\[
\left(\begin{array}{c} x_1 \\ x_2 \\ x_3 \\ x_4\end{array}\right) =
\left(\begin{array}{c} -\frac{7}{4} - \frac{21}{8}x_4 \\ -\frac{9}{2} -
\frac{11}{4}x_4 \\ -\frac{19}{4} - \frac{25}{8}x_4 \\ x_4\end{array}\right)
\]
where $x_4$ is any real number.

\soln Note that the system of linear equations is related to the
augmented matrix
\begin{verbatim}
A = 
    2        3       -4        1        2
    3       -1       -1        2        4
    1       -7        5       -1        6
\end{verbatim}
which row reduces to
\begin{verbatim}
A =
    1       3/2      -2       1/2       1
    0        1     -10/11    -1/11    -2/11
    0        0        1      25/8    -19/4
\end{verbatim}
The system can then be solved by back substitution.

\end{solution}
\end{exercise}


\begin{exercise} \label{c2.3.4}
{\bf Comment:} {\rm To understand the point of this exercise you
must begin by typing the \Matlab command {\tt format short e}.
This command will set a format in which you can see the
difficulties that sometimes arise in numerical computations.}

Consider the following two $3\times 3$-matrices:
\begin{matlabEquation}\label{MATLAB:14}
A=\left( \begin{array}{rrr}
     1  &  3  &  4\\
     2  &  1  &  1\\
    -4  &  3  &  5
\end{array}\right) \AND
B=\left( \begin{array}{rrr}
     3  &  1  &  4\\
     1  &  2  &  1\\
     3  & -4  &  5
\end{array}\right).
\end{matlabEquation}
Note that matrix $B$ is obtained from matrix $A$ by interchanging the
first two columns.
\begin{itemize}
\item[(a)] Use \Matlab to put $A$ into row echelon form using the
transformations
\begin{enumerate}
\item Subtract $2$ times the $1^{st}$ row from the $2^{nd}$.
\item Add $4$ times the $1^{st}$ row to the $3^{rd}$.
\item Divide the $2^{nd}$ row by $-5$.
\item Subtract $15$ times the $2^{nd}$ row from the $3^{rd}$.
\end{enumerate}
\item[(b)] Put $B$ by hand into row echelon form using the
transformations
\begin{enumerate}
\item Divide the $1^{st}$ row by $3$.
\item Subtract the $1^{st}$ row from the $2^{nd}$.
\item Subtract $3$ times the $1^{st}$ row from the $3^{rd}$.
\item Multiply the $2^{nd}$ row by $3/5$.
\item Add $5$ times the $2^{nd}$ row to the $3^{rd}$.
\end{enumerate}
\item[(c)] Use \Matlab to put $B$ into row echelon form using the
same transformations as in part (b).
\item[(d)] Discuss the outcome of the three transformations.  Is
there a difference in the results?  Would you expect to see a
difference?  Could the difference be crucial when solving a system
of linear equations?
\end{itemize}

\begin{solution}

(a)
\begin{verbatim}
A =
   1.0000e+00   3.0000e+00   4.0000e+00
            0   1.0000e+00   1.4000e+00
            0            0            0
\end{verbatim}

(b)
$B = \left(\begin{array}{rrr}
1 & \frac{1}{3} & \frac{4}{3} \\
0 & 1 & -\frac{1}{5} \\
0 & 0 & 0\end{array} \right)$

(c)
\begin{verbatim}
B =
   1.0000e+00   3.3333e-01   1.3333e+00
            0   1.0000e+00  -2.0000e-01
            0            0   2.2204e-16
\end{verbatim}


(d) Note that switching the first two columns of matrix {\tt A} produces
matrix {\tt B}.  Suppose that {\tt A} and {\tt B} represent the left-hand
sides of linear systems with the same vector representing the right-hand
sides.  If the solution of system {\tt A} is $(x,y,z) = (a,b,c)$, then the
solution of system {\tt B} should be $(x,y,z) = (b,a,c)$.  However,
according to the row reduced matrices produced by \Matlab, the system
corresponding to {\tt B} has a unique solution, and the system
corresponding to {\tt A} does not.  Row reducing {\tt B} by hand shows
that there is not a unique solution.  \Matlab calculations provide one because of 
roundoff error.  The division by 3 in the first step of the row reduction
for {\tt B} causes a rounding inaccuracy.  Because of this, \Matlab
eventually computes a very small nonzero value for {\tt B}$(3,3)$
rather than the correct answer of 0.

\end{solution}
\end{exercise}

\begin{exercise} \label{c2.3.5}
Find a cubic polynomial
\[
p(x) = ax^3 + bx^2 + cx + d
\]
so that $p(1)=2$, $p(2)=3$, $p'(-1)=-1$, and $p'(3)=1$.

\begin{solution}

\ans The polynomial $p(x) = -\frac{3}{44}x^3 + \frac{5}{11}x^2 +
\frac{5}{44}x + \frac{3}{2}$ satisfies the given conditions.

\soln Let $p(x) = ax^3 + bx^2 + cx + d$ be the general cubic polynomial. 
Then $p'(x) = 3ax^2 + 2bx + c$.  We can write the conditions on $p$ as a
system of linear equations:
\[
\begin{array}{rrrrrrrrr}
a & + & b & + & c & + & d & = & 2 \\
8a & + & 4b & + & 2c & + & d & = & 3 \\
3a & - & 2b & + & c & & & = & -1 \\
27a & + & 6b & + & c & & & = & 1\end{array}
\]
Using \Matlab, we solve for $a$, $b$, $c$, and $d$, obtaining the
vector of coefficients:
\begin{verbatim}
ans =
    -3/44
     5/11
     5/44
     3/2
\end{verbatim}






\end{solution}
\end{exercise}



\end{document}
