\documentclass{ximera}

\input{../preamble.tex}

\title{Solving Linear Equations}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

\normalsize

The primary motivation for the study of vectors and matrices is based
on the study of solving systems of linear equations.  The algorithms
that enable us to find solutions are themselves based on certain kinds
of matrix manipulations.  In these algorithms, matrices serve as a
shorthand for calculation, rather than as a basis for a theory.  We will
see later that these matrix manipulations do lead to a rich theory of how
to solve systems of linear equations.  But our first step is just to see
how these equations are actually solved.

We begin with a discussion in Section~\ref{S:2.1} of how to write systems
of linear equations in terms of matrices.  We also show by example how
complicated writing down the answer to such systems can be.  In
Section~\ref{S:2.2}, we recall that solution sets to systems of linear
equations in two and three variables are lines and planes.

The best known and probably the most efficient method for solving
systems of linear equations (especially with a moderate to large number
of unknowns) is Gaussian elimination.  The idea behind this method,
which is introduced in Section~\ref{S:Gauss}, is to manipulate matrices
by elementary row operations to reduced echelon form.  It is then possible
just to look at the reduced echelon form matrix and to read off the
solutions to the linear system, if any.  The process of reading off the
solutions is formalized in Section~\ref{S:2.4}; see Theorem~\ref{number}.
Our discussion of solving linear equations is presented with equations
whose coefficients are real numbers --- though most of our examples have
just integer coefficients.  The methods work just as well with complex
numbers, and this generalization is discussed in Section~\ref{S:specialcoeff}.

Throughout this chapter, we alternately discuss the theory and show how
calculations that are tedious when done by hand can easily be performed
by computer using \Matlabp.  The chapter ends with a proof of the
uniqueness of row echelon form (a topic of theoretical importance) in
Section~\ref{S:uniquerowechelon}.  This section is included mainly for
completeness and need not be covered on a first reading.


\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../linearAlgebra"
%%% End:
