\documentclass{ximera}

\input{../preamble.tex}

\title{Uniqueness of Reduced Echelon Form}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{S:uniquerowechelon}


In this section we prove Theorem~\ref{uniquerowechelon}, which
states that every matrix is row equivalent to precisely one
reduced echelon form matrix.  

\noindent {\bf Proof of Theorem~\ref{uniquerowechelon}:}
Suppose that $E$ and $F$ are two $m\times n$ reduced echelon
matrices that are row equivalent to $A$.  Since elementary row
operations are invertible, the two matrices $E$ and $F$ are
row equivalent.  Thus, the systems of linear equations associated to 
the $m\times (n+1)$ matrices $(E|0)$ and $(F|0)$ must have exactly the
same set of solutions.  It is the fact that the solution sets
of the linear equations associated to $(E|0)$ and $(F|0)$ are
identical that allows us to prove that $E=F$.

Begin by renumbering the variables $x_1,\ldots,x_n$ so that the equations 
associated to $(E|0)$ have the form: 
\begin{equation}  \label{e1-ell2a}
\begin{array}{rcl}
  x_1  & = &  - a_{1,\ell+1}x_{\ell+1} - \cdots - a_{1,n}x_n\\
  x_2  & = &  - a_{2,\ell+1}x_{\ell+1} - \cdots - a_{2,n}x_n\\
\vdots &   &    \vdots \\
x_\ell & = &  - a_{\ell,\ell+1}x_{\ell+1} - \cdots - a_{\ell,n}x_n.
\end{array}
\end{equation}
In this form, pivots of $E$ occur in the columns $1,\ldots,\ell$.  We begin 
by showing that the matrix $F$ also has pivots in columns $1,\ldots,\ell$. 
Moreover, there is a unique solution to these equations for {\em every\/}
choice of numbers $x_{\ell+1},\ldots,x_n$.  

Suppose that the pivots of $F$ do not occur in columns $1,\ldots,\ell$.  Then
there is a row in $F$ whose first nonzero entry occurs in a column $k>\ell$. 
This row corresponds to an equation
\[
x_k = c_{k+1}x_{k+1} + \cdots + c_nx_n.
\]
Now, consider solutions that satisfy
\[
x_{\ell+1} = \cdots = x_{k-1} = 0 \AND x_{k+1} = \cdots = x_n = 0.
\]
In the equations associated to the matrix $(E|0)$, there is a unique solution
associated with every number $x_k$; while in the equations associated to the 
matrix $(F|0)$, $x_k$ must be zero to be a solution.  This argument
contradicts the fact that the $(E|0)$ equations and the $(F|0)$
equations have the same solutions.  So the pivots of $F$ must also occur in 
columns $1,\ldots,\ell$, and the equations associated to $F$ must have the 
form:
\begin{equation} \label{e1-ell2b}
\begin{array}{rcl}
x_1  & = & - \hat{a}_{1,\ell+1}x_{\ell+1} - \cdots - \hat{a}_{1,n}x_n \\
x_2  & = & - \hat{a}_{2,\ell+1}x_{\ell+1} - \cdots - \hat{a}_{2,n}x_n \\
\vdots &   &    \vdots  \\
x_\ell & = & - \hat{a}_{\ell,\ell+1}x_{\ell+1} - \cdots - \hat{a}_{\ell,n}x_n
\end{array}
\end{equation}
where $\hat{a}_{i,j}$ are scalars.

To complete this proof, we show that $a_{i,j}=\hat{a}_{i,j}$.  These 
equalities are verified as follows.  There is just one solution to each 
system \Ref{e1-ell2a} and \Ref{e1-ell2b} of the form
\[
x_{\ell+1}=1,\; x_{\ell+2}=\cdots=x_n=0.
\]
These solutions are
\[
(-a_{1,\ell+1}, \ldots, -a_{\ell,\ell+1},1,0,\cdots,0)
\]
for \Ref{e1-ell2a} and
\[
(-\hat{a}_{1,\ell+1},\ldots,-\hat{a}_{\ell,\ell+1},1,0\cdots,0)
\]
for \Ref{e1-ell2b}. It follows that $a_{j,\ell+1}=\hat{a}_{j,\ell+1}$
for $j=1,\ldots,\ell$.  Complete this proof by repeating this argument. 
Just inspect solutions of the form
\[
x_{\ell+1}=0,\; x_{\ell+2}=1,\; x_{\ell+3}=\cdots= x_n=0
\]
through
\[
x_{\ell+1}=\cdots = x_{n-1}=0,\; x_n=1.
\]



\end{document}
