\documentclass{ximera}

\input{../preamble.tex}

\title{Linear Differential Equations}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle


\label{C:LDE}

\normalsize

Primarily our discussions of differential equations have focused on 
two issues: generalizations of solutions of the differential equation 
$\dot{x}=\lambda x$ to systems of equations and the qualitative 
interpretation of numerically obtained phase portraits for autonomous
nonlinear systems.   Except for finding closed form solutions of 
systems of equations $\dot{X}=AX$ (in the plane, Chapter~\ref{Chap:Planar}, 
or in Jordan normal form, Section~\ref{sec:LinHomSys}), we have solved in 
closed form virtually no other differential equation.  In this chapter and 
the next two we focus on finding closed form solutions to a variety of 
differential equations.  This chapter and the next discuss constant 
coefficient linear equations, both homogeneous and inhomogeneous, 
while Chapter~\ref{chap:SingleOdes} discusses nonconstant coefficient
and nonlinear equations.  

The system of differential equations $\dot{X}=AX$ is a constant coefficient, 
first order, homogeneous, linear system of ordinary differential equations. 
We begin this chapter (Section~\ref{S:SEOC}) by discussing how to solve 
the systems $\dot{X}=AX$ in the given coordinates (rather than by first 
transforming $A$ to Jordan normal form, as we did in 
Section~\ref{sec:LinHomSys}).  For example, we present a formula for 
computing $e^{tA}$ for any matrix $A$.  Later we solve some linear 
equations that are neither homogeneous nor first order.   

In Section~\ref{sec:HighOrder} we discuss how to solve higher order 
linear differential equations by reducing them to first order systems.  
This section generalizes the reduction of second order equations to planar
systems described in Section~\ref{S:SOE}.  We will see that generally it is 
easier to solve higher order equations in closed form than to solve first 
order systems.  Section~\ref{sec:2norderinhom} discusses the solution of an 
inhomogeneous higher order equation by undetermined coefficients.  The method 
of undetermined coefficients is most easily understood in the language of 
linear differential operators, and this language is introduced in 
Section~\ref{S:LDO}.

The chapter ends with a discussion of resonance in Section~\ref{S:resonance}. 
Here we use explicit closed form solutions found by the method of 
undetermined coefficients to understand a physically motivated phenomenon 
that occurs in solutions to forced second order equations.  


\end{document}
