\documentclass{ximera}
\input{../../preamble.tex}
\begin{document}

\noindent In Exercises~\ref{c6.2.5A} -- \ref{c6.2.5C} we use
Theorem~\ref{T:linODEsoln}, the uniqueness of solutions to initial value
problems, in perhaps a surprising way.
\begin{exercise}  \label{c6.2.5A}
Prove that
\[
e^{t+s} = e^te^s
\]
for all real numbers $s$ and $t$.  {\bf Hint:}
\begin{itemize}
\item[(a)]  Fix $s$ and verify that $y(t) = e^{t+s}$ is a solution to the
initial value problem
\begin{equation}  \label{E:init1}
\begin{array}{rcl}
\frac{dx}{dt} & = & x \\
x(0) & = & e^s
\end{array}
\end{equation}
\item[(b)] Fix $s$ and verify that $z(t) = e^te^s$ is also a solution to
\eqref{E:init1}.
\item[(c)]  Use Theorem~\ref{T:linODEsoln} to conclude that $y(t)=z(t)$ for
every $s$.
\end{itemize}

\begin{solution}

(a) To verify that $y(t) = e^{t + s}$ is a solution to the initial value
problem, first substitute $y(t)$ into the left hand side of the equation.
Using the chain rule, obtain
\[
\frac{dy}{dt}(t) = \frac{d}{dt}(e^{t + s}) = \frac{d}{dt}(t + s)e^{t + s}
= e^{t + s}.
\]
Then substitute $y(t)$ into the right hand side of the equation, obtaining
\[
y(t) = e^{t + s}.
\]
Thus, the left hand and right hand sides are equal, so $y(t)$ is a
solution to the differential equation.  Finally, check to see that
$y(t)$ satisfies the initial value:
\[
y(0) = e^{0 + s} = e^s
\]
as desired.

(b) Similarly, verify that $z(t) = e^te^s$ is a solution to the initial
value problem by substituting $z(t)$ into each side of the differential
equation:
\[
\frac{dz}{dt}(t) = \frac{d}{dt}(e^te^s) = e^te^s \AND
z(t) = e^te^s.
\]
Note that the results are equal, so that $z(t)$ is a solution to the
differential equation.  Since
\[
z(0) = e^0e^s = e^s,
\]
it follows that $z(t)$ is also a solution to the initial value problem.

(c) By Theorem~\ref{exist&unique}, if
$f(x)$ is differentiable near $x_0$, and if $\frac{df}{dx}$ is continuous,
then there is a unique solution to the differential equation $\dot{x} = f(x)$
with initial condition $x(0) = x_0$.  Let $f(x) = x$ and let $x(0) = e^s$.
Then, as shown in (a) and (b) of this problem, $e^{t + s}$ and $e^te^s$ are
both solutions.  However, by Theorem~\ref{exist&unique}, there is only one
solution, so $e^{t + s} = e^te^s$.

\end{solution}
\end{computerExercise}
\begin{computerExercise}  \label{c6.2.5B}
Let $A$ be an $n\times n$ matrix.  Prove that
\[
e^{(t+s)A} = e^{tA}e^{sA}
\]
for all real numbers $s$ and $t$.  {\bf Hint:}
\begin{itemize}
\item[(a)]  Fix $s\in\R$ and $X_0\in\R^n$ and verify that
$Y(t) = e^{(t+s)A}X_0$ is a solution to the initial value problem
\begin{equation}  \label{E:init2}
\begin{array}{rcl}
\frac{dX}{dt} & = & AX \\
X(0) & = & e^{sA}X_0
\end{array}
\end{equation}
\item[(b)] Fix $s$ and verify that $Z(t) = e^{tA}\left(e^{sA}X_0\right)$ is
also a solution to \eqref{E:init2}.
\item[(c)]  Use the $n$ dimensional version of Theorem~\ref{T:linODEsoln} to
conclude that $Y(t)=Z(t)$ for every $s$ and every $X_0$.
\end{itemize}
{\bf Remark:}  Compare the result in this exercise with the calculation in
Exercise~\ref{c6.2.5}.

\begin{solution}

(a) To verify that $Y(t)$ is a solution to the initial value problem
\eqref{E:init2}, first substitute $Y(t)$ into the left hand side of the
equation.  Using the chain rule, obtain
\[
\frac{dY}{dt}(t) = \frac{d}{dt}(e^{(t + s)A}X_0) =
\frac{d}{dt}((t + s)A)e^{(t + s)A}X_0 = Ae^{(t + s)A}X_0.
\]
Then substitute $Y(t)$ into the right hand side of the equation, obtaining
\[
AY(t) = Ae^{(t + s)A}X_0.
\]
Thus, the left hand and right hand sides of the equation are equal, so
$Y(t)$ is a solution to the differential equation.  Finally, check to see
that $Y(t)$ satisfies the initial value:
\[
Y(0) = e^{(0 + s)A}X_0 = e^{sA}X_0,
\]
as desired.

(b) Similarly, verify that $Z(t) = e^{tA}(e^{sA}X_0)$ is a solution to the
initial value problem by substituting $Z(t)$ into each side of the
differential equation:
\[
\frac{dZ}{dt}(t) = \frac{d}{dt}(e^{tA}(e^{sA}X_0))
= Ae^{tA}(e^{sA}X_0) \AND
AZ(t) = Ae^{tA}(e^{sA}X_0).
\]
Since the results are equal, $Z(t)$ is a solution to the differential
equation.  Evaluating at $t = 0$, we find
\[
Z(0) = e^{0}(e^{sA}X_0) = e^{sA}X_0,
\]
from which it follows that $Z(t)$ is also a solution to the initial
value problem.

(c) Since $Y(t)$ and $Z(t)$ are both solutions to the initial value problem
\[
\begin{array}{rcl}
\frac{dX}{dt} & = & AX \\
X(0) & = & e^{sA}X_0,
\end{array}
\]
it follows from the uniqueness part of Theorem~\ref{exist&unique} that
$Y(t) = Z(t)$.  Thus, $e^{(t + s)A} = e^{tA}e^{sA}$, as desired.


\end{solution}
\end{computerExercise}
\begin{computerExercise}  \label{c6.2.5C}
Prove that
\begin{equation}  \label{E:0-110E}
\exp\left(t\mattwo{0}{-1}{1}{0}\right) =
\mattwo{\cos t}{-\sin t}{\sin t}{\cos t}.
\end{equation}
{\bf Hint:}
\begin{itemize}
\item[(a)] Verify that $X_1(t) = \vectwo{\cos t}{\sin t}$ and
$X_2(t) = \vectwo{-\sin t}{\cos t}$ are solutions to the initial value problems
\begin{equation}  \label{E:init3}
\begin{array}{rcl}
\dps\frac{dX}{dt} & = & \mattwo{0}{-1}{1}{0}X \\
X(0) & = & e_j
\end{array}
\end{equation}
for $j=1,2$.
\item[(b)] Since $X_j(0)=e_j$, use Theorem~\ref{T:linODEsoln} to verify that
\begin{equation}   \label{E:0-110}
X_j(t) = \exp\left(t\mattwo{0}{-1}{1}{0}\right)e_j.
\end{equation}
\item[(c)]  Show that \eqref{E:0-110} proves \eqref{E:0-110E}
\end{itemize}

\begin{solution}

(a) To verify that $X_1(t) = (\cos t,\sin t)^t$ is a solution to the
initial value problem \eqref{E:init3}, substitute $X_1(t)$ into the left
hand side of the differential equation, obtaining
\[
\frac{dX_1}{dt} = \frac{d}{dt}\vectwo{\cos t}{\sin t} =
\vectwo{-\sin t}{\cos t}.
\]
Then substitute $X_1(t)$ into the right hand side of the differential
equation, obtaining
\[
\mattwo{0}{-1}{1}{0}X_1 = \mattwo{0}{-1}{1}{0}\vectwo{\cos t}{\sin t}
= \vectwo{-\sin t}{\cos t}.
\]
Since the two sides are equal, $X_1(t)$ is a solution to
\eqref{E:init3}.  Further, since
\[
X_1(0) = \vectwo{\cos 0}{\sin 0} = \vectwo{1}{0} = e_1,
\]
$X_1(t)$ is a solution to the given initial value problem.

\para Similarly, to verify that $X_2(t) = (-\sin t,\cos t)^t$ is a
solution to the initial value problem, substitute $X_2(t)$ into the
left hand side of the differential equation, obtaining
\[
\frac{dX_2}{dt} = \frac{d}{dt}\vectwo{-\sin t}{\cos t} =
\vectwo{-\cos t}{-\sin t}.
\]
Then substitute $X_2(t)$ into the right hand side of the differential
equation, obtaining
\[
\mattwo{0}{-1}{1}{0}X_2 = \mattwo{0}{-1}{1}{0}\vectwo{-\sin t}{\cos t}
= \vectwo{-\cos t}{-\sin t}.
\]
Since the two sides of the differential equation are equal, and since
\[
X_2(0) = \vectwo{-\sin 0}{\cos 0} = \vectwo{0}{1} = e_2,
\]
$X_2(t)$ is a solution to the initial value problem \eqref{E:init3}.

(b) By Theorem~\ref{T:linODEsoln}, the unique
solution to \eqref{E:init3} with initial condition $X(0) = e_j$ is
\[
Y_j(t) \equiv \exp\left(t\mattwo{0}{-1}{1}{0}\right)e_j.
\]
We showed in part (a) that $X_j(t)$ satisfies this initial value
problem.  Thus, by the uniqueness part of Theorem~\ref{exist&unique},
$X_j(t) = Y_j(t)$, as desired.

(c) By part (b), $X_j(t)$ is equal to $Y_j(t)$, which is defined as
the $j^{th}$ column of the matrix
\[
\exp\left(t\mattwo{0}{-1}{1}{0}\right).
\]
By definition, $X_j(t)$ is equal to the $j^{th}$ column of the matrix
\[
\mattwo{\cos t}{-\sin t}{\sin t}{\cos t}.
\]
Thus,
\[
\exp\left(t\mattwo{0}{-1}{1}{0}\right) =
\mattwo{\cos t}{-\sin t}{\sin t}{\cos t}.
\]

\end{solution}
\end{exercise}
\end{document}
